---
title: "01a. Stepwise conditional analysis"
description: "(1) Are modules A & B both individually associated with ASD risk, and (2) Are modules A & B both independently associated with ASD risk? (i.e., does module B still show significant ASD enrichment, even after accounting for module A?)\nCombined across all studies (testing all modules simultaneously)."
date: 2026-08-04
format: html
---

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "02_module_identification")
```

Load data

```{r load_data}
# Gene network assignments
df_network_data <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_gene-networks-all.RDS"))

# Module eigengene data
df_module_kme <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_co-expression-kME.RDS"))

# ASD rare variant data (with Z-score transformation)
df_asd_rarevar_zscore <- readRDS(file.path(analysis_dir, "00_data_format", "00_01_ASD-rarevar-stats.RDS"))
```


# --- 1) Format data --- #

Prep data for model

## 1.1 Convert module membership to binary tibble

```{r format_data}
# Gene network data: Pivot wider & dummy code each column (within network)
# genes are rows, modules are columns
# value = 1 if gene is present in module
df_mods_binary <- df_network_data %>% 
    mutate(
        data = map(
            .x = data,
            .f = ~ .x %>% 
                mutate(present = 1) %>% 
                dplyr::select(module, ensembl_gene_id, present) %>% 
                distinct() %>% 
                filter(!is.na(ensembl_gene_id)) %>% # remove unmapped gene IDs (N=0; all gene names were mapped)
                pivot_wider(
                    id_cols = ensembl_gene_id, #c(ensembl_gene_id, gene_name),
                    names_from = module,
                    values_from = present,
                    values_fill = 0
                ) %>%
                mutate(across(-ensembl_gene_id, as.numeric))
                #mutate(across(-c(ensembl_gene_id, gene_name), as.numeric)) 
        )
    ) %>% 
    pull(data) %>% 
    purrr::reduce(., ~ inner_join(.x, .y, by = join_by(ensembl_gene_id))) %>% 
    dplyr::select(-none) # remove "none" column that was create in Wang data (for unmapped background genes)
```

## 1.2 Combine module membership with ASD rare variant stats

```{r combine_module_ASD_stats}
# Merge gene-level module membership data with gene-level ASD metrics
df_step_model <- left_join(df_asd_rarevar_zscore %>% 
                               dplyr::rename(gene_name_asd = gene_name),
                           df_mods_binary,
                           by = join_by(ensembl_gene_id)
) %>%
    na.omit()

print(paste0("N = ", 
             nrow(df_step_model), # 11,324
             " genes shared across all studies when we join by ensembl_gene_id")
)

# Get the names of the module columns
module_cols_all <- df_network_data %>% 
    unnest(cols = data) %>% 
    ungroup %>% 
    filter(module != "none") %>%  # removed regulon "none" from Wang et al data
    pull(module) %>% 
    unique
```

## 1.3 Example QQ

Quick look at regression model fit for a single module:

```{r model_diagnostics, fig.width = 6, fig.height = 5}
## QQplot
example_model <- lm(z_fdr ~ `M2-gene` + z_union_cds + z_log_union_cds, data = df_step_model)
p_qq <- autoplot(example_model, 
                 colour = 'steelblue', 
                 size = 0.5, 
                 label = FALSE, 
                 smooth.colour = "black")
p_qq
ggsave(
    filename = file.path(figures_dir, "supplement", "02_00_M2-gene-model-diagnostics.png"), 
    plot = p_qq, 
    width = 6, 
    height = 5, 
    dpi = 300)
```


# --- 2) Identify modules individually and independently associated with ASD risk --- #

## 2.1 STEP 1: Initial pre-screen

```{r initial_model}
# ======================================================================== #
# 0. INITIAL STEP: Run LR model with only length covariates (baseline)
# ======================================================================== #

# Baseline (transcript length) covariates
base_predictors <- c("z_union_cds", "z_log_union_cds")

# Initial model formula (M0: covariates only)
initial_formula <- paste0("z_fdr ~ ", paste(base_predictors, collapse = " + "))

# Run initial model (M0: covariates only)
current_model <- lm(as.formula(initial_formula), data = df_step_model)
```

```{r pre_screen}
# ======================================================================== #
# 1. PRE-SCREENING STEP: Test *individual* association of each module
# ======================================================================== #

pre_screen_results <- purrr::map_dfr(
    .x = module_cols_all,
    .f = function(mod) {
        
        # print(paste0("Running individual association for ", mod))
        
        quoted_module <- paste0("`", mod, "`")
        candidate_predictors <- c(base_predictors, quoted_module)
        candidate_formula <- paste0("z_fdr ~ ", paste(candidate_predictors, collapse = " + "))
        
        # Run model M_module (module + covariates)
        candidate_model <- tryCatch(
            lm(as.formula(candidate_formula),
               data = df_step_model),
            error = function(e) { return(NULL) }
        )
        
        if (is.null(candidate_model)) {
            return(tibble(module = mod, p_value = Inf, success = FALSE))
        }
        
        # Capture summary statistics for each model (representing the individual association of each module)
        sum_mod <- summary(candidate_model)
        ci_mod  <- stats::confint(candidate_model) # !!! Linear CIs not exponentiated !!!
        # Find the row corresponding to the module to extract statistics
        mod_row <- which(rownames(sum_mod$coefficients) %in% c(mod, quoted_module))
        
        p_val <- sum_mod$coefficients[mod_row, "Pr(>|t|)"]
        beta_val <- sum_mod$coefficients[mod_row, "Estimate"]
        conf_low <- ci_mod[mod_row, 1]
        conf_high <- ci_mod[mod_row, 2]    
        
        tibble(module = mod, 
               p_value = p_val, 
               beta = beta_val, 
               conf_low = conf_low, 
               conf_high = conf_high
               )
        
    }
)

# Adjust p-values for screening
df_individual_assoc <- pre_screen_results %>% 
    mutate(p_adj = p.adjust(p_value, method = "BH")) %>% 
    arrange(p_adj) %>% 
    
    # add module label
    left_join(
        df_network_data %>% 
            unnest(cols = c(data)) %>% 
            dplyr::select(module, module_label) %>% 
            unique
    ) %>% 
    dplyr::select(study, module, module_label, beta, conf_low, conf_high, p_value, p_adj)

# Filter modules that pass the initial threshold and extract for forward selection screening
candidate_screen_p <- 0.05
candidate_modules <- df_individual_assoc %>%
    dplyr::filter(p_adj < candidate_screen_p & beta > 0) %>% # include only those that were individually (& positively) associated with ASD risk *after* correction for multiple comparisons
    dplyr::pull(module)

print(paste0("N = ", 
             length(candidate_modules), 
             " modules are significantly associated with ASD and will be tested for independent association in Step 2")
)
```


## 2.2 STEP 2: Stepwise conditional linear model

Write the function to run a stepwise linear analysis \n
Only include modules that passed the pre-screening process (i.e., are marginally associated with ASD status) \n
Report independent association statistics for all modules that survived the forward screening process \n

```{r stepwise_forward_selection}
# Initialize for stepwise loop
# The baseline model is refit here, not just in the chunk above, because the loop
# overwrites current_model: re-running this chunk without also re-running the
# initial_model chunk would otherwise start the search from the previous run's
# final model and admit a module on a non-nested (negative df) F-test
current_model <- lm(as.formula(initial_formula), data = df_step_model)
current_predictors <- base_predictors
remaining_modules <- candidate_modules
selected_modules <- c()

# Set threshold for independent association
# Loop will stop if no modules meet this threshold
p_significance_threshold <- 0.001

# Run loop
repeat{
    if (length(remaining_modules) == 0) break
    
    test_results <- purrr::map_dfr(
        .x = remaining_modules,
        .f = function(mod) {
            quoted_module <- paste0("`", mod, "`")
            tmp_predictors <- c(current_predictors, quoted_module)
            candidate_formula <- paste0("z_fdr ~ ", paste(tmp_predictors, collapse = " + "))
            
            candidate_model <- tryCatch(
                lm(as.formula(candidate_formula), 
                   data = df_step_model),
                error = function(e) { return(NULL) }
            )
            if (is.null(candidate_model)) return(tibble(module = mod, p_value = Inf))
            
            # Run LRT to see if module adds value to the existing model (use an F-test to compare)
            lrt_res <- stats::anova(current_model, candidate_model)
            tibble(module = mod, 
                   p_value = lrt_res$`Pr(>F)`[2], 
                   model = list(candidate_model)
            )
        }
    )
    
    # select module with the lowest LRT p-value
    best_candidate <- test_results %>% dplyr::slice_min(p_value, n = 1, with_ties = FALSE)
    
    # add module to our final set if its p-value is below the threshold, else end the loop
    if (nrow(best_candidate) > 0 && best_candidate$p_value < p_significance_threshold) {
        selected_modules <- c(selected_modules, best_candidate$module) # Mark as an independent signal
        current_predictors <- c(current_predictors, paste0("`", best_candidate$module, "`"))
        remaining_modules <- setdiff(remaining_modules, best_candidate$module)
        current_model <- best_candidate$model[[1]]
        message(paste("Added independent module:", 
                      best_candidate$module, 
                      " (LRT P =", 
                      format.pval(best_candidate$p_value), 
                      ")")
                )
    } else {
        break
    }
}
```


## 2.3 Extract final results

```{r final_results}
df_stepwise_res <- df_individual_assoc %>%
    dplyr::filter(module %in% selected_modules) %>%
    dplyr::select(module, module_label, beta, conf_low, conf_high, p_value, p_adj)
```


# --- 3) Format results --- #

## 3.1 Add module size and standardized beta

```{r format_stepwise_res}
# Standardize beta across modules using SD of the outcome variable (Cohen's d interpretation)
sd_z_fdr <- sd(df_step_model$z_fdr)

# Determine module sizes; count(study, module) preserves the study each module belongs to
df_gene_set_size <- df_network_data %>%
    unnest(cols = c(data)) %>%
    count(study, module)

# Filter pre_screen_results to only include those that survived stepwise
df_stepwise_res2 <- df_stepwise_res %>%
    
    # positively associated modules only at FDR threshold
    # filter(beta > 0 & p_adj < 0.001) %>% 
    
    # add gene set size info
    left_join(df_gene_set_size, by = join_by(module)) %>% 
    distinct() %>% 
    
    # add standardized beta
    mutate(beta_std = beta / sd_z_fdr) %>%   # standardized beta: SD units of ASD risk per module membership
    arrange(p_adj) %>% 
    
    # rearrange
    dplyr::select(study, module, module_label, n, beta, beta_std, everything())
```


## 3.1 Identify modules whose "hub" (or transcription factor) is also a risk gene

```{r hub_risk_gene}
# Pull ASD risk gene list (FDR < 0.001)
asd_risk_genes <- df_asd_rarevar_zscore %>% 
    filter(FDR < 0.001 & !FLAG) %>% 
    pull(ensembl_gene_id)

# Identify regulons where the transcription factor is a risk gene
risk_gene_regulons <- df_network_data %>% 
    filter(study %in% c("Wamsley", "Wang")) %>%  # filter for regulons only
    unnest(cols = c(data)) %>% 
    filter(
        gene_name == str_remove(module_label, " .*") &
            ensembl_gene_id %in% asd_risk_genes
    ) %>%
    pull(module) %>% 
    unique

# Identify co-expression modules whose top hub by kME is a risk gene
df_mod_asd_hubs <- df_module_kme %>%
    filter(module == kME_module) %>% # ensure the kME value corresponds to the module the gene/isoform is in
    group_by(module) %>%
    arrange(module, -kME) %>%
    mutate(rank = row_number()) %>%
    filter(rank == 1) %>% # filter for the single top hub gene by kME in each module
    filter(ensembl_gene_id %in% asd_risk_genes)
risk_gene_mods <- df_mod_asd_hubs %>% 
    pull(module) %>% 
    unique %>% 
    as.character()

# Identify which of these modules / regulons with 'hub' risk genes was also significant in the analysis
sig_risk_gene_mods <- intersect(df_stepwise_res2$module, c(risk_gene_regulons, risk_gene_mods))

print(
    paste0(
        "Modules where hub or TF is risk gene: ",
        paste0(sig_risk_gene_mods, collapse = ", ")
    )
)
```

## 3.2 Combine the results into one tibble, including gene set (module or regulon) size

```{r combine_results}
# Combine model results
df_final_res <- df_stepwise_res2 %>%
    
    # identify if hub / TF is a risk gene
    mutate(hub_is_risk_gene = module %in% sig_risk_gene_mods) %>% 
    
    # select & reorder variables
    dplyr::select(study, module, module_label, n, hub_is_risk_gene, everything())

# Save full results
l_stepwise_res <- list(
    input_data = df_step_model,
    lm_results = df_final_res
)
```

## 3.3 Save results

```{r save_res}
output_path <- file.path(notebook_out_dir, 
                         paste0("02_00_stepwise-linear-res-step1p", candidate_screen_p, 
                                "-step2p", p_significance_threshold, 
                                ".RDS")
)
saveRDS(l_stepwise_res, file = output_path)
```


# --- 4) Export tables --- #

## 4.1 Table S5a: Significant module stats along with risk genes in module

```{r risk_gene_table}
# Get risk genes in each module
df_sig_module_risk_genes <- df_network_data %>% 
    unnest(cols = c(data)) %>% 
    filter(module %in% df_final_res$module & 
               ensembl_gene_id %in% asd_risk_genes) %>% 
    group_by(study, module) %>%
    dplyr::select(study, module, module_label, ensembl_gene_id, gene_name) %>% 
    arrange(gene_name) %>% 
    distinct() %>% 
    summarise(
        n_risk_genes = n(),
        risk_genes_in_module = str_c(gene_name, collapse = ", "),
        .groups = "drop"
    ) %>% 
    distinct()

# Format data to get list of risk genes (FDR < 0.001) associated with each module
df_sig_mod_res <- df_final_res %>% 
   left_join(
       df_sig_module_risk_genes
   ) %>% 
    dplyr::rename(module_n = n)
    
    
# Write to xlsx
df_sig_mod_res %>% 
    write_csv(file.path(notebook_out_dir, 
                        paste0("02_00_sig-module-res-step1p", candidate_screen_p, 
                               "-step2p", p_significance_threshold, 
                               ".csv"))
              )
```

## 4.2 Table S5b: Marginal stats for all modules

```{r marginal_stats_table}
df_individual_assoc %>%
    write_csv(file.path(notebook_out_dir, "02_00_marginal-stats-all-modules.csv"))
```

## 4.3 Co-expression modules whose top hub by kME is an ASD risk gene

```{r hub_risk_genes}
asd_risk_genes <- df_asd_rarevar_zscore %>% 
    filter(FDR < 0.001 & !FLAG) %>% 
    pull(ensembl_gene_id) %>% 
    unique
df_hub_risk_genes <- df_module_kme %>% 
    filter(module == kME_module & module %in% df_stepwise_res$module) %>%
    dplyr::select(-kME_module) %>% 
    group_by(module) %>% 
    
    # filter for the single top hub gene by kME in each module
    arrange(module, -kME) %>%
    mutate(kME_rank = row_number()) %>%
    filter(kME_rank == 1) %>%
    ungroup %>% 
    
    # filter for ASD risk genes
    filter(ensembl_gene_id %in% asd_risk_genes) %>% 
    
    # add module label
    left_join(df_stepwise_res %>% dplyr::select(module, module_label)) %>% 
    dplyr::select(study, module, module_label, everything()) %>% 
    
    # add ASD TADA stats
    left_join(
        df_asd_rarevar_zscore %>% dplyr::select(ensembl_gene_id, TADA_FDR = FDR, TADA_BF = BF)
    )
df_hub_risk_genes %>%
    write_csv(file.path(notebook_out_dir, "02_00_module-hub-risk-genes.csv"))
```



# --- 5) Supplement Figure --- #

## 5.1 Step 1 candidate threshold sensitivity

Compare the independently associated (selected) module set across STEP 1 (candidate pre-screen) FDR thresholds (0.05, 0.01, 0.001), holding the STEP 2 conditional threshold fixed at p < 0.001. Tile fill encodes each module's -log10(FDR); white = not selected at that threshold. Rows are ordered by significance (most significant at top). Because the marginal association statistic is identical across runs, a module's shade is constant across the thresholds at which it is selected.

```{r read_step1_sensitivity}
# Read the three STEP 1 threshold runs; lm_results = final significant modules for each
step1_files <- c(
    "0.05"  = "02_00_stepwise-linear-res-step1p0.05-step2p0.001.RDS",
    "0.01"  = "02_00_stepwise-linear-res-step1p0.01-step2p0.001.RDS",
    "0.001" = "02_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS"
)

df_step1_sensitivity <- imap_dfr(step1_files, function(file, step1_thresh) {
    readRDS(file.path(notebook_out_dir, file))$lm_results %>%
        dplyr::transmute(module, module_label, step1_threshold = step1_thresh, p_value, p_adj)
})
```

```{r plot_step1_sensitivity, fig.height = 5.5, fig.width = 3.2}
threshold_levels <- c("0.05", "0.01", "0.001")   # lenient -> stringent (left -> right)

# module -> display label
module_labels <- df_step1_sensitivity %>% distinct(module, module_label)

# Order modules by significance (most significant at top of y-axis)
module_order <- df_step1_sensitivity %>%
    group_by(module) %>%
    summarise(min_p_adj = min(p_adj), .groups = "drop") %>%
    arrange(desc(min_p_adj)) %>%   # least significant first -> lands at bottom of y-axis
    pull(module)

# Complete the module x threshold grid; neg_log10_fdr is NA where a module was not selected
df_step1_heatmap <- df_step1_sensitivity %>%
    mutate(neg_log10_fdr = -log10(p_adj)) %>%
    dplyr::select(module, step1_threshold, neg_log10_fdr) %>%
    complete(step1_threshold = threshold_levels, module = module_order) %>%
    mutate(
        step1_threshold = factor(step1_threshold, levels = threshold_levels),
        module          = factor(module, levels = module_order)
    )

p_step1_sensitivity <- ggplot(df_step1_heatmap, aes(x = step1_threshold, y = module, fill = neg_log10_fdr)) +
    geom_tile(color = "grey85", linewidth = 0.3) +
    scale_fill_gradient(low = "#C4C9EC", high = "#3B4CC0", na.value = "white",
                        name = expression(-log[10](FDR))) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(labels = deframe(module_labels), expand = c(0, 0)) +
    labs(
        x = "Candidate module threshold (FDR)",
        y = NULL,
        title = "Independently associated modules\nacross candidate thresholds"
    ) +
    theme(
        axis.text.y  = element_text(size = 6),
        panel.grid   = element_blank(),
        panel.border = element_blank(),
        legend.key.size = unit(0.3, "cm"),
        legend.margin = margin(l = -5),
        plot.title = element_text(face = "bold")
    )

p_step1_sensitivity

ggsave(
    filename = file.path(figures_dir, "supplement", "02_00_stepwise_linear_step1p_sensitivity.pdf"),
    plot = p_step1_sensitivity,
    height = 5.5,
    width = 3.2,
    dpi = 300
)
```
