---
title: "Module-level O/E enrichment analyses"
description: "Observed/expected de novo mutation burden for each significant module from the stepwise conditional analysis, across proband and sibling cohorts (ASD-D, ASD-DA, all probands, siblings)."
date: 2026-07-10
format: html
---

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set output directory for notebook analysis
notebook_out_dir <- file.path(analysis_dir, "06_gene_set_DDID_stratification")
```

Load data

```{r load_data}
# ASD rare variant data
df_asd_rarevar_zscore <- readRDS(file.path(analysis_dir, "00_data_format", "00_01_ASD-rarevar-stats.RDS"))

# Network data (gene-module assignments & kME)
df_network_data <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_gene-networks-all.RDS"))
df_module_kme <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_co-expression-kME.RDS"))

# Significant modules from stepwise conditional analysis
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "01_module_identification", "01_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results

# Module topic assignments (top topic per module, with colors)
df_module_topics <- readRDS(file.path(analysis_dir, "02_module_enrichments", "02_00_topic-module-gamma.RDS")) %>%
    filter(top_topic == "yes") %>%
    select(module, topic_label, topic_color)

# Background mutation rates + full proband/sibling counts
df_asd_dnm <- read_tsv(
    file.path(data_dir, "full_results_wcounts_2025-10-08.txt"),
    show_col_types = FALSE
) %>%
    clean_names() %>%
    mutate(mu_total = mu_lof + mu_mis2 + mu_mis1)

# ASD-D / ASD-DA proband counts (ASC)
df_counts_asd_d   <- read_xlsx(file.path(data_dir, "DDID", "counts_asd_ddid.xlsx"))   %>% clean_names()
df_counts_asd_da <- read_xlsx(file.path(data_dir, "DDID", "counts_asd_noddid.xlsx")) %>% clean_names()

# Kaplanis DDD counts (all ASD-D; sex-stratified Male/Female columns)
# N = 1,194 (adds to ASC ASD-D N = 12,647 → total ASD-D N = 13,841)
df_counts_kaplanis <- read_tsv(
    file.path(data_dir, "DDID", "Kaplanis_DDD_ASD_counts_by_gene_new_mis_cats_2025-04-09.txt"),
    show_col_types = FALSE
) %>%
    clean_names()
```


# --- 1) O/E enrichment across all modules --- #

The O/E analysis is run identically over every module; the module-level figures in
Section 1.6 filter to the stepwise-significant set (`df_oe` / `df_pairwise`), while the
all-module figures (Section 3) use the full result (`df_oe_all` / `df_pairwise_all`).

## 1.1 Format data

Combine counts into one table keyed by gene name

```{r format_counts}
# Proband and sibling counts are in the ASD DNM file; ASD-D/ASD-DA split from separate files
df_counts <- df_asd_dnm %>%
    select(gene, mu_total, mu_lof, mu_mis2, mu_mis1,
           ptv_proband, mis2_proband, mis1_proband,
           ptv_sibling, mis2_sibling, mis1_sibling) %>%
    mutate(
        count_proband = ptv_proband + mis2_proband + mis1_proband,
        count_sibling = ptv_sibling + mis2_sibling + mis1_sibling
    ) %>%
    left_join(
        # ASC ASD-D counts + Kaplanis DDD counts (summed Male + Female per variant type)
        df_counts_asd_d %>%
            mutate(count_asd_d_proband = ptv_proband + mis2_proband + mis1_proband) %>%
            select(gene, count_asd_d_proband) %>%
            left_join(
                df_counts_kaplanis %>%
                    mutate(count_kaplanis = (ptv_proband_male + ptv_proband_female) +
                                            (mis2_proband_male + mis2_proband_female) +
                                            (mis1_proband_male + mis1_proband_female)) %>%
                    select(gene, count_kaplanis),
                by = "gene"
            ) %>%
            mutate(count_asd_d_proband = count_asd_d_proband + replace_na(count_kaplanis, 0)) %>%
            select(gene, count_asd_d_proband),
        by = "gene"
    ) %>%
    left_join(
        df_counts_asd_da %>%
            mutate(count_asd_da_proband = ptv_proband + mis2_proband + mis1_proband) %>%
            select(gene, count_asd_da_proband),
        by = "gene"
    )
```

Extract gene members for each significant module from `df_network_data`

```{r build_gene_sets}
# Full individual-association stats (used to annotate each module's tier)
df_all_module_stats <- read_csv(
    file.path(analysis_dir, "01_module_identification", "01_00_marginal-stats-all-modules.csv")
) %>%
    clean_names()

# Every module (excluding the unassigned "none"), annotated by ASD-association tier
# The O/E analysis is run identically across ALL modules
# the module figures filter this to the stepwise-significant set
df_all_modules <- df_network_data %>%
    unnest(cols = data) %>%
    ungroup() %>%
    filter(module != "none") %>%
    select(study, module, module_label) %>%
    distinct() %>%
    left_join(df_all_module_stats %>% select(module, beta, p_adj), by = "module") %>%
    mutate(
        tier = case_when(
            module %in% df_stepwise_res$module        ~ "Stepwise significant",
            !is.na(p_adj) & p_adj < 0.001 & beta > 0 & !(module %in% df_stepwise_res$module)   ~ "Marginal (ASD-enriched)",
            !is.na(p_adj) & p_adj < 0.001 & beta < 0 & !(module %in% df_stepwise_res$module)   ~ "Marginal (ASD-depleted)",
            TRUE                                       ~ "Not associated"
        ),
        tier = factor(tier, levels = c("Stepwise significant", "Marginal (ASD-enriched)",
                                       "Marginal (ASD-depleted)", "Not associated"))
    )

# Gene members of every module (used for O/E and the per-module scatter)
df_module_genes <- df_network_data %>%
    unnest(cols = data) %>%
    ungroup() %>%
    filter(module != "none") %>%
    select(study, module, gene_name, ensembl_gene_id) %>%
    distinct()
```

## 1.2 O/E analysis helper functions

```{r helpers}
# --- O/E with Garwood exact Poisson CI --- #
# Garwood (1936) exact Poisson confidence interval for the O/E ratio
# p-value: exact Poisson test of H0: obs ~ Poisson(exp), i.e. O/E = 1
oe_garwood <- function(obs, exp) {
    oe   <- obs / exp
    ci_l <- if (obs > 0) qchisq(0.025, 2 * obs) / (2 * exp) else 0
    ci_h <- qchisq(0.975, 2 * (obs + 1)) / (2 * exp)
    p    <- if (obs > 0 && exp > 0)
                poisson.test(obs, T = exp, alternative = "two.sided")$p.value
            else NA_real_
    list(oe = oe, ci_l = ci_l, ci_h = ci_h, p = p)
}


# --- Pairwise conditional binomial test --- #

pairwise_binom <- function(obs1, exp1, obs2, exp2) {
    obs1 <- as.numeric(obs1)[1]; exp1 <- as.numeric(exp1)[1]
    obs2 <- as.numeric(obs2)[1]; exp2 <- as.numeric(exp2)[1]

    # Completely uninformative: missing inputs or both groups have zero observed
    if (any(is.na(c(obs1, obs2, exp1, exp2))) || (obs1 == 0 && obs2 == 0))
        return(tibble(rate_ratio = NA_real_, ci_l = NA_real_, ci_h = NA_real_, p = NA_real_))

    rel <- (obs1 / exp1) / (obs2 / exp2)   # = 0 when obs1 = 0; Inf when obs2 = 0
    p0  <- exp1 / (exp1 + exp2)
    p   <- binom.test(obs1, obs1 + obs2, p0, alternative = "two.sided")$p.value

    # Log-scale SE requires both counts > 0; return rate_ratio + p but no CI when one is zero
    if (obs1 == 0 || obs2 == 0)
        return(tibble(rate_ratio = rel, ci_l = NA_real_, ci_h = NA_real_, p = p))

    se <- sqrt(1 / obs1 + 1 / obs2)
    tibble(
        rate_ratio = rel,
        ci_l   = exp(log(rel) - qnorm(0.975) * se),
        ci_h   = exp(log(rel) + qnorm(0.975) * se),
        p      = p
    )
}

# Cohort sample sizes
N_cohorts <- list(
    asd_da_proband = 24839L,
    asd_d_proband   = 13841L,
    proband        = 38680L,
    sibling        =  9567L
)
```

## 1.3 Calculate O/E per module per cohort

```{r oe_per_module}
cohort_spec <- tribble(
    ~cohort,           ~count_col,            ~n,
    "proband",         "count_proband",        N_cohorts$proband,
    "sibling",         "count_sibling",        N_cohorts$sibling,
    "asd_d_proband",    "count_asd_d_proband",   N_cohorts$asd_d_proband,
    "asd_da_proband",  "count_asd_da_proband", N_cohorts$asd_da_proband
)

# O/E for EVERY module (loops over ~1,150 modules, so it's cached)
oe_cache <- file.path(notebook_out_dir, "06_00_all-modules-OE.RDS")

if (file.exists(oe_cache)) {
    df_oe_all <- readRDS(oe_cache)
} else {
    df_oe_all <- map2_dfr(
        .x = cohort_spec$cohort,
        .y = cohort_spec$count_col,
        .f = function(cohort_name, cnt_col) {

            n_cohort <- cohort_spec %>% filter(cohort == cohort_name) %>% pull(n)

            df_all_modules %>%
                select(study, module, module_label) %>%
                pmap_dfr(function(study, module, module_label) {

                    genes_in_mod <- df_module_genes %>%
                        filter(module == !!module) %>%
                        pull(gene_name) %>%
                        unique()

                    m <- df_counts %>% filter(gene %in% genes_in_mod)

                    obs  <- sum(m[[cnt_col]], na.rm = TRUE)
                    exp  <- sum(m$mu_total,   na.rm = TRUE) * 2 * n_cohort
                    ci   <- oe_garwood(obs, exp)

                    tibble(
                        study     = study,
                        module    = module,
                        module_label = module_label,
                        cohort    = cohort_name,
                        n_genes   = length(genes_in_mod),
                        n_matched = nrow(m),
                        n_cohort  = n_cohort,
                        obs       = obs,
                        exp       = round(exp, 4),
                        oe        = ci$oe,
                        ci_l      = ci$ci_l,
                        ci_h      = ci$ci_h,
                        p         = ci$p
                    )
                })
        }
    ) %>%
        group_by(cohort) %>%
        mutate(p_fdr = p.adjust(p, method = "BH")) %>%
        ungroup() %>%
        arrange(p)

    saveRDS(df_oe_all, oe_cache)
}
```

## 1.4 Cohort pairwise comparisons within each module

For each module, compare:
- Proband vs Sibling (overall enrichment)
- ASD-D_Proband vs ASD-DA_Proband (ASD-D specificity)
- ASD-D_Proband vs Sibling
- ASD-DA_Proband vs Sibling

```{r cohort_pairs}
cohort_pairs <- list(
    c("proband",        "sibling"),
    c("asd_d_proband",   "asd_da_proband"),
    c("asd_d_proband",   "sibling"),
    c("asd_da_proband", "sibling")
)

# Pairwise comparisons for EVERY module (cached). FDR is corrected across all modules
# and cohort contrasts; the stepwise subset (df_pairwise) is derived below for 1.6.
pairwise_cache <- file.path(notebook_out_dir, "06_00_all-modules-pairwise-RR.RDS")

if (file.exists(pairwise_cache)) {
    df_pairwise_all <- readRDS(pairwise_cache)
} else {
    df_pairwise_all <- df_all_modules %>%
        select(study, module, module_label) %>%
        pmap_dfr(function(study, module, module_label) {
            map_dfr(cohort_pairs, function(pair) {
                ga <- pair[1]; gb <- pair[2]

                sa <- df_oe_all %>% filter(module == !!module, cohort == ga)
                sb <- df_oe_all %>% filter(module == !!module, cohort == gb)

                pw <- pairwise_binom(sa$obs, sa$exp, sb$obs, sb$exp)

                bind_cols(
                    tibble(
                        study      = study,
                        module     = module,
                        module_label = module_label,
                        comparison = paste(ga, "vs", gb)
                    ),
                    pw
                )
            })
        }) %>%
        mutate(p_fdr = p.adjust(p, method = "BH"))

    saveRDS(df_pairwise_all, pairwise_cache)
}
```

## 1.5 Export tables

```{r export_tables}
df_oe_all %>%
    arrange(study, module, cohort) %>%
    write_csv(file.path(notebook_out_dir, "06_00_module_OE_ratios.csv"))

df_pairwise_all %>%
    arrange(study, module, comparison) %>%
    write_csv(file.path(notebook_out_dir, "06_00_module_cohort_pairwise.csv"))
```


# --- 2) GO enrichment on ASD-DA-biased modules --- #

Modules significantly biased toward ASD-DA probands (ASD without DD/ID 
over ASD-D probands — `asd_d_proband vs asd_da_proband` relative risk < 1 at
FDR < 0.05 — are tested for GO over-representation (Fisher's exact, one-sided), to ask
which biological programs distinguish the ASD-DA-preferential modules.

## 2.1 Define ASD-DA-biased modules

```{r define_asd_da_modules}
sig_thresh <- 0.05

# ASD-DA-biased = significantly enriched in ASD-DA vs ASD-D probands (rate_ratio < 1)
# --- OR --- #
# Enriched in ASD-DA relative to siblings but NOT ASD-D relative to siblings

# Pull the relevant comparisons into one row per module (wide), so criteria that live on
# different rows of the long pairwise table can be combined
df_mod_rr <- df_pairwise_all %>%
    filter(comparison %in% c("asd_d_proband vs asd_da_proband",
                             "asd_d_proband vs sibling",
                             "asd_da_proband vs sibling")) %>%
    mutate(key = recode(comparison,
        "asd_d_proband vs asd_da_proband" = "contrast",
        "asd_d_proband vs sibling"        = "asd_d_sib",
        "asd_da_proband vs sibling"       = "asd_da_sib")) %>%
    select(module, key, rate_ratio, p_fdr) %>%
    pivot_wider(names_from = key, values_from = c(rate_ratio, p_fdr))

# A module is ASD-DA-preferential if EITHER:
#   (A) direct contrast — significantly higher burden in ASD-DA than ASD-D probands
#       (contrast rate ratio < 1 at FDR < 0.05); the cleaner test of differential burden.
#   (B) vs-siblings specificity — significantly enriched over siblings in ASD-DA but NOT
#       in ASD-D. Caveat: non-significance in ASD-D vs siblings is not evidence of no
#       ASD-D enrichment (could be underpowered), so (B) is descriptive, weaker than (A).
df_asd_da_biased <- df_mod_rr %>%
    filter(!str_detect(module, "M0_grey")) %>%
    mutate(
        sig_contrast   = coalesce(rate_ratio_contrast   < 1 & p_fdr_contrast   < sig_thresh, FALSE),
        enr_asd_da_sib = coalesce(rate_ratio_asd_da_sib > 1 & p_fdr_asd_da_sib < sig_thresh, FALSE),
        enr_asd_d_sib  = coalesce(rate_ratio_asd_d_sib  > 1 & p_fdr_asd_d_sib  < sig_thresh, FALSE),
        sig_specific   = enr_asd_da_sib & !enr_asd_d_sib,
        criterion = case_when(
            sig_contrast & sig_specific ~ "both",
            sig_contrast                ~ "direct contrast",
            sig_specific                ~ "ASD-DA-specific vs siblings",
            TRUE                        ~ NA_character_
        ),
        # single rate_ratio / p_fdr (the direct contrast) for downstream ordering + joins
        rate_ratio = rate_ratio_contrast,
        p_fdr      = p_fdr_contrast
    ) %>%
    filter(!is.na(criterion))

cat(sprintf("%d ASD-DA-preferential modules (%d direct contrast, %d ASD-DA-specific vs sib, %d both)\n",
            nrow(df_asd_da_biased),
            sum(df_asd_da_biased$criterion == "direct contrast"),
            sum(df_asd_da_biased$criterion == "ASD-DA-specific vs siblings"),
            sum(df_asd_da_biased$criterion == "both")))
```

## 2.2 GO over-representation

```{r prep_go_ora}
# GO term names + gene annotations (org.Hs.eg.db / GO.db loaded in setup.R)
go_terms <- AnnotationDbi::select(
    GO.db, keys = keys(GO.db), columns = c("TERM", "ONTOLOGY"), keytype = "GOID"
) %>% as_tibble() %>% clean_names() %>% dplyr::rename(go_id = goid)

go_mapping <- AnnotationDbi::select(
    org.Hs.eg.db, keys = keys(org.Hs.eg.db, keytype = "ENSEMBL"),
    columns = c("GO", "ONTOLOGY"), keytype = "ENSEMBL"
) %>% as_tibble() %>% clean_names() %>%
    dplyr::rename(go_id = go, ensembl_gene_id = ensembl) %>%
    dplyr::select(-evidence)

go_background <- go_mapping %>% pull(ensembl_gene_id) %>% unique()

# GO term → gene list (keep terms with 5–1000 genes)
go_genes_list <- go_mapping %>%
    group_by(go_id) %>%
    summarise(genes = list(unique(ensembl_gene_id)), .groups = "drop") %>%
    filter(lengths(genes) >= 5 & lengths(genes) <= 1000) %>%
    deframe()

# Per-study background + per-module Ensembl gene sets for the ASD-DA-biased modules
df_network_unnested <- df_network_data %>% unnest(cols = data) %>% ungroup()

study_genes_list <- df_network_unnested %>%
    group_by(study) %>%
    summarise(study_genes = list(unique(ensembl_gene_id)), .groups = "drop") %>%
    deframe()

df_asd_da_gene_sets <- df_network_unnested %>%
    filter(module %in% df_asd_da_biased$module) %>%
    group_by(study, module) %>%
    summarise(ensembl_ids = list(unique(ensembl_gene_id)), .groups = "drop")
```

```{r run_go_ora}
# Fisher's exact ORA of each ASD-DA-biased module against every GO term it overlaps
# by > 1 gene. Universe = study network ∩ GO background; BH-corrected within module.
df_go_res <- pmap_dfr(
    list(df_asd_da_gene_sets$study, df_asd_da_gene_sets$module, df_asd_da_gene_sets$ensembl_ids),
    function(study_i, module_i, mod_genes) {
        go_overlap <- names(go_genes_list)[
            map_lgl(go_genes_list, ~ length(intersect(.x, mod_genes)) > 1)
        ]
        map_dfr(go_overlap, function(gid) {
            ORA_fisher(
                gene_list1       = mod_genes,
                gene_list2       = go_genes_list[[gid]],
                background_list1 = study_genes_list[[study_i]],
                background_list2 = go_background,
                alternative      = "greater"
            ) %>%
                mutate(study = study_i, module = module_i, go_id = gid)
        })
    }
) %>%
    left_join(go_terms, by = "go_id") %>%
    group_by(study, module) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"), .before = n_overlap) %>%
    ungroup() %>%
    dplyr::rename(n_module = n_list1, n_pathway = n_list2) %>%
    mutate(pct_module = n_overlap / n_module * 100) %>%
    left_join(df_asd_da_biased %>% select(module, rate_ratio, criterion), by = "module") %>%
    dplyr::select(study, module, go_id, term, ontology, everything()) %>%
    arrange(module, p_value)
```

```{r export_go}
saveRDS(df_go_res, file.path(analysis_dir, "06_gene_set_DDID_stratification",
                             "06_00_ASD-DA-biased-GO-enrichments.RDS"))

df_go_res %>%
    filter(p_value < 0.05) %>%
    write_csv(file.path(analysis_dir, "06_gene_set_DDID_stratification",
                         "06_00_ASD-DA-biased-GO-enrichments.csv"))
```

## 2.3 Top GO terms per module (not currently included in supp mat)

```{r plot_go_dot, fig.height = 7, fig.width = 9}
# Modules ordered by ASD-DA bias (most biased = smallest rate_ratio first)
mod_order_go <- df_asd_da_biased %>% arrange(rate_ratio) %>% pull(module)

top_go <- df_go_res %>%
    filter(p_adj < 0.05, !is.na(term)) %>%
    group_by(module) %>%
    slice_min(p_adj, n = 5, with_ties = FALSE) %>%
    ungroup()

p_go_dot <- top_go %>%
    mutate(module = factor(module, levels = mod_order_go)) %>%
    ggplot(aes(x = module, y = reorder(term, -log10(p_adj)),
               size = pct_module, color = -log10(p_adj))) +
    geom_point() +
    scale_color_viridis_c(option = "plasma", name = "-log10(FDR)") +
    scale_size_continuous(range = c(1, 5), name = "% module genes") +
    facet_grid(ontology ~ ., scales = "free", space = "free") +
    labs(x = NULL, y = NULL, title = "Top GO terms for ASD-DA-biased modules") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text = element_text(face = "bold"), strip.background = element_blank())

p_go_dot
```


# --- 3) Supplemental plots --- #

## 3.1 Panel A: Forest plot of all cohorts O/E with ASD-D:ASD-DA significance

```{r oe_forest, fig.height = 2.7, fig.width = 3.5}
# Filter for stepwise significant modules to plot
df_oe_plot <- df_oe_all %>% 
    filter(module %in% df_stepwise_res$module)

# Module order: ASD-D O/E descending
oe_mod_order <- df_oe_plot %>% 
    filter(cohort == "asd_d_proband") %>%
    arrange(desc(oe)) %>%
    pull(module_label)

# Filter for stepwise significant modules
df_pairwise_plot <- df_pairwise_all %>%
    filter(module %in% df_stepwise_res$module)

# Significance labels from the ASD-D vs ASD-DA pairwise comparison
df_asd_d_asd_da_sig <- df_pairwise_plot %>% 
    filter(comparison == "asd_d_proband vs asd_da_proband") %>%
    mutate(
        sig_lab = case_when(
            p_fdr < 0.001 ~ "***",
            p_fdr < 0.01  ~ "**",
            p_fdr < 0.05  ~ "*",
            TRUE          ~ ""
        )
    ) %>%
    select(module_label, sig_lab)

# Star x-position = max CI upper bound across all four cohorts
df_oe_star <- df_oe_plot %>%
    group_by(module_label) %>%
    summarise(x_star = max(ci_h, na.rm = TRUE), .groups = "drop") %>%
    left_join(df_asd_d_asd_da_sig, by = "module_label") %>%
    filter(sig_lab != "") %>%
    mutate(module_label = factor(module_label, levels = rev(oe_mod_order)))

# Assign cohort colors
oe_cohort_colors <- c(
    "asd_d_proband" = "#C0392B", 
    "asd_da_proband" = "#4472C4", 
    "sibling" = "grey40"
)

# Plot
p_oe_forest <- df_oe_plot %>% 
    filter(cohort != "proband") %>% 
    
    mutate(
        module_label = factor(module_label, levels = rev(oe_mod_order)),
        cohort = factor(cohort, levels = names(oe_cohort_colors))
    ) %>%
    
    ggplot(aes(x = oe, y = module_label, color = cohort)) +
    geom_vline(xintercept = 1, linetype = "dashed",
               color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h),
        position = position_dodge(width = 0.7),
        size = 0.075, linewidth = 0.25
    ) +
    geom_text(
        data        = df_oe_star,
        aes(x = x_star, y = module_label, label = sig_lab),
        inherit.aes = FALSE,
        hjust = -0.15, vjust = 0.75, size = 2.8, color = "black"
    ) +
    scale_color_manual(
        values = oe_cohort_colors,
        labels = c("ASD-D probands", "ASD-DA probands", "Siblings"),
        name   = NULL
    ) +
    labs(
        x     = "O/E ratio",
        y     = NULL,
        title = "DNM burden across modules"
    ) +
    coord_cartesian(clip = "off") +
    theme(
        legend.position = c(0.75, 0.3),
        #legend.position = "none", # legend will be in Yundan's figure
        legend.margin   = margin(t = -2, b = -5),
        legend.title    = element_text(size = 6),
        legend.text     = element_text(size = 5.5),
        axis.text.x     = element_text(size = 5.5),
        axis.text.y     = element_text(size = 6),
        axis.title      = element_text(size = 6),
        panel.border    = element_blank(),
        axis.line       = element_line(linewidth = 0.25),
        plot.title      = element_text(size = 7, face = "bold")
    )

p_oe_forest
```

## 3.2 Panel B: Scatter plot of ASD-DA vs siblings (x) × ASD-D vs siblings (y)

```{r rr_scatter, fig.height = 2.7, fig.width = 4}
# Genes per module with at least one mutation in ASD-D or ASD-DA probands
df_module_n_mut <- df_module_genes %>%
    left_join(df_counts %>% select(gene, count_asd_d_proband, count_asd_da_proband),
              by = c("gene_name" = "gene")) %>%
    filter((count_asd_d_proband + count_asd_da_proband) > 0) %>%
    group_by(module) %>%
    summarise(n_genes_mut = n_distinct(gene_name), .groups = "drop")

# Pivot wider & calculate # of genes with at least one mutation in the modules
df_scatter <- df_pairwise_plot %>%
    filter(comparison %in% c("asd_d_proband vs sibling", "asd_da_proband vs sibling")) %>%
    select(module, module_label, comparison, rate_ratio, ci_l, ci_h) %>%
    mutate(group = if_else(comparison == "asd_d_proband vs sibling", "asd_d", "asd_da")) %>%
    select(-comparison) %>%
    pivot_wider(
        id_cols     = c(module, module_label),
        names_from  = group,
        values_from = c(rate_ratio, ci_l, ci_h)
    ) %>%
    left_join(df_module_topics,  by = "module") %>%
    left_join(df_module_n_mut,   by = "module")

# Labels for full plot only (outliers)
full_plot_labels <- c("TCF4 (+)", "BHLHE22 (-)", "ZNF281 (+)")

# Labels for zoomed plot only (everything except outliers)
zoom_labels <- c(
    "MEF2C (+)",
    "CUX1 (+)",
    "SRCIN1",
    "THRB (-)",
    "SOX11 (+)",
    "VEZF1",
    "NR3C1 (+)",
    "EP300",
    "JUND (+)"
)

# Full plot - only outlier labels, no zoom
p_full <- df_scatter %>%
    ggplot(aes(x = rate_ratio_asd_da, y = rate_ratio_asd_d, fill = topic_color)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(size = n_genes_mut), shape = 21, color = "white", stroke = 0.5, alpha = 0.75) +
    ggrepel::geom_text_repel(
        data = . %>% filter(module_label %in% full_plot_labels),
        aes(label = module_label, color = topic_color),
        size               = 2,
        box.padding        = 0.75,
        point.padding      = 0.25,
        force              = 2.5,
        force_pull         = 0.8,
        min.segment.length = 0.0,
        segment.color      = "grey70",
        segment.size       = 0.2,
        max.overlaps       = Inf,
        show.legend        = FALSE
    ) +
    # Draw zoom rectangle
    annotate("rect",
        xmin = 1.2, xmax = 2.5,
        ymin = 1.95,   ymax = 6.3,
        fill = NA, color = "grey70", linewidth = 0.3, linetype = "dashed"
    ) +
    scale_size_continuous(range = c(1.5, 3.5), name = "# genes with ≥1 mutation") +
    scale_y_log10() +
    scale_x_log10() +
    scale_color_identity() +
    scale_fill_identity(
        guide  = guide_legend(order = 2, override.aes = list(size = 3)),
        labels = setNames(df_scatter$topic_label, df_scatter$topic_color),
        name   = NULL
    ) +
    guides(size = guide_legend(order = 1, override.aes = list(color = "black", fill = "grey60"))) +
    labs(x = "ASD-DA vs siblings (log10 RR)", y = "ASD-D vs siblings (log10 RR)") +
    theme(
        legend.text      = element_text(size = 5.5),
        legend.title     = element_text(size = 6, margin = margin(b = 0)),
        axis.text        = element_text(size = 5.5),
        axis.title       = element_text(size = 6)
    )

# Zoomed plot - only cluster labels, grey background
p_zoom <- df_scatter %>%
    filter(rate_ratio_asd_da >= 1.3, rate_ratio_asd_da <= 2.5,
           rate_ratio_asd_d   >= 2,   rate_ratio_asd_d   <= 6.1) %>%
    ggplot(aes(x = rate_ratio_asd_da, y = rate_ratio_asd_d, fill = topic_color)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_point(aes(size = n_genes_mut), shape = 21, color = "white", stroke = 0.5, alpha = 0.75) +
    ggrepel::geom_text_repel(
        data = . %>% filter(module_label %in% zoom_labels),
        aes(label = module_label, color = topic_color),
        size               = 2,
        box.padding        = 0.75,
        point.padding      = 0.25,
        force              = 2.5,
        force_pull         = 0.8,
        min.segment.length = 0.0,
        segment.color      = "grey70",
        segment.size       = 0.2,
        max.overlaps       = Inf,
        show.legend        = FALSE
    ) +
    scale_size_continuous(range = c(1.5, 3.5), guide = "none") +
    scale_y_log10(limits = c(2.1, 6.1)) +
    scale_x_log10(limits = c(1.3, 2.4)) +
    scale_color_identity() +
    scale_fill_identity(guide = "none") +
    labs(x = NULL, y = NULL) +
    theme(
        legend.position = "none",
        axis.text       = element_text(size = 5.5),
        axis.title      = element_text(size = 6),
        panel.background = element_rect(fill = "grey97"),
        panel.border     = element_rect(color = "grey70", linewidth = 0.3, fill = NA)
    )

# Combine with patchwork
p_rr_scatter <- p_full + plot_spacer() + p_zoom +
    plot_layout(nrow = 1, widths = c(0.75, -0.1, 1),
                guides = "collect") &
    plot_annotation(title = "Relative module enrichments") &
    theme(plot.title = element_text(size = 7, face = "bold"),
          legend.position = "bottom",
          legend.direction = "horizontal",
          legend.title.position = "top",
          legend.margin = margin(t = -5)
    )

p_rr_scatter
```

## 3.3 Panel C: Scatter plot of ASD-DA vs siblings (x) × ASD-D vs siblings (y) across all modules

```{r all_modules_scatter, fig.width = 7.09, fig.height = 4.5}
# Color by significance vs siblings:
#   Both    = sig in ASD-D and ASD-DA vs siblings (green)
#   ASD-D    = sig in ASD-D vs siblings only (red)
#   ASD-DA  = sig in ASD-DA vs siblings only (blue)
#   Neither = neither significant (gray)
col_sig_group <- c(
    "Both"       = "#885371",
    "ASD-D only"  = "#C0392B",
    "ASD-DA only"= "#4472C4",
    "Neither"    = "grey75"
)

# Module sizes (total genes per module)
df_all_module_sizes <- df_module_genes %>%
    group_by(module) %>%
    summarise(n_genes = n_distinct(gene_name), .groups = "drop")

# One row per module: rate ratio + significance (FDR < 0.05 & enriched) vs siblings, for
# both ASD-D and ASD-DA, then classify by which contrast(s) are significant.
df_scatter_all <- df_pairwise_all %>%
    filter(comparison %in% c("asd_d_proband vs sibling", "asd_da_proband vs sibling"),
           !str_detect(module, "M0_grey")) %>%                 # drop grey modules
    mutate(group  = if_else(comparison == "asd_d_proband vs sibling", "asd_d", "asd_da"),
           is_sig = coalesce(p_fdr < 0.05 & rate_ratio > 1, FALSE)) %>%
    select(module, module_label, group, rate_ratio, is_sig) %>%
    pivot_wider(names_from = group, values_from = c(rate_ratio, is_sig)) %>%
    left_join(df_all_module_sizes, by = "module") %>%
    mutate(sig_group = case_when(
        is_sig_asd_d  &  is_sig_asd_da ~ "Both",
        is_sig_asd_d  & !is_sig_asd_da ~ "ASD-D only",
        !is_sig_asd_d &  is_sig_asd_da ~ "ASD-DA only",
        TRUE                           ~ "Neither"
    ) %>% factor(levels = names(col_sig_group))) %>%
    # draw "Neither" first so significant points render on top
    arrange(sig_group != "Neither")

# Generate plot
p_scatter_all <- df_scatter_all %>%
    ggplot(aes(x = rate_ratio_asd_da, y = rate_ratio_asd_d)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(fill = sig_group, size = n_genes),
               shape = 21, color = "white", stroke = 0.2) +
    # Label ASD-DA-only significant modules using module_label
    ggrepel::geom_text_repel(
        data = df_scatter_all %>% filter(sig_group == "ASD-DA only"),
        aes(label = module_label),
        color = col_sig_group["ASD-DA only"],
        size = 2, nudge_x = 0.35, direction = "y", hjust = 0,
        box.padding = 0.3, point.padding = 0.2,
        min.segment.length = 0, segment.color = "grey30", segment.size = 0.3,
        max.overlaps = Inf, show.legend = FALSE
    ) +
    scale_fill_manual(
        values = col_sig_group,
        name   = "Significant vs siblings",
        labels = c(
            "Both"    = "ASD-D & ASD-DA",
            "ASD-D only"    = "ASD-D only",
            "ASD-DA only"  = "ASD-DA only",
            "Neither" = "Neither"
        ),
        guide = guide_legend(override.aes = list(size = 3, color = "white"))
    ) +
    scale_size_continuous(
        range = c(1, 4),
        name  = "Module size\n(# genes)",
        guide = guide_legend(
            override.aes = list(fill = "grey50", color = "white")
        )
    ) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
        x     = "ASD-DA vs siblings (log RR)",
        y     = "ASD-D vs siblings (log RR)",
        title = "All module relative enrichments"
    ) +
    theme(
        plot.title           = element_text(size = 7, face = "bold"),
        axis.title           = element_text(size = 6),
        axis.text            = element_text(size = 5.5),
        legend.title         = element_text(size = 6),
        legend.text          = element_text(size = 5.5),
        legend.position      = c(0.01, 0.99),
        legend.justification = c(0, 1),
        legend.box           = "vertical",
        legend.spacing.y     = unit(0.05, "cm"),
        legend.margin        = margin(2, 2, 2, 2),
        legend.background    = element_rect(fill = scales::alpha("white", 0.7),
                                            color = "grey85", linewidth = 0.2),
        legend.key.size      = unit(0.25, "cm")
    )

p_scatter_all
```

## 3.4 Panel D: Box plots of OE distributions

```{r all_modules_oe_box, fig.height = 5, fig.width = 2.5}
dodge_w      <- 0.8
cohort_order <- names(oe_cohort_colors)   # asd_d_proband, asd_da_proband, sibling

# Cohort pairs to test (internal names)
cohort_pairs_box <- list(
    c("asd_d_proband",  "sibling"),
    c("asd_da_proband", "sibling"),
    c("asd_d_proband",  "asd_da_proband")
)

df_oe_box_data <- df_oe_all %>%
    left_join(df_all_modules %>% select(module, tier), by = "module") %>%
    filter(cohort %in% cohort_order, !is.na(oe), !is.na(tier)) %>%
    mutate(
        cohort = factor(cohort, levels = cohort_order),
        tier   = factor(tier, levels = levels(df_all_modules$tier))
    )

# --- Wilcoxon tests: compare cohorts within each tier (internal cohort names) --- #
stat_box <- df_oe_box_data %>%
    group_by(tier) %>%
    group_modify(~ {
        oe_split <- split(.x$oe, .x$cohort)
        map_dfr(cohort_pairs_box, function(pair) {
            if (all(pair %in% names(oe_split)))
                tibble(group1 = pair[1], group2 = pair[2],
                       p = wilcox.test(oe_split[[pair[1]]], oe_split[[pair[2]]])$p.value)
        })
    }) %>%
    mutate(
        p.adj    = p.adjust(p, method = "BH"),
        p.signif = case_when(
            p.adj < 0.0001 ~ "****", p.adj < 0.001 ~ "***",
            p.adj < 0.01   ~ "**",   p.adj < 0.05  ~ "*", TRUE ~ "ns"
        )
    ) %>%
    ungroup() %>%
    filter(p.signif != "ns")

# --- Bracket x positions for the dodged cohort layout --- #
# offset for cohort i = dodge_w * (i − (n+1)/2) / n; brackets stacked per tier (narrowest lowest)
cohort_offsets <- setNames(
    dodge_w * (seq_along(cohort_order) - (length(cohort_order) + 1) / 2) / length(cohort_order),
    cohort_order
)
tier_pos <- setNames(seq_along(levels(df_oe_box_data$tier)), levels(df_oe_box_data$tier))

stat_box <- stat_box %>%
    mutate(xmin = tier_pos[as.character(tier)] + cohort_offsets[group1],
           xmax = tier_pos[as.character(tier)] + cohort_offsets[group2]) %>%
    group_by(tier) %>%
    arrange(abs(xmax - xmin)) %>%
    mutate(y.position = 20 * 1.6^(row_number() - 1)) %>%
    ungroup()

y_max_plot <- if (nrow(stat_box) > 0) max(stat_box$y.position) * 2 else 50

p_oe_box <- df_oe_box_data %>%
    ggplot(aes(x = tier, y = oe, fill = cohort, color = cohort)) +
    ggbeeswarm::geom_quasirandom(
        alpha = 0.15, size = 0.4, dodge.width = dodge_w, show.legend = FALSE
    ) +
    geom_boxplot(
        alpha = 0.25, outlier.shape = NA, width = 0.6,
        linewidth = 0.15, position = position_dodge(dodge_w)
    ) +
    geom_boxplot(
        color = "black", outlier.shape = NA, width = 0.6,
        linewidth = 0.35, position = position_dodge(dodge_w)
    ) +
    # Significance brackets: geom_segment + geom_text for reliable log-scale positioning
    geom_segment(data = stat_box, inherit.aes = FALSE,
                 aes(x = xmin, xend = xmax, y = y.position, yend = y.position),
                 color = "grey20", linewidth = 0.3) +
    geom_segment(data = stat_box, inherit.aes = FALSE,
                 aes(x = xmin, xend = xmin, y = y.position / 1.25, yend = y.position),
                 color = "grey20", linewidth = 0.3) +
    geom_segment(data = stat_box, inherit.aes = FALSE,
                 aes(x = xmax, xend = xmax, y = y.position / 1.25, yend = y.position),
                 color = "grey20", linewidth = 0.3) +
    geom_text(data = stat_box, inherit.aes = FALSE,
              aes(x = (xmin + xmax) / 2, y = y.position, label = p.signif),
              vjust = 0.2, size = 2, color = "grey20", angle = 90) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.4) +
    scale_fill_manual(values = oe_cohort_colors,
                      labels = c("ASD-D probands", "ASD-DA probands", "Siblings"), name = NULL) +
    scale_color_manual(values = oe_cohort_colors, name = NULL, guide = "none") +
    scale_x_discrete(labels = c(
        "Stepwise significant"    = "Stepwise\nsignificant",
        "Marginal (ASD-enriched)" = "Marginal\n(ASD-enriched)",
        "Marginal (ASD-depleted)" = "Marginal\n(ASD-depleted)",
        "Not associated"          = "Not\nassociated"
    )) +
    scale_y_log10() +
    # flip so tiers run down the y-axis and O/E along the x-axis
    coord_flip(ylim = c(0.1, y_max_plot)) +
    labs(x = NULL, y = "O/E ratio (log scale)",
         title = "O/E by module tier and cohort"
         ) +
    theme(plot.title  = element_text(size = 7, face = "bold"),
          axis.title  = element_text(size = 6),
          axis.text.x = element_text(size = 5.5),   # O/E axis (after flip)
          axis.text.y = element_text(size = 6.5),   # tier labels (after flip)
          legend.position = "none")                 # legend is shown in Panel A

p_oe_box
```

## 3.5 Combined panel

```{r oe_ddid_combined, fig.height = 7, fig.width = 7.09}
row_top <- p_oe_forest + wrap_elements(full = p_rr_scatter) +
    plot_layout(widths = c(1, 2))

row_bottom <- p_scatter_all + (p_oe_box + coord_flip()) +
    plot_layout(widths = c(2, 1))

p_module_ddid_combined <- row_top / row_bottom +
    plot_layout(heights = c(1.25, 2)) +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 10))

p_module_ddid_combined

ggsave(
    filename = file.path(figures_dir, "supplement", "06_00_module-DDID-OE-enrichment.pdf"),
    plot  = p_module_ddid_combined,
    height    = 7,
    width     = 7.09,
    dpi       = 300
)
```


# ------------------------------------------------------------------------------


## DEPR

## 1 Proband vs Sibling enrichment

```{r plot_proband_sib, fig.height = 5, fig.width = 4}
sig_thresh <- 0.05

p_prob_sib <- df_pairwise %>%
    filter(comparison == "proband vs sibling") %>%
    mutate(
        #module   = factor(module, levels = rev(mod_order)),
        sig      = p_fdr < sig_thresh,
        sig_lab  = case_when(
            p_fdr < 0.001 ~ "***",
            p_fdr < 0.01  ~ "**",
            p_fdr < sig_thresh ~ "*",
            TRUE          ~ ""
        )
    ) %>%
    ggplot(aes(x = rate_ratio, y = reorder(module, rate_ratio))) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h, color = sig),
        size = 0.3, linewidth = 0.4
    ) +
    geom_text(
        aes(x = ci_h, label = sig_lab),
        hjust = -0.3, size = 2.5, color = "black"
    ) +
    scale_color_manual(values = c("TRUE" = "#2E7D32", "FALSE" = "grey60"), guide = NULL) +
    labs(
        x     = "Relative risk (proband / sibling)",
        y     = NULL,
        title = "Proband vs Sibling"
    )

p_prob_sib

f_save_plot(
    plot_obj  = p_prob_sib,
    plot_name = "05_proband_vs_sibling_RR.pdf",
    overwrite = TRUE,
    height    = 5,
    width     = 4,
    dpi       = 300
)
```

## 2 ASD-D vs ASD-DA enrichment

```{r plot_asd_d_asd_da, fig.height = 5, fig.width = 4}
p_asd_d <- df_pairwise %>%
    filter(comparison == "asd_d_proband vs asd_da_proband") %>%
    mutate(
        #module  = factor(module, levels = rev(mod_order)),
        sig     = p_fdr < sig_thresh,
        sig_lab = case_when(
            p_fdr < 0.001 ~ "***",
            p_fdr < 0.01  ~ "**",
            p_fdr < sig_thresh ~ "*",
            TRUE          ~ ""
        )
    ) %>%
    ggplot(aes(x = rate_ratio, y = reorder(module, rate_ratio))) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h, color = sig),
        size = 0.3, linewidth = 0.4
    ) +
    geom_text(
        aes(x = ci_h, label = sig_lab),
        hjust = -0.3, size = 2.5, color = "black"
    ) +
    scale_color_manual(values = c("TRUE" = "#C0392B", "FALSE" = "grey60"), guide = NULL) +
    labs(
        x     = "Relative risk (ASD-D / ASD-DA)",
        y     = NULL,
        title = "ASD-D vs ASD-DA probands"
    )

p_asd_d

f_save_plot(
    plot_obj  = p_asd_d,
    plot_name = "05_ASD-D_vs_ASD-DA_RR.pdf",
    overwrite = TRUE,
    height    = 5,
    width     = 4,
    dpi       = 300
)
```

## 3 ASD-D vs Siblings and ASD-DA vs Siblings

```{r plot_vs_sib, fig.height = 5.5, fig.width = 5}
# Order modules by ASD-D vs sibling RR (descending)
mod_order_sib <- df_pairwise %>%
    filter(comparison == "asd_d_proband vs sibling") %>%
    arrange(rate_ratio) %>%
    pull(module)

p_vs_sib <- df_pairwise %>%
    filter(comparison %in% c("asd_d_proband vs sibling", "asd_da_proband vs sibling")) %>%
    mutate(
        group  = if_else(comparison == "asd_d_proband vs sibling", "ASD-D", "ASD-DA"),
        module = factor(module, levels = mod_order_sib),
        color_group = case_when(
            p_fdr < sig_thresh & group == "ASD-D"   ~ "ASD-D",
            p_fdr < sig_thresh & group == "ASD-DA" ~ "ASD-DA",
            TRUE                                    ~ "ns"
        )
    ) %>%
    ggplot(aes(x = rate_ratio, y = module, color = color_group)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h),
        position = position_dodge(width = 0.6),
        size = 0.25, linewidth = 0.4
    ) +
    scale_color_manual(
        values = c("ASD-D" = "#C0392B", "ASD-DA" = "#4472C4", "ns" = "grey70"),
        labels = c("ASD-D" = "ASD-D (FDR < 0.05)", "ASD-DA" = "ASD-DA (FDR < 0.05)", "ns" = "ns"),
        name   = NULL
    ) +
    coord_cartesian(xlim = c(NA, 30)) +
    labs(
        x     = "Relative risk vs siblings",
        y     = NULL,
        title = "Proband enrichment vs siblings"
    ) +
    theme(legend.position = "top")

p_vs_sib

f_save_plot(
    plot_obj  = p_vs_sib,
    plot_name = "05_ASD-D_ASD-DA_vs_sibling_RR.pdf",
    overwrite = TRUE,
    height    = 5.5,
    width     = 5,
    dpi       = 300
)
```

## 4 Combined figure

```{r plot_combined, fig.height = 7, fig.width = 12}
p_combined <- (p_oe | (p_prob_sib / p_asd_d) | free(p_scatter_sib) ) +
    plot_layout(widths = c(1, 1, 1.5)) +
    plot_annotation(tag_levels = "a")

p_combined

f_save_plot(
    plot_obj  = p_combined,
    plot_name = "05_module_OE_combined.png",
    overwrite = TRUE,
    height    = 7,
    width     = 12,
    dpi       = 300
)
```

**Figure legend.** Observed/expected (O/E) de novo mutation burden across the 28 co-expression modules and gene regulatory network regulons independently associated with ASD rare-variant risk (identified in notebook 01b). De novo variants included: protein-truncating variants (PTVs), damaging missense variants (Mis2, MPC > 2), and moderate-impact missense variants (Mis1, 1 < MPC ≤ 2). Expected mutation counts were derived from published per-gene mutation rate models (Samocha et al.) scaled by cohort size: *E* = Σ(μ~lof~ + μ~mis2~ + μ~mis1~) × 2*N*, where *N* is the number of diploid individuals in each cohort. Cohorts: all probands (*N* = 38,680), unaffected siblings (*N* = 9,567), probands with ASD and developmental delay or intellectual disability (ASD-D; *N* = 13,841, comprising 12,647 ASC ASD-D probands and 1,194 Kaplanis DDD probands), and probands with ASD without ASD-D (ASD-DA; *N* = 24,839).

**(a)** O/E ratio for each module and cohort. Points show the ratio of observed to expected de novo mutation counts; horizontal bars show the 95% Garwood exact Poisson confidence interval (Garwood 1936). Modules are ordered by descending O/E in all probands. A dashed vertical line marks O/E = 1 (no enrichment). **(b)** Relative risk versus unaffected siblings for ASD-D probands (red) and ASD-DA probands (blue), shown side by side for each module. Modules are ordered by descending ASD-D vs sibling RR. Both groups show mutation burden elevated above siblings, with ASD-D probands showing a consistently stronger signal. **(c)** Pairwise relative risk comparing all probands to unaffected siblings for each module. **(d)** Pairwise relative risk comparing ASD-D to ASD-DA probands. For panels (b–d), the relative risk (RR) is computed as the ratio of O/E ratios between the two groups: RR = (obs₁/exp₁) / (obs₂/exp₂). Statistical significance was assessed with a two-sided exact binomial test using the expected mutation rate as the null probability, and p-values were corrected for multiple comparisons across all cohort contrasts using the Benjamini-Hochberg false discovery rate (FDR) procedure. Asterisks denote FDR thresholds: \*p < 0.05, \*\*p < 0.01, \*\*\*p < 0.001.



Plots of the all-module O/E result (`df_oe_all` / `df_pairwise_all`, computed in
Section 1), with modules annotated by association tier: stepwise-significant, marginally
associated (FDR < 0.01, β > 0), marginally depleted, or not associated with ASD risk.

## All modules plots

### 3.5.2 O/E distribution across all modules

```{r all_modules_oe_dist, fig.height = 3, fig.width = 7}
col_tier <- c(
    "Stepwise significant"       = "#922B21",
    "Marginal (ASD-enriched)"   = "#CA6F1E",
    "Marginal (ASD-depleted)"   = "steelblue",
    "Not associated"     = "gray60"
)

p_oe_dist <- df_oe_all %>%
    left_join(df_all_modules %>% select(module, tier), by = "module") %>%
    filter(cohort %in% c("asd_d_proband", "asd_da_proband", "proband", "sibling")) %>%
    mutate(
        cohort = factor(cohort,
                        levels = c("asd_d_proband", "asd_da_proband", "proband", "sibling"),
                        labels = cohort_labels[c("asd_d_proband", "asd_da_proband", "proband", "sibling")])
    ) %>%
    
    ggplot(aes(x = oe, fill = tier, color = tier)) +
    geom_density(alpha = 0.3, linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.4) +
    scale_fill_manual(values  = col_tier, name = NULL) +
    scale_color_manual(values = col_tier, name = NULL) +
    scale_x_log10() +
    facet_wrap(vars(cohort), nrow = 1) +
    labs(
        x     = "O/E ratio (log scale)",
        y     = "Density",
        title = "O/E distribution by module tier and cohort"
    ) +
    theme(legend.position = "top",
          legend.key.size = unit(0.5, "cm"),
          strip.text = element_text(face = "bold"),
          strip.background = element_blank()
    )

p_oe_dist

f_save_plot(
    plot_obj  = p_oe_dist,
    plot_name = "05_all-module_OE_distribution.png",
    overwrite = TRUE,
    height    = 3,
    width     = 7,
    dpi       = 300
)
```

# --- 4) Plots for manuscript --- #

## 4.0 Prep

```{r prep}
## Load data

# Significant modules
df_pairwise <- read_xlsx(file.path(analysis_dir, "06_gene_set_DDID_stratification", "06_00_module_cohort_pairwise.xlsx"))
df_oe <- read_xlsx(file.path(analysis_dir, "06_gene_set_DDID_stratification", "06_00_module_OE_ratios.xlsx"))

# All modules
df_oe_all <- readRDS(file.path(analysis_dir, "06_gene_set_DDID_stratification", "06_00_all-modules-OE.RDS"))
df_pairwise_all <- readRDS(file.path(analysis_dir, "06_gene_set_DDID_stratification", "06_00_all-modules-pairwise.RDS"))

# Set colors and labels
cohort_labels <- c(
    #proband       = "All probands",
    sibling       = "Siblings",
    asd_d_proband  = "ASD-D probands",
    asd_da_proband = "ASD-DA probands"
)

col_cohort <- c(
    #proband        = "#2E7D32",
    sibling        = "gray40",
    asd_d_proband   = "#C0392B",
    asd_da_proband = "#4472C4"
)

# Hub gene map: highest-kME gene per co-expression module (across all modules)
hub_gene_map_all <- df_module_kme %>%
    filter(module == kME_module) %>%
    group_by(module) %>%
    slice_max(kME, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module, hub_gene = gene_name)

# Display labels for the 28 stepwise-significant modules
df_module_labels <- df_oe_all %>%
    dplyr::select(module) %>%
    distinct() %>%
    left_join(hub_gene_map_all, by = "module") %>%
    mutate(
        is_regulon   = is.na(hub_gene),
        regulon_gene = str_extract(module, "^[^_\\-\\+]+"),
        regulon_sign = str_extract(str_remove(module, "^[^_\\-\\+]+"), "[\\+\\-]"),
        module_label = if_else(
            is_regulon,
            paste0(regulon_gene, " (", regulon_sign, ")"),
            hub_gene
        ),
        module_label = if_else(module_label == "43894", "MARCHF4", module_label)
    )  %>% 
    
    # manually change the weird number gene
    mutate(
        module_label = ifelse(module_label == "43894", "MARCHF4", module_label)
    )
```



# --- 5) Tables for manuscript --- #

```{r all_modules_export}
write_xlsx(
    list(
        "A. All module OE ratios" = df_oe_all,
        "B. All module pairwise RRs" = df_pairwise_all %>% filter(!str_detect(module, "M0_grey")) %>% arrange(p_fdr)
    ),
    path = file.path(analysis_dir, "06_gene_set_DDID_stratification", "06_00_all-module-OE-ratios.xlsx")
)
```

