---
title: "Module-level O/E enrichment analyses"
description: "Observed/expected de novo mutation burden for SynGO and GO pathways, across proband and sibling cohorts (ASD-D, ASD-DA, all probands, siblings)."
date: 2026-07-16
format: html
---

# --- 0) Setup --- 

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- paste0(here::here(), "/")
source(paste0(project_dir, "setup.R"))

# Set output directory for notebook analysis
notebook_out_dir <- file.path(analysis_dir, "07_ddid_stratification")
```

```{r load_data}
# Background mutation rates + full proband/sibling counts
df_asd_dnm <- read_tsv(
    file.path(data_dir, "full_results_wcounts_2025-10-08.txt"),
    show_col_types = FALSE
) %>%
    clean_names() %>%
    mutate(mu_total = mu_lof + mu_mis2 + mu_mis1)

# ASD-D / ASD-DA proband counts (ASC)
df_counts_asd_d   <- read_xlsx(file.path(data_dir, "DDID", "counts_asd_ddid.xlsx"))   %>% clean_names()
df_counts_asd_da <- read_xlsx(file.path(data_dir, "DDID", "counts_asd_noddid.xlsx")) %>% clean_names()

# Kaplanis DDD counts (all ASD-D; sex-stratified Male/Female columns)
# N = 1,194 (adds to ASC ASD-D N = 12,647 → total ASD-D N = 13,841)
df_counts_kaplanis <- read_tsv(
    file.path(data_dir, "DDID", "Kaplanis_DDD_ASD_counts_by_gene_new_mis_cats_2025-04-09.txt"),
    show_col_types = FALSE
) %>%
    clean_names()

# SynGO data
syngo_dir      <- file.path(data_dir, "SynGO")
df_syngo_ontol <- read_xlsx(file.path(syngo_dir, "syngo_ontologies.xlsx"))
```


# --- 1) O/E enrichment across all SynGO pathways --- #

## 1.1 Format data

```{r format_counts_syngo}
# Proband and sibling counts from the full file; ASD-D/ASD-DA from separate files
df_counts <- df_asd_dnm %>%
    select(gene, mu_total,
           ptv_proband, mis2_proband, mis1_proband,
           ptv_sibling, mis2_sibling, mis1_sibling) %>%
    mutate(
        count_proband = ptv_proband + mis2_proband + mis1_proband,
        count_sibling = ptv_sibling + mis2_sibling + mis1_sibling
    ) %>%
    left_join(
        df_counts_asd_d %>%
            mutate(count_asd_d_proband = ptv_proband + mis2_proband + mis1_proband) %>%
            select(gene, count_asd_d_proband) %>%
            left_join(
                df_counts_kaplanis %>%
                    mutate(count_kaplanis = (ptv_proband_male + ptv_proband_female) +
                                            (mis2_proband_male + mis2_proband_female) +
                                            (mis1_proband_male + mis1_proband_female)) %>%
                    select(gene, count_kaplanis),
                by = "gene"
            ) %>%
            mutate(count_asd_d_proband = count_asd_d_proband + replace_na(count_kaplanis, 0)) %>%
            select(gene, count_asd_d_proband),
        by = "gene"
    ) %>%
    left_join(
        df_counts_asd_da %>%
            mutate(count_asd_da_proband = ptv_proband + mis2_proband + mis1_proband) %>%
            select(gene, count_asd_da_proband),
        by = "gene"
    )
```

Build one gene set per SynGO pathway using the annotations file (one row per
gene–pathway pair). Only pathways with ≥ 5 genes matched in the count data are kept.

```{r build_syngo_gene_sets}
# IDs that appear as someone's parent_id have children → not leaf nodes
has_children <- na.omit(df_syngo_ontol$parent_id)

# One row per pathway with a list of its genes. hgnc_symbol is a comma-space-delimited
# string of genes per pathway, so split it to one gene per row before grouping.
df_syngo_gene_sets <- df_syngo_ontol %>%
    select(id, name, domain, hgnc_symbol) %>%
    filter(!is.na(hgnc_symbol)) %>%
    separate_rows(hgnc_symbol, sep = ", ") %>%
    mutate(hgnc_symbol = str_trim(hgnc_symbol)) %>%
    group_by(id, name, domain) %>%
    summarise(genes = list(unique(hgnc_symbol)), .groups = "drop") %>%
    mutate(
        n_genes       = map_int(genes, length),
        matched_genes = map(genes, ~ .x[.x %in% df_counts$gene]),   # genes present in the DNM counts
        n_matched     = map_int(matched_genes, length)
    ) %>%
    filter(n_matched >= 5) %>%
    left_join(
        df_syngo_ontol %>% select(id, name_hierarchical, shortname, parent_id),
        by = "id"
    ) %>%
    
    # identify if pathway is root, intermediate, or leaf
    mutate(
        category = case_when(
            is.na(parent_id)        ~ "root",
            id %in% has_children ~ "intermediate",
            TRUE                    ~ "leaf"
        )
    )

cat(sprintf(
    "%d SynGO pathways with >= 5 matched genes (%d BP, %d CC)\n",
    nrow(df_syngo_gene_sets),
    sum(df_syngo_gene_sets$domain == "BP"),
    sum(df_syngo_gene_sets$domain == "CC")
))
cat(sprintf(
    "  root: %d  |  intermediate: %d  |  leaf: %d\n",
    sum(df_syngo_gene_sets$category == "root"),
    sum(df_syngo_gene_sets$category == "intermediate"),
    sum(df_syngo_gene_sets$category == "leaf")
))
```


## 1.2 O/E analysis helper functions

```{r helpers_syngo}
# --- Garwood (1936) exact Poisson CI for O/E ratio --- #
# p-value: exact Poisson test of H0: obs ~ Poisson(exp), i.e. O/E = 1
oe_garwood <- function(obs, exp) {
    oe   <- obs / exp
    ci_l <- if (obs > 0) qchisq(0.025, 2 * obs) / (2 * exp) else 0
    ci_h <- qchisq(0.975, 2 * (obs + 1)) / (2 * exp)
    p    <- if (obs > 0 && exp > 0)
                poisson.test(obs, T = exp, alternative = "two.sided")$p.value
            else NA_real_
    list(oe = oe, ci_l = ci_l, ci_h = ci_h, p = p)
}

# --- Pairwise conditional binomial test --- #
pairwise_binom <- function(obs1, exp1, obs2, exp2) {
    obs1 <- as.numeric(obs1)[1]; exp1 <- as.numeric(exp1)[1]
    obs2 <- as.numeric(obs2)[1]; exp2 <- as.numeric(exp2)[1]

    # Completely uninformative: missing inputs or both groups have zero observed
    if (any(is.na(c(obs1, obs2, exp1, exp2))) || (obs1 == 0 && obs2 == 0))
        return(tibble(rel_rr = NA_real_, ci_l = NA_real_, ci_h = NA_real_, p = NA_real_))

    rel <- (obs1 / exp1) / (obs2 / exp2)   # = 0 when obs1 = 0; Inf when obs2 = 0
    p0  <- exp1 / (exp1 + exp2)
    p   <- binom.test(obs1, obs1 + obs2, p0, alternative = "two.sided")$p.value

    # Log-scale SE requires both counts > 0; return rel_rr + p but no CI when one is zero
    if (obs1 == 0 || obs2 == 0)
        return(tibble(rel_rr = rel, ci_l = NA_real_, ci_h = NA_real_, p = p))

    se <- sqrt(1 / obs1 + 1 / obs2)
    tibble(
        rel_rr = rel,
        ci_l   = exp(log(rel) - qnorm(0.975) * se),
        ci_h   = exp(log(rel) + qnorm(0.975) * se),
        p      = p
    )
}

# Cohort sample sizes
N_cohorts <- list(
    asd_da_proband = 24839L,
    asd_d_proband   = 13841L,
    proband        = 38680L,
    sibling        =  9567L
)
```

## 1.3 Calculate O/E per SynGO pathway per cohort

```{r oe_per_pathway_syngo}
cohort_spec <- tribble(
    ~cohort,            ~count_col,             ~n,
    "proband",          "count_proband",         N_cohorts$proband,
    "sibling",          "count_sibling",         N_cohorts$sibling,
    "asd_d_proband",    "count_asd_d_proband",    N_cohorts$asd_d_proband,
    "asd_da_proband",   "count_asd_da_proband",  N_cohorts$asd_da_proband
)

# Run analysis across all cohorts for each pathway
# Load from cache if it's already been run
oe_cache <- file.path(notebook_out_dir, "07_01_SynGO-OE.RDS")

if (file.exists(oe_cache)) {
    df_oe_syngo <- readRDS(oe_cache)
} else {
    
    df_oe_syngo <- map2_dfr(
        .x = cohort_spec$cohort,
        .y = cohort_spec$count_col,
        .f = function(cohort_name, cnt_col) {
            
            n_cohort <- cohort_spec %>% filter(cohort == cohort_name) %>% pull(n)
            
            df_syngo_gene_sets %>%
                select(id, name, domain, matched_genes, n_genes, n_matched, category) %>%
                pmap_dfr(function(id, name, domain, matched_genes, n_genes, n_matched, category) {
                    
                    m   <- df_counts %>% filter(gene %in% matched_genes)
                    obs <- sum(m[[cnt_col]], na.rm = TRUE)
                    exp <- sum(m$mu_total,   na.rm = TRUE) * 2 * n_cohort
                    ci  <- oe_garwood(obs, exp)
                    
                    tibble(
                        id     = id,
                        name   = name,
                        domain = domain,
                        category  = category,
                        cohort    = cohort_name,
                        n_genes   = n_genes,
                        n_matched = n_matched,
                        n_cohort  = n_cohort,
                        obs       = obs,
                        exp       = round(exp, 4),
                        oe        = ci$oe,
                        ci_l      = ci$ci_l,
                        ci_h      = ci$ci_h,
                        p         = ci$p
                    )
                })
        }
    ) %>%
        group_by(cohort) %>%
        mutate(p_fdr = p.adjust(p, method = "BH")) %>%
        ungroup() %>%
        arrange(p)
    
    saveRDS(df_oe_syngo, oe_cache)
    
}
```

## 1.4 Cohort pairwise comparisons within each SynGO term

For each pathway, compare:
- Proband vs Sibling (overall enrichment)
- ASD-D_Proband vs ASD-DA_Proband (ASD-D specificity)
- ASD-D_Proband vs Sibling
- ASD-DA_Proband vs Sibling

```{r cohort_pairs_syngo}
cohort_pairs <- list(
    c("proband",        "sibling"),
    c("asd_d_proband",   "asd_da_proband"),
    c("asd_d_proband",   "sibling"),
    c("asd_da_proband", "sibling")
)

# Run analysis across all pairwise combinations of cohorts
# Load from cache if it's already been run
pairwise_cache <- file.path(notebook_out_dir, "07_01_SynGO-pairwise-RR.RDS")

if (file.exists(pairwise_cache)) {
    df_pairwise_syngo <- readRDS(pairwise_cache)
} else {
    
    df_pairwise_syngo <- df_syngo_gene_sets %>%
        select(id) %>%
        pmap_dfr(function(id) {
            map_dfr(cohort_pairs, function(pair) {
                ga <- pair[1]; gb <- pair[2]
                sa <- df_oe_syngo %>% filter(id == !!id, cohort == ga)
                sb <- df_oe_syngo %>% filter(id == !!id, cohort == gb)
                bind_cols(
                    tibble(id = id, comparison = paste(ga, "vs", gb)),
                    pairwise_binom(sa$obs, sa$exp, sb$obs, sb$exp)
                )
            })
        }) %>%
        mutate(p_fdr = p.adjust(p, method = "BH")) %>%
        left_join(
            df_syngo_gene_sets %>% select(id, name, domain, n_matched, category),
            by = "id"
        ) %>%
        arrange(p)
    
    saveRDS(df_pairwise_syngo, pairwise_cache)
    
}
```

## 1.5 Export tables

```{r export_tables}
df_oe_syngo %>%
    arrange(cohort, p) %>%
    write_csv(file.path(notebook_out_dir, "07_01_SynGO_OE_ratios.csv"))

df_pairwise_syngo %>%
    arrange(comparison, p) %>%
    write_csv(file.path(notebook_out_dir, "07_01_SynGO_cohort_pairwise_RR.csv"))
```


# --- 2) GO analysis --- #

## 2.1 Format data

Build one gene set per GO pathway using `org.Hs.eg.db` (all evidence codes) and
`GO.db` for term metadata and hierarchy. Pathways with fewer than 5 or more
than 1,000 matched genes in the count data are excluded.

```{r build_go_gene_sets}
# --- Gene symbol --> GO term annotations --- #

go_annot_raw <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = keys(org.Hs.eg.db, keytype = "SYMBOL"),
    columns = c("GO", "ONTOLOGY"),
    keytype = "SYMBOL"
) %>%
    as_tibble() %>%
    clean_names() %>%
    filter(!is.na(go), !is.na(ontology)) %>%
    select(gene = symbol, id = go, domain = ontology) %>%
    distinct()

# --- GO term names --- #
go_terms <- AnnotationDbi::select(
    GO.db,
    keys    = unique(go_annot_raw$id),
    columns = "TERM",
    keytype = "GOID"
) %>%
    as_tibble() %>%
    clean_names() %>%
    rename(id = goid, name = term)

# --- Hierarchy: identify root / intermediate / leaf nodes --- #
# A term has children if it appears as a key with non-NULL entries in *CHILDREN
go_children_list <- c(
    as.list(GOBPCHILDREN),
    as.list(GOCCCHILDREN),
    as.list(GOMFCHILDREN)
)
has_children_ids <- names(go_children_list)[
    !sapply(go_children_list, function(x) is.null(x) || all(is.na(x)))
]

# A term is a root if its only parent is the virtual "all" node
go_parents_list <- c(
    as.list(GOBPPARENTS),
    as.list(GOCCPARENTS),
    as.list(GOMFPARENTS)
)
root_ids <- names(go_parents_list)[
    sapply(go_parents_list, function(x) is.null(x) || all(x == "all"))
]

# --- Build gene sets per GO term --- #
df_go_gene_sets <- go_annot_raw %>%
    left_join(go_terms, by = "id") %>%
    filter(!is.na(name)) %>%
    group_by(id, name, domain) %>%
    summarise(genes = list(unique(gene)), .groups = "drop") %>%
    mutate(
        n_genes       = map_int(genes, length),
        matched_genes = map(genes, ~ .x[.x %in% df_counts$gene]),
        n_matched     = map_int(matched_genes, length)
    ) %>%
    filter(n_matched >= 5, n_matched <= 1000) %>%
    mutate(
        category = case_when(
            id %in% root_ids         ~ "root",
            id %in% has_children_ids ~ "intermediate",
            TRUE                        ~ "leaf"
        )
    )

cat(sprintf(
    "%d GO pathways with 5–1000 matched genes (BP: %d, CC: %d, MF: %d)\n",
    nrow(df_go_gene_sets),
    sum(df_go_gene_sets$domain == "BP"),
    sum(df_go_gene_sets$domain == "CC"),
    sum(df_go_gene_sets$domain == "MF")
))
cat(sprintf(
    "  root: %d  |  intermediate: %d  |  leaf: %d\n",
    sum(df_go_gene_sets$category == "root"),
    sum(df_go_gene_sets$category == "intermediate"),
    sum(df_go_gene_sets$category == "leaf")
))
```


## 2.2 O/E per GO pathway per cohort

```{r oe_per_pathway_go}
# Run analysis across all cohorts for each pathway
# Load from cache if it's already been run
oe_cache <- file.path(notebook_out_dir, "07_01_all-GO-OE.RDS")

if (file.exists(oe_cache)) {
    df_oe_go <- readRDS(oe_cache)
} else {
    
    df_oe_go <- map2_dfr(
        .x = cohort_spec$cohort,
        .y = cohort_spec$count_col,
        .f = function(cohort_name, cnt_col) {
            
            n_cohort <- cohort_spec %>% filter(cohort == cohort_name) %>% pull(n)
            
            df_go_gene_sets %>%
                select(id, name, domain, matched_genes, n_genes, n_matched, category) %>%
                pmap_dfr(function(id, name, domain, matched_genes, n_genes, n_matched, category) {
                    
                    m   <- df_counts %>% filter(gene %in% matched_genes)
                    obs <- sum(m[[cnt_col]], na.rm = TRUE)
                    exp <- sum(m$mu_total,   na.rm = TRUE) * 2 * n_cohort
                    ci  <- oe_garwood(obs, exp)
                    
                    tibble(
                        id     = id,
                        name   = name,
                        domain = domain,
                        category  = category,
                        cohort    = cohort_name,
                        n_genes   = n_genes,
                        n_matched = n_matched,
                        n_cohort  = n_cohort,
                        obs       = obs,
                        exp       = round(exp, 4),
                        oe        = ci$oe,
                        ci_l      = ci$ci_l,
                        ci_h      = ci$ci_h,
                        p         = ci$p
                    )
                })
        }
    ) %>%
        group_by(cohort) %>%
        mutate(p_fdr = p.adjust(p, method = "BH")) %>%
        ungroup() %>%
        arrange(p)
    
    saveRDS(df_oe_go, oe_cache)
    
}
```

## 2.3 Cohort pairwise comparisons

```{r cohort_pairs_go}
# Run analysis across all pairwise combinations of cohorts
# Load from cache if it's already been run
pairwise_cache <- file.path(notebook_out_dir, "07_01_all-GO-pairwise-RR.RDS")

if (file.exists(pairwise_cache)) {
    df_pairwise_go <- readRDS(pairwise_cache)
} else {
    
    df_pairwise_go <- df_go_gene_sets %>%
        select(id) %>%
        pmap_dfr(function(id) {
            map_dfr(cohort_pairs, function(pair) {
                ga <- pair[1]; gb <- pair[2]
                sa <- df_oe_go %>% filter(id == !!id, cohort == ga)
                sb <- df_oe_go %>% filter(id == !!id, cohort == gb)
                bind_cols(
                    tibble(id = id, comparison = paste(ga, "vs", gb)),
                    pairwise_binom(sa$obs, sa$exp, sb$obs, sb$exp)
                )
            })
        }) %>%
        mutate(p_fdr = p.adjust(p, method = "BH")) %>%
        left_join(
            df_go_gene_sets %>% select(id, name, domain, n_matched, category),
            by = "id"
        ) %>%
        arrange(p)
    
    saveRDS(df_pairwise_go, pairwise_cache)
    
}
```


# --- 3) Plots --- #

```{r plot_globals}
sig_thresh <- 0.05

cohort_labels <- c(
    proband        = "All probands",
    sibling        = "Siblings",
    asd_d_proband   = "ASD-D probands",
    asd_da_proband = "ASD-DA probands"
)

col_sig_group <- c(
    "Both"       = "#885371",
    "ASD-D only"  = "#C0392B",
    "ASD-DA only"= "#4472C4",
    "Neither"    = "grey75"
)
# # SynGO pathway order within each domain: sort by proband O/E descending
# pathway_order_syngo <- df_oe_syngo %>%
#     filter(cohort == "proband") %>%
#     arrange(go_domain, desc(oe)) %>%
#     pull(go_name)
# 
# # GO pathway order within each domain: sort by proband O/E descending
# pathway_order_go <- df_oe_go %>%
#     filter(cohort == "proband") %>%
#     arrange(go_domain, desc(oe)) %>%
#     pull(go_name)
```

## 3.1 SynGO scatter: ASD-DA vs siblings (x) × ASD-D vs siblings (y)

```{r syngo_scatter, fig.height = 2.6, fig.width = 7.1}
# --- Build wide data: one row per pathway --- #
df_scatter_syngo <- df_pairwise_syngo %>%
    filter(comparison %in% c("asd_d_proband vs sibling", "asd_da_proband vs sibling")) %>%
    select(id, name, domain, comparison, rel_rr, ci_l, ci_h, n_matched, p_fdr) %>%
    mutate(group = if_else(comparison == "asd_d_proband vs sibling", "asd_d", "asd_da")) %>%
    select(-comparison) %>%
    pivot_wider(
        id_cols     = c(id, name, domain, n_matched),
        names_from  = group,
        values_from = c(rel_rr, ci_l, ci_h, p_fdr)
    ) %>%
    left_join(
        df_pairwise_syngo %>%
            filter(comparison == "asd_d_proband vs asd_da_proband") %>%
            select(id, p_fdr_asd_d_asd_da = p_fdr, rel_rr_asd_d_asd_da = rel_rr),
        by = "id"
    ) %>%
    mutate(
        sig_asd_d_asd_da = replace_na(p_fdr_asd_d_asd_da < sig_thresh, FALSE),
        sig_asd_d        = replace_na(p_fdr_asd_d   < sig_thresh, FALSE),
        sig_asd_da      = replace_na(p_fdr_asd_da < sig_thresh, FALSE),
        sig_group = case_when(
            sig_asd_d  &  sig_asd_da ~ "Both",
            sig_asd_d  & !sig_asd_da ~ "ASD-D only",
            !sig_asd_d &  sig_asd_da ~ "ASD-DA only",
            TRUE                    ~ "Neither"
        ),
        sig_group = factor(sig_group,
                           levels = c("Both", "ASD-D only", "ASD-DA only", "Neither"))
    )

# Pathways explicitly cited in the manuscript text (to label)
highlight_ids <- c(
    "GO:0099056",  # integral component of presynaptic membrane        (DDID-biased)
    "GO:0099092",  # postsynaptic density intracellular component       (DDID-biased)
    "GO:0050804",  # modulation of chemical synaptic transmission       (DDID-biased)
    "GO:0099560"   # synapse adhesion between pre- and post-synapse     (NoDDID only)
)

# Create df to label specific pathways
df_scatter_labels <- bind_rows(
    # Manuscript-cited pathways
    df_scatter_syngo %>%
        filter(id %in% highlight_ids),
    # Top 2 per domain among pathways sig in BOTH ASD-D and ASD-DA vs siblings
    df_scatter_syngo %>%
        filter(sig_group %in% c("Both", "ASD-D only")) %>%
        group_by(domain, sig_group) %>%
        slice_min(p_fdr_asd_d, n = 2, with_ties = FALSE) %>%
        ungroup()
) %>%
    distinct(id, .keep_all = TRUE) %>%
    mutate(go_label = str_remove(name, "\\s*\\(GO:[0-9]+\\)\\s*$"))

# Plot
p_syngo_scatter <- df_scatter_syngo %>%
    arrange(sig_group != "Neither") %>%
    ggplot(aes(x = rel_rr_asd_da, y = rel_rr_asd_d, color = sig_group)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(size = n_matched, fill = sig_group), alpha = 0.8, shape = 21, color = "white") +
    ggrepel::geom_text_repel(
        data = df_scatter_labels,
        aes(label = str_wrap(go_label, width = 25)),
        size = 1.8, max.overlaps = Inf,
        min.segment.length = 0,
        segment.color = "grey50", segment.size = 0.3,
        box.padding = 1.5,
        force = 6, force_pull = 0.5
    ) +
    scale_color_manual(values = col_sig_group, name = "Sig vs siblings", guide = "none") +
    scale_fill_manual(values = col_sig_group, name = "Sig vs siblings") +
    scale_size_continuous(range = c(1, 5), name = "# genes with \n≥1 mutation") +
    scale_x_log10() +
    scale_y_log10() +
    guides(size = guide_legend(override.aes = aes(shape = 21, color = "black", fill = "grey60"))) +
    facet_wrap(~ domain, ncol = 2) +
    labs(
        x     = "ASD-DA vs siblings (log RR)",
        y     = "ASD-D vs siblings (log RR)",
        title = "SynGO pathway relative enrichments"
    ) +
    theme(
        legend.position  = "right",
        strip.text       = element_text(size = 7, face = "bold"),
        strip.background = element_blank(),
        legend.text      = element_text(size = 5.5),
        legend.title     = element_text(size = 6),
        legend.margin    = margin(l = -5),
        axis.text        = element_text(size = 5.5),
        axis.title       = element_text(size = 6),
        plot.title = element_text(size = 7, hjust = 0.5, face = "bold")
    )


p_syngo_scatter
```

## 3.2 GO scatter: ASD-DA vs siblings (x) × ASD-D vs siblings (y)

```{r go_scatter, fig.width = 7.09, fig.height = 3}
# Data prep
df_scatter_go <- df_pairwise_go %>%
    filter(comparison %in% c("asd_d_proband vs sibling", "asd_da_proband vs sibling")) %>%
    select(id, name, domain, category, comparison,
           rel_rr, ci_l, ci_h, n_matched, p_fdr) %>%
    mutate(group = if_else(comparison == "asd_d_proband vs sibling", "asd_d", "asd_da")) %>%
    select(-comparison) %>%
    pivot_wider(
        id_cols     = c(id, name, domain, category, n_matched),
        names_from  = group,
        values_from = c(rel_rr, ci_l, ci_h, p_fdr)
    ) %>%
    left_join(
        df_pairwise_go %>%
            filter(comparison == "asd_d_proband vs asd_da_proband") %>%
            select(id, p_fdr_asd_d_asd_da = p_fdr, rel_rr_asd_d_asd_da = rel_rr),
        by = "id"
    ) %>%
    mutate(
        sig_asd_d_asd_da = replace_na(p_fdr_asd_d_asd_da < sig_thresh, FALSE),
        sig_asd_d        = replace_na(p_fdr_asd_d   < sig_thresh, FALSE),
        sig_asd_da      = replace_na(p_fdr_asd_da < sig_thresh, FALSE),
        sig_group = case_when(
            sig_asd_d  &  sig_asd_da ~ "Both",
            sig_asd_d  & !sig_asd_da ~ "ASD-D only",
            !sig_asd_d &  sig_asd_da ~ "ASD-DA only",
            TRUE                    ~ "Neither"
        ),
        sig_group = factor(sig_group,
                           levels = c("Both", "ASD-D only", "ASD-DA only", "Neither"))
    )

# Plot
p_go_scatter <- df_scatter_go %>%
    arrange(sig_group != "Neither") %>%
    ggplot(aes(x = rel_rr_asd_da, y = rel_rr_asd_d, fill = sig_group)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(size = n_matched), shape = 21, color = "white", stroke = 0.2, alpha = 0.7) +
    ggrepel::geom_text_repel(
        data = . %>% filter(sig_group == "ASD-DA only" & rel_rr_asd_da > 1),
        aes(label = str_wrap(name, width = 25), color = sig_group),
        size = 1.8, max.overlaps = 30,
        nudge_x = 0.3, nudge_y = -0.3,
        direction = "both",
        min.segment.length = 0,
        segment.color = "grey50", segment.size = 0.3,
        box.padding = 0.4
    ) +
    scale_fill_manual(values = col_sig_group, name = "Sig vs siblings") +
    scale_color_manual(values = col_sig_group, guide = "none") +
    scale_size_continuous(range = c(0.5, 4), name = "# pathway genes",
                          guide = guide_legend(
                              override.aes = list(fill = "grey50", color = "white")
                          )) +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ domain, nrow = 1) +
    labs(
        x     = "ASD-DA vs siblings (log RR)",
        y     = "ASD-D vs siblings (log RR)",
        title = "GO pathway relative enrichments"
    ) +
    theme(
        plot.title = element_text(size = 7, hjust = 0.5, face = "bold"),
        legend.position  = "none",
        legend.box = "horizontal",
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_go_scatter
```


## 3.3 Combine panels

```{r SynGO_GO_OE, fig.height = 6, fig.width = 7.09}
p_syngo_ddid_combined <- p_syngo_scatter / free(p_go_scatter) +
    plot_layout(heights = c(1, 1.25)) +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 10))

p_syngo_ddid_combined

ggsave(
    filename = file.path(figures_dir, "supplement", "07_01_SynGO-DDID-OE-enrichment.pdf"),
    plot  = p_syngo_ddid_combined,
    height    = 6,
    width     = 7.09,
    dpi       = 300
)
```



# ------------------------------------------------------------------------------

### 3.2.1 SynGO O/E forest plot — DDID and NoDDID cohorts

Significant pathways only (FDR < 0.05 in at least one of DDID vs siblings or NoDDID vs
siblings). Ordered by DDID O/E within each domain; colour encodes significance group
from the scatter plot above.

```{r plot_syngo_oe_forest, fig.height = 6.69, fig.width = 7.09}
# Significant pathways in either DDID or NoDDID vs siblings
sig_go_ids_syngo <- df_pairwise_syngo %>%
    filter(
        comparison %in% c("ddid_proband vs sibling", "noddid_proband vs sibling"),
        p_fdr < sig_thresh
    ) %>%
    pull(go_id) %>%
    unique()

# Pathway order: DDID O/E descending within each domain
forest_order_syngo <- df_oe_syngo %>%
    filter(cohort == "ddid_proband", go_id %in% sig_go_ids_syngo) %>%
    arrange(go_domain, desc(oe)) %>%
    mutate(go_name_clean = str_remove(go_name, "\\s*\\(GO:[0-9]+\\)\\s*$")) %>%
    pull(go_name_clean)

# Join significance group; derive per-point colour
df_oe_forest_syngo <- df_oe_syngo %>%
    filter(
        cohort %in% c("ddid_proband", "noddid_proband"),
        go_id  %in% sig_go_ids_syngo
    ) %>%
    left_join(
        df_scatter_syngo %>% select(go_id, sig_group),
        by = "go_id"
    ) %>%
    mutate(
        cohort_label = if_else(cohort == "ddid_proband", "DDID", "NoDDID"),
        point_sig = case_when(
            cohort_label == "DDID"   & sig_group %in% c("Both", "DDID only")   ~ "DDID",
            cohort_label == "NoDDID" & sig_group %in% c("Both", "NoDDID only") ~ "NoDDID",
            TRUE                                                                 ~ "ns"
        ) %>% factor(levels = c("DDID", "NoDDID", "ns")),
        go_name_clean = str_remove(go_name, "\\s*\\(GO:[0-9]+\\)\\s*$") %>% str_wrap(40),
        go_name_clean = factor(go_name_clean, levels = rev(forest_order_syngo %>% str_wrap(40))),
        sig_group     = factor(sig_group, levels = c("Both", "DDID only", "NoDDID only", "Neither"))
    )

col_point_sig <- c("DDID" = "#C0392B", "NoDDID" = "#4472C4", "ns" = "grey70")
lab_point_sig <- c("DDID" = "DDID (FDR < 0.05)", "NoDDID" = "NoDDID (FDR < 0.05)", "ns" = "ns")

# Extract shared legend from a dummy plot so it sits above all sub-plots
legend_dummy <- ggplot(
    tibble(x = 1, y = 1, point_sig = factor(names(col_point_sig), levels = names(col_point_sig))),
    aes(x, y, color = point_sig)
) +
    geom_point() +
    scale_color_manual(values = col_point_sig, labels = lab_point_sig,
                       name = NULL, drop = FALSE) +
    theme(legend.position = "top",
          legend.key.size = unit(0.35, "cm"),
          legend.text     = element_text(size = 6))

shared_legend <- cowplot::get_legend(legend_dummy)

# Helper: one sub-plot per sig_group row (no internal legend)
make_forest_row <- function(df, grp, show_title = FALSE) {
    ggplot(df, aes(x = oe, y = go_name_clean, color = point_sig)) +
        geom_vline(xintercept = 1, linetype = "dashed",
                   color = "grey50", linewidth = 0.4) +
        geom_pointrange(
            aes(xmin = ci_l, xmax = ci_h),
            position = position_dodge(width = 0.6),
            size = 0.25, linewidth = 0.4
        ) +
        scale_color_manual(values = col_point_sig, labels = lab_point_sig,
                           name = NULL, drop = FALSE) +
        facet_grid2(. ~ go_domain, independent = "all", scales = "free") +
        labs(
            x     = if (show_title) "O/E ratio" else NULL,
            y     = NULL,
            title = grp
        ) +
        theme(
            legend.position  = "none",
            axis.text.y      = element_text(size = 6),
            axis.text.x      = element_text(size = 7),
            axis.title.x     = element_text(size = 7),
            plot.title       = element_text(size = 7, face = "bold", hjust = 0.5),
            plot.margin      = margin(t = 1, r = 5, b = 1, l = 5),
            strip.text       = element_text(face = "bold", size = 7),
            strip.background = element_blank(),
            panel.spacing    = unit(0.1, "lines")
        )
}

sig_group_levels <- c("Both", "DDID only", "NoDDID only")

n_paths_syngo <- df_oe_forest_syngo %>%
    filter(sig_group %in% sig_group_levels) %>%
    group_by(sig_group) %>%
    summarise(n = n_distinct(go_name_clean), .groups = "drop") %>%
    mutate(sig_group = factor(sig_group, levels = sig_group_levels)) %>%
    arrange(sig_group)

forest_rows_syngo <- map(sig_group_levels, function(grp) {
    make_forest_row(
        df         = df_oe_forest_syngo %>% filter(sig_group == grp),
        grp        = grp,
        show_title = grp == tail(sig_group_levels, 1)
    )
})

p_syngo_oe_forest <- patchwork::wrap_plots(
    c(list(patchwork::wrap_elements(shared_legend)), forest_rows_syngo),
    ncol    = 1,
    heights = c(0.8, n_paths_syngo$n)
)

p_syngo_oe_forest

f_save_plot(
    plot_obj  = p_syngo_oe_forest,
    plot_name = "04a_syngo_OE_forest_sig.png",
    overwrite = TRUE,
    height    = 6.69,
    width     = 7.09,
    dpi       = 300
)
```

### 3.2.2 SynGO DDID vs NoDDID O/E scatter

Each point is a SynGO pathway; x = NoDDID O/E, y = DDID O/E. Colour and labels
follow the same significance grouping as above.

```{r plot_syngo_scatter_oe, fig.height = 4, fig.width = 9}
df_scatter_oe_syngo <- df_oe_syngo %>%
    filter(cohort %in% c("ddid_proband", "noddid_proband")) %>%
    select(go_id, go_name, go_domain, n_matched, cohort, oe, ci_l, ci_h) %>%
    mutate(cohort = if_else(cohort == "ddid_proband", "ddid", "noddid")) %>%
    pivot_wider(
        id_cols     = c(go_id, go_name, go_domain, n_matched),
        names_from  = cohort,
        values_from = c(oe, ci_l, ci_h)
    ) %>%
    left_join(
        df_scatter_syngo %>% select(go_id, sig_group),
        by = "go_id"
    )

p_syngo_scatter_oe <- df_scatter_oe_syngo %>%
    arrange(sig_group != "Neither") %>%
    ggplot(aes(x = oe_asd_da, y = oe_asd_d, color = sig_group)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(size = n_matched), alpha = 0.8) +
    ggrepel::geom_text_repel(
        data = . %>% filter(sig_group == "NoDDID only"),
        aes(label = str_wrap(str_remove(go_name, "\\s*\\(GO:[0-9]+\\)\\s*$"), width = 30)),
        size = 2.2, max.overlaps = 25,
        nudge_x = 0.3, nudge_y = -0.3,
        direction = "both",
        min.segment.length = 0,
        segment.color = "grey50", segment.size = 0.3,
        box.padding = 0.4
    ) +
    scale_color_manual(values = col_sig_group, name = "Sig vs siblings") +
    scale_size_continuous(range = c(1, 5), name = "# pathway genes") +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ go_domain, ncol = 2) +
    labs(
        x     = "NoDDID O/E (log scale)",
        y     = "DDID O/E (log scale)",
        title = "SynGO DDID vs NoDDID O/E"
    ) +
    theme(
        legend.position  = "right",
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_syngo_scatter_oe

f_save_plot(
    plot_obj  = p_syngo_scatter_oe,
    plot_name = "04c_syngo_asd_d_asd_da_scatter_OE.png",
    overwrite = TRUE,
    height    = 4,
    width     = 9,
    dpi       = 300
)
```


# --- 4) Table export --- #

## 4.1 SynGO interim tables

```{r export_syngo_tables}
df_oe_syngo %>%
    write_xlsx(file.path(notebook_out_dir, "07_02_syngo_OE_ratios.xlsx"))

df_pairwise_syngo %>%
    write_xlsx(file.path(notebook_out_dir, "07_02_syngo_cohort_pairwise.xlsx"))
```

## 4.2 GO interim tables

```{r export_go_tables}
df_oe_go %>%
    write_xlsx(file.path(notebook_out_dir, "07_02_go_OE_ratios.xlsx"))

df_pairwise_go %>%
    write_xlsx(file.path(notebook_out_dir, "07_02_go_cohort_pairwise.xlsx"))
```

## 4.3 Manuscript tables

```{r export_ms_tables}
write_xlsx(
    list(
        "A. SynGO pathway OE ratios" = df_oe_syngo,
        "B. SynGO pathway pairwise RRs" = df_pairwise_syngo
    ),
    path = file.path(notebook_out_dir, "07_02_all-SynGO-OE-ratios.xlsx")
)

write_xlsx(
    list(
        "A. GO pathway OE ratios (p < 0.05)" = df_oe_go %>% filter(p < 0.05),
        "B. GO pathway pairwise RRs (p < 0.05)" = df_pairwise_go %>% arrange(p_fdr) %>% filter(p < 0.05)
    ),
    path = file.path(notebook_out_dir, "07_02_all-GO-OE-ratios.xlsx")
)
```

# 3.3 DEPR

## 3.3.1 SynGO O/E ratio per pathway per cohort

```{r plot_syngo_oe, fig.height = 22, fig.width = 7}
p_syngo_oe <- df_oe_syngo %>%
    mutate(
        go_name = factor(go_name, levels = rev(pathway_order_syngo)),
        cohort  = factor(
            cohort,
            levels = c("ddid_proband", "noddid_proband", "proband", "sibling"),
            labels = cohort_labels[c("ddid_proband", "noddid_proband", "proband", "sibling")]
        )
    ) %>%
    ggplot(aes(x = oe, y = go_name, color = cohort)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h),
        position = position_dodge(width = 0.7),
        size = 0.2, linewidth = 0.3
    ) +
    scale_color_manual(values = setNames(col_cohort, cohort_labels[names(col_cohort)])) +
    facet_grid(go_domain ~ ., scales = "free_y", space = "free_y") +
    labs(
        x     = "Observed / Expected",
        y     = NULL,
        color = NULL,
        title = "SynGO pathway O/E ratio by cohort"
    ) +
    theme(
        legend.position  = "top",
        axis.text.y      = element_text(size = 5),
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_syngo_oe

f_save_plot(
    plot_obj  = p_syngo_oe,
    plot_name = "05c_syngo_OE_ratio.pdf",
    overwrite = TRUE,
    height    = 22,
    width     = 7,
    dpi       = 300
)
```

## 3.3.2 SynGO DDID vs NoDDID enrichment

```{r plot_syngo_asd_d_asd_da, fig.height = 22, fig.width = 5}
p_syngo_asd_d <- df_pairwise_syngo %>%
    filter(comparison == "ddid_proband vs noddid_proband") %>%
    mutate(
        go_name = factor(go_name, levels = rev(pathway_order_syngo)),
        sig     = p_fdr < sig_thresh,
        sig_lab = case_when(
            p_fdr < 0.001      ~ "***",
            p_fdr < 0.01       ~ "**",
            p_fdr < sig_thresh ~ "*",
            TRUE               ~ ""
        )
    ) %>%
    ggplot(aes(x = rel_rr, y = go_name)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h, color = sig),
        size = 0.2, linewidth = 0.3
    ) +
    geom_text(
        aes(x = ci_h, label = sig_lab),
        hjust = -0.3, size = 2, color = "black"
    ) +
    scale_color_manual(
        values = c("TRUE" = "#C0392B", "FALSE" = "grey60"),
        guide  = NULL
    ) +
    facet_grid(go_domain ~ ., scales = "free_y", space = "free_y") +
    labs(
        x     = "Relative risk (DDID / NoDDID)",
        y     = NULL,
        title = "SynGO DDID vs NoDDID"
    ) +
    theme(
        axis.text.y      = element_text(size = 5),
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_syngo_asd_d

f_save_plot(
    plot_obj  = p_syngo_asd_d,
    plot_name = "05c_syngo_asd_d_vs_asd_da_RR.pdf",
    overwrite = TRUE,
    height    = 22,
    width     = 5,
    dpi       = 300
)
```

## 3.3.3 SynGO DDID and NoDDID vs siblings

```{r plot_syngo_vs_sib, fig.height = 22, fig.width = 6}
pathway_order_sib_syngo <- df_pairwise_syngo %>%
    filter(comparison == "ddid_proband vs sibling") %>%
    arrange(go_domain, rel_rr) %>%
    pull(go_name)

p_syngo_vs_sib <- df_pairwise_syngo %>%
    filter(comparison %in% c("ddid_proband vs sibling", "noddid_proband vs sibling")) %>%
    mutate(
        group      = if_else(comparison == "ddid_proband vs sibling", "DDID", "NoDDID"),
        go_name    = factor(go_name, levels = pathway_order_sib_syngo),
        color_group = case_when(
            p_fdr < sig_thresh & group == "DDID"   ~ "DDID",
            p_fdr < sig_thresh & group == "NoDDID" ~ "NoDDID",
            TRUE                                    ~ "ns"
        )
    ) %>%
    ggplot(aes(x = rel_rr, y = go_name, color = color_group)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h),
        position = position_dodge(width = 0.6),
        size = 0.2, linewidth = 0.3
    ) +
    scale_color_manual(
        values = c("DDID" = "#C0392B", "NoDDID" = "#4472C4", "ns" = "grey70"),
        labels = c("DDID" = "DDID (FDR < 0.05)", "NoDDID" = "NoDDID (FDR < 0.05)", "ns" = "ns"),
        name   = NULL
    ) +
    facet_grid(go_domain ~ ., scales = "free_y", space = "free_y") +
    coord_cartesian(xlim = c(NA, 30)) +
    labs(
        x     = "Relative risk vs siblings",
        y     = NULL,
        title = "SynGO proband enrichment vs siblings"
    ) +
    theme(
        legend.position  = "top",
        axis.text.y      = element_text(size = 5),
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_syngo_vs_sib

f_save_plot(
    plot_obj  = p_syngo_vs_sib,
    plot_name = "05c_syngo_asd_d_asd_da_vs_sibling_RR.pdf",
    overwrite = TRUE,
    height    = 22,
    width     = 6,
    dpi       = 300
)
```

## 3.3.4 SynGO total mutations vs DDID:NoDDID relative risk

```{r plot_syngo_bubble, fig.height = 2.5, fig.width = 7}
# Total observed de novo mutations across DDID + NoDDID probands per pathway
df_total_mut <- df_oe_syngo %>%
    filter(cohort %in% c("ddid_proband", "noddid_proband")) %>%
    group_by(go_id) %>%
    summarise(total_obs = sum(obs, na.rm = TRUE), .groups = "drop")

df_ms_syngo_bubble <- df_scatter_syngo %>%
    left_join(df_total_mut, by = "go_id") %>%
    mutate(go_label = str_remove(go_name, "\\s*\\(GO:[0-9]+\\)\\s*$")) %>%
    arrange(sig_group == "Neither")   # sig points on top

col_sig_group_depr <- c(
    "Both"       = "#2E7D32",
    "DDID only"  = "#C0392B",
    "NoDDID only"= "#4472C4",
    "Neither"    = "grey75"
)

p_ms_syngo_bubble <- df_ms_syngo_bubble %>%
    ggplot(aes(x = total_obs, y = rel_rr_asd_d_asd_da, color = sig_group)) +
    geom_hline(yintercept = 1, linetype = "dashed",
               color = "grey50", linewidth = 0.4) +
    geom_point(aes(size = n_matched), alpha = 0.85) +
    ggrepel::geom_text_repel(
        data = . %>% filter(sig_group != "Neither"),
        aes(label = str_wrap(go_label, 25)),
        size = 2, max.overlaps = Inf,
        box.padding = 0.4, point.padding = 0.3,
        min.segment.length = 0,
        segment.color = "grey50", segment.size = 0.3,
        show.legend = FALSE
    ) +
    scale_color_manual(values = col_sig_group_depr, name = "Sig vs siblings") +
    scale_size_continuous(range = c(1, 5), name = "# pathway genes") +
    scale_y_log10() +
    facet_wrap(~ go_domain, ncol = 2) +
    labs(
        x     = "Total de novo mutations (DDID + NoDDID probands)",
        y     = "DDID:NoDDID relative risk (log scale)",
        title = "SynGO DDID vs NoDDID enrichment"
    ) +
    theme(
        legend.position  = "right",
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_ms_syngo_bubble

f_save_plot(
    plot_obj  = p_ms_syngo_bubble,
    plot_name = "ms_syngo-bubble.pdf",
    overwrite = TRUE,
    height    = 2.5,
    width     = 7,
    dpi       = 300
)
```

## 3.3.5 GO O/E ratio per pathway per cohort

Forest plots are restricted to FDR-significant pathways in at least one cohort;
the full results are in the exported table.

```{r plot_go_oe, fig.height = 14, fig.width = 7}
df_oe_go_sig <- df_oe_go %>%
    group_by(go_id) %>%
    filter(any(p_fdr < sig_thresh)) %>%
    ungroup()

p_go_oe <- df_oe_go_sig %>%
    mutate(
        go_name = factor(go_name, levels = rev(pathway_order_go)),
        cohort  = factor(
            cohort,
            levels = c("ddid_proband", "noddid_proband", "proband", "sibling"),
            labels = cohort_labels[c("ddid_proband", "noddid_proband", "proband", "sibling")]
        )
    ) %>%
    ggplot(aes(x = oe, y = go_name, color = cohort)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h),
        position = position_dodge(width = 0.7),
        size = 0.2, linewidth = 0.3
    ) +
    scale_color_manual(values = setNames(col_cohort, cohort_labels[names(col_cohort)])) +
    facet_grid(go_domain ~ ., scales = "free_y", space = "free_y") +
    labs(
        x     = "Observed / Expected",
        y     = NULL,
        color = NULL,
        title = "GO pathway O/E ratio by cohort (FDR < 0.05 in ≥ 1 cohort)"
    ) +
    theme(
        legend.position  = "top",
        axis.text.y      = element_text(size = 5),
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_go_oe

f_save_plot(
    plot_obj  = p_go_oe,
    plot_name = "04d_go_OE_ratio.pdf",
    overwrite = TRUE,
    height    = 14,
    width     = 7,
    dpi       = 300
)
```

## 3.3.6 GO DDID vs NoDDID enrichment

```{r plot_go_asd_d_asd_da, fig.height = 14, fig.width = 5}
df_pairwise_go_asd_d_sig <- df_pairwise_go %>%
    filter(comparison == "ddid_proband vs noddid_proband", p_fdr < sig_thresh)

pathway_order_go_asd_d <- df_pairwise_go %>%
    filter(comparison == "ddid_proband vs noddid_proband",
           go_id %in% df_pairwise_go_asd_d_sig$go_id) %>%
    arrange(go_domain, rel_rr) %>%
    pull(go_name)

p_go_asd_d <- df_pairwise_go %>%
    filter(comparison == "ddid_proband vs noddid_proband",
           go_id %in% df_pairwise_go_asd_d_sig$go_id) %>%
    mutate(
        go_name = factor(go_name, levels = pathway_order_go_asd_d),
        sig_lab = case_when(
            p_fdr < 0.001      ~ "***",
            p_fdr < 0.01       ~ "**",
            p_fdr < sig_thresh ~ "*",
            TRUE               ~ ""
        ),
        sig = p_fdr < sig_thresh
    ) %>%
    ggplot(aes(x = rel_rr, y = go_name)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h, color = sig),
        size = 0.2, linewidth = 0.3
    ) +
    geom_text(
        aes(x = ci_h, label = sig_lab),
        hjust = -0.3, size = 2, color = "black"
    ) +
    scale_color_manual(
        values = c("TRUE" = "#C0392B", "FALSE" = "grey60"),
        guide  = NULL
    ) +
    facet_grid(go_domain ~ ., scales = "free_y", space = "free_y") +
    labs(
        x     = "Relative risk (DDID / NoDDID)",
        y     = NULL,
        title = "GO DDID vs NoDDID (FDR < 0.05)"
    ) +
    theme(
        axis.text.y      = element_text(size = 5),
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_go_asd_d

f_save_plot(
    plot_obj  = p_go_asd_d,
    plot_name = "04d_go_asd_d_vs_asd_da_RR.pdf",
    overwrite = TRUE,
    height    = 14,
    width     = 5,
    dpi       = 300
)
```

## 3.3.7 GO DDID and NoDDID vs siblings

```{r plot_go_vs_sib, fig.height = 14, fig.width = 6}
df_pairwise_go_sib_sig <- df_pairwise_go %>%
    filter(comparison %in% c("ddid_proband vs sibling", "noddid_proband vs sibling"),
           p_fdr < sig_thresh)

pathway_order_go_sib <- df_pairwise_go %>%
    filter(comparison == "ddid_proband vs sibling",
           go_id %in% df_pairwise_go_sib_sig$go_id) %>%
    arrange(go_domain, rel_rr) %>%
    pull(go_name)

p_go_vs_sib <- df_pairwise_go %>%
    filter(comparison %in% c("ddid_proband vs sibling", "noddid_proband vs sibling"),
           go_id %in% df_pairwise_go_sib_sig$go_id) %>%
    mutate(
        group       = if_else(comparison == "ddid_proband vs sibling", "DDID", "NoDDID"),
        go_name     = factor(go_name, levels = pathway_order_go_sib),
        color_group = case_when(
            p_fdr < sig_thresh & group == "DDID"   ~ "DDID",
            p_fdr < sig_thresh & group == "NoDDID" ~ "NoDDID",
            TRUE                                    ~ "ns"
        )
    ) %>%
    ggplot(aes(x = rel_rr, y = go_name, color = color_group)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_pointrange(
        aes(xmin = ci_l, xmax = ci_h),
        position = position_dodge(width = 0.6),
        size = 0.2, linewidth = 0.3
    ) +
    scale_color_manual(
        values = c("DDID" = "#C0392B", "NoDDID" = "#4472C4", "ns" = "grey70"),
        labels = c("DDID" = "DDID (FDR < 0.05)", "NoDDID" = "NoDDID (FDR < 0.05)", "ns" = "ns"),
        name   = NULL
    ) +
    facet_grid(go_domain ~ ., scales = "free_y", space = "free_y") +
    coord_cartesian(xlim = c(NA, 30)) +
    labs(
        x     = "Relative risk vs siblings",
        y     = NULL,
        title = "GO proband enrichment vs siblings (FDR < 0.05)"
    ) +
    theme(
        legend.position  = "top",
        axis.text.y      = element_text(size = 5),
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_go_vs_sib

f_save_plot(
    plot_obj  = p_go_vs_sib,
    plot_name = "04d_go_asd_d_asd_da_vs_sibling_RR.pdf",
    overwrite = TRUE,
    height    = 14,
    width     = 6,
    dpi       = 300
)
```

## 3.3.8 GO DDID vs NoDDID scatter (each vs siblings)

Each point is a GO pathway; x = NoDDID vs sibling RR, y = DDID vs sibling RR.
Points are coloured by significance group; only NoDDID-only significant pathways are labelled.

```{r plot_go_scatter, fig.height = 4, fig.width = 12}
df_scatter_go <- df_pairwise_go %>%
    filter(comparison %in% c("ddid_proband vs sibling", "noddid_proband vs sibling")) %>%
    select(go_id, go_name, go_domain, category, comparison,
           rel_rr, ci_l, ci_h, n_matched, p_fdr) %>%
    mutate(group = if_else(comparison == "ddid_proband vs sibling", "ddid", "noddid")) %>%
    select(-comparison) %>%
    pivot_wider(
        id_cols     = c(go_id, go_name, go_domain, category, n_matched),
        names_from  = group,
        values_from = c(rel_rr, ci_l, ci_h, p_fdr)
    ) %>%
    left_join(
        df_pairwise_go %>%
            filter(comparison == "ddid_proband vs noddid_proband") %>%
            select(go_id, p_fdr_asd_d_asd_da = p_fdr, rel_rr_asd_d_asd_da = rel_rr),
        by = "go_id"
    ) %>%
    mutate(
        sig_asd_d_asd_da = replace_na(p_fdr_asd_d_asd_da < sig_thresh, FALSE),
        sig_asd_d        = replace_na(p_fdr_asd_d   < sig_thresh, FALSE),
        sig_asd_da      = replace_na(p_fdr_asd_da < sig_thresh, FALSE),
        sig_group = case_when(
            sig_asd_d  &  sig_asd_da ~ "Both",
            sig_asd_d  & !sig_asd_da ~ "DDID only",
            !sig_asd_d &  sig_asd_da ~ "NoDDID only",
            TRUE                    ~ "Neither"
        ),
        sig_group = factor(sig_group,
                           levels = c("Both", "DDID only", "NoDDID only", "Neither"))
    )

p_go_scatter <- df_scatter_go %>%
    arrange(sig_group != "Neither") %>%
    ggplot(aes(x = rel_rr_asd_da, y = rel_rr_asd_d, color = sig_group)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(size = n_matched), alpha = 0.7) +
    ggrepel::geom_text_repel(
        data = . %>% filter(sig_group == "NoDDID only" & rel_rr_asd_da > 1),
        aes(label = str_wrap(go_name, width = 25)),
        size = 2.2, max.overlaps = 30,
        nudge_x = 0.3, nudge_y = -0.3,
        direction = "both",
        min.segment.length = 0,
        segment.color = "grey50", segment.size = 0.3,
        box.padding = 0.4
    ) +
    scale_color_manual(values = col_sig_group, name = "Sig vs siblings") +
    scale_size_continuous(range = c(0.5, 4), name = "# pathway genes") +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ go_domain, ncol = 3) +
    labs(
        x     = "NoDDID vs siblings (log RR)",
        y     = "DDID vs siblings (log RR)",
        title = "GO DDID vs NoDDID enrichment relative to siblings"
    ) +
    theme(
        legend.position  = "right",
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_go_scatter

f_save_plot(
    plot_obj  = p_go_scatter,
    plot_name = "04d_go_asd_d_asd_da_scatter.png",
    overwrite = TRUE,
    height    = 4,
    width     = 12,
    dpi       = 300
)
```

## 3.3.9 GO DDID vs NoDDID O/E scatter

Each point is a GO pathway; x = NoDDID O/E, y = DDID O/E. Colour and labels
follow the same significance grouping as above.

```{r plot_go_scatter_oe, fig.height = 4, fig.width = 12}
df_scatter_oe_go <- df_oe_go %>%
    filter(cohort %in% c("ddid_proband", "noddid_proband")) %>%
    select(go_id, go_name, go_domain, n_matched, cohort, oe, ci_l, ci_h) %>%
    mutate(cohort = if_else(cohort == "ddid_proband", "ddid", "noddid")) %>%
    pivot_wider(
        id_cols     = c(go_id, go_name, go_domain, n_matched),
        names_from  = cohort,
        values_from = c(oe, ci_l, ci_h)
    ) %>%
    left_join(
        df_scatter_go %>% select(go_id, sig_group),
        by = "go_id"
    )

p_go_scatter_oe <- df_scatter_oe_go %>%
    arrange(sig_group != "Neither") %>%
    ggplot(aes(x = oe_asd_da, y = oe_asd_d, color = sig_group)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey50", linewidth = 0.4) +
    geom_vline(xintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_hline(yintercept = 1, linetype = "dotted", color = "grey70", linewidth = 0.3) +
    geom_point(aes(size = n_matched), alpha = 0.7) +
    ggrepel::geom_text_repel(
        data = . %>% filter(sig_group == "NoDDID only"),
        aes(label = str_wrap(go_name, width = 25)),
        size = 2.2, max.overlaps = 30,
        nudge_x = 0.3, nudge_y = -0.3,
        direction = "both",
        min.segment.length = 0,
        segment.color = "grey50", segment.size = 0.3,
        box.padding = 0.4
    ) +
    scale_color_manual(values = col_sig_group, name = "Sig vs siblings") +
    scale_size_continuous(range = c(0.5, 4), name = "# pathway genes") +
    scale_x_log10() +
    scale_y_log10() +
    facet_wrap(~ go_domain, ncol = 3) +
    labs(
        x     = "NoDDID O/E (log scale)",
        y     = "DDID O/E (log scale)",
        title = "GO DDID vs NoDDID O/E"
    ) +
    theme(
        legend.position  = "right",
        strip.text       = element_text(face = "bold"),
        strip.background = element_blank()
    )

p_go_scatter_oe

f_save_plot(
    plot_obj  = p_go_scatter_oe,
    plot_name = "04d_go_asd_d_asd_da_scatter_OE.pdf",
    overwrite = TRUE,
    height    = 7,
    width     = 12,
    dpi       = 300
)
```

## 3.3.10 GO O/E forest plot — significant pathways (top by p-value)

Significant pathways only (FDR < 0.05 in at least one of DDID vs siblings or NoDDID vs
siblings). Ordered by DDID O/E within each domain; patchwork stacks sig-group rows with
heights proportional to pathway count. Shared legend sits above all rows.

```{r plot_go_oe_forest, fig.height = 8, fig.width = 6}
sig_go_ids_all <- df_pairwise_go %>%
    filter(
        comparison %in% c("ddid_proband vs sibling", "noddid_proband vs sibling"),
        p_fdr < sig_thresh
    ) %>%
    pull(go_id) %>%
    unique()

# "Both": top 5 per ontology by min(p_fdr_asd_d, p_fdr_asd_da)
top5_both_ids <- df_oe_go %>%
    filter(cohort %in% c("ddid_proband", "noddid_proband"),
           go_id %in% sig_go_ids_all) %>%
    select(go_id, go_domain, cohort, p_fdr) %>%
    left_join(df_scatter_go %>% select(go_id, sig_group), by = "go_id") %>%
    filter(sig_group == "Both") %>%
    pivot_wider(id_cols = c(go_id, go_domain),
                names_from  = cohort,
                values_from = p_fdr,
                names_prefix = "p_fdr_") %>%
    mutate(min_p_fdr = pmin(p_fdr_asd_d_proband, p_fdr_asd_da_proband, na.rm = TRUE)) %>%
    group_by(go_domain) %>%
    slice_min(min_p_fdr, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    pull(go_id)

# "DDID only": top 5 per ontology by DDID p_fdr
top5_asd_d_only_ids <- df_oe_go %>%
    filter(cohort == "ddid_proband", go_id %in% sig_go_ids_all) %>%
    left_join(df_scatter_go %>% select(go_id, sig_group), by = "go_id") %>%
    filter(sig_group == "DDID only") %>%
    group_by(go_domain) %>%
    slice_min(p_fdr, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    pull(go_id)

# "NoDDID only": keep all
noddid_only_ids <- df_scatter_go %>%
    filter(sig_group == "NoDDID only", go_id %in% sig_go_ids_all) %>%
    pull(go_id)

sig_go_ids_go <- unique(c(top5_both_ids, top5_asd_d_only_ids, noddid_only_ids))

# Pathway order: DDID O/E descending within each domain
forest_order_go <- df_oe_go %>%
    filter(cohort == "ddid_proband", go_id %in% sig_go_ids_go) %>%
    arrange(go_domain, desc(oe)) %>%
    pull(go_name)

df_oe_forest_go <- df_oe_go %>%
    filter(
        cohort %in% c("ddid_proband", "noddid_proband"),
        go_id  %in% sig_go_ids_go
    ) %>%
    left_join(
        df_scatter_go %>% select(go_id, sig_group),
        by = "go_id"
    ) %>%
    mutate(
        cohort_label = if_else(cohort == "ddid_proband", "DDID", "NoDDID"),
        point_sig = case_when(
            cohort_label == "DDID"   & sig_group %in% c("Both", "DDID only")   ~ "DDID",
            cohort_label == "NoDDID" & sig_group %in% c("Both", "NoDDID only") ~ "NoDDID",
            TRUE                                                                 ~ "ns"
        ) %>% factor(levels = c("DDID", "NoDDID", "ns")),
        go_name_clean = str_wrap(go_name, width = 30),
        go_name_clean = factor(go_name_clean,
                               levels = rev(str_wrap(forest_order_go, width = 30))),
        sig_group     = factor(sig_group,
                               levels = c("Both", "DDID only", "NoDDID only", "Neither"))
    )

# Shared legend extracted from a dummy plot with all levels guaranteed
legend_dummy_go <- ggplot(
    tibble(x = 1, y = 1,
           point_sig = factor(names(col_point_sig), levels = names(col_point_sig))),
    aes(x, y, color = point_sig)
) +
    geom_point() +
    scale_color_manual(values = col_point_sig, labels = lab_point_sig,
                       name = NULL, drop = FALSE) +
    theme(legend.position = "top",
          legend.key.size = unit(0.35, "cm"),
          legend.text     = element_text(size = 6))

shared_legend_go <- cowplot::get_legend(legend_dummy_go)

# Helper: one sub-plot per sig_group (faceted by go_domain)
make_forest_row_go <- function(df, grp, show_title = FALSE) {
    ggplot(df, aes(x = oe, y = go_name_clean, color = point_sig)) +
        geom_vline(xintercept = 1, linetype = "dashed",
                   color = "grey50", linewidth = 0.4) +
        geom_pointrange(
            aes(xmin = ci_l, xmax = ci_h),
            position = position_dodge(width = 0.6),
            size = 0.25, linewidth = 0.4
        ) +
        scale_color_manual(values = col_point_sig, labels = lab_point_sig,
                           name = NULL, drop = FALSE) +
        facet_grid2(. ~ go_domain, independent = "all", scales = "free") +
        labs(
            x     = if (show_title) "Observed / expected" else NULL,
            y     = NULL,
            title = grp
        ) +
        theme(
            legend.position  = "none",
            axis.text.y      = element_text(size = 7),
            axis.text.x      = element_text(size = 7),
            axis.title.x     = element_text(size = 7),
            plot.title       = element_text(size = 7, face = "bold", hjust = 0.5),
            plot.margin      = margin(t = 1, r = 5, b = 1, l = 5),
            strip.text       = element_text(face = "bold", size = 7),
            strip.background = element_blank(),
            panel.spacing    = unit(0.1, "lines")
        )
}

n_paths_go <- df_oe_forest_go %>%
    filter(sig_group %in% sig_group_levels) %>%
    group_by(sig_group) %>%
    summarise(n = n_distinct(go_name_clean), .groups = "drop") %>%
    mutate(sig_group = factor(sig_group, levels = sig_group_levels)) %>%
    arrange(sig_group)

forest_rows_go <- map(sig_group_levels, function(grp) {
    make_forest_row_go(
        df         = df_oe_forest_go %>% filter(sig_group == grp),
        grp        = grp,
        show_title = grp == tail(sig_group_levels, 1)
    )
})

p_go_oe_forest <- patchwork::wrap_plots(
    c(list(patchwork::wrap_elements(shared_legend_go)), forest_rows_go),
    ncol    = 1,
    heights = c(0.8, n_paths_go$n)
) +
    patchwork::plot_annotation(title = "GO O/E — significant pathways")

p_go_oe_forest

f_save_plot(
    plot_obj  = p_go_oe_forest,
    plot_name = "04d_go_OE_forest_sig.png",
    overwrite = TRUE,
    height    = 8,
    width     = 6,
    dpi       = 300
)
```

## 3.3.11 GO O/E forest plot — top DDID RR vs siblings

Same layout as above but "Both" and "DDID only" rows are limited to the top 5 pathways
per ontology by highest DDID vs siblings relative risk (rather than lowest p-value).

```{r plot_go_oe_forest_rr, fig.height = 8, fig.width = 6}
top5_rr_ids <- df_pairwise_go %>%
    filter(comparison == "ddid_proband vs sibling", go_id %in% sig_go_ids_all) %>%
    left_join(df_scatter_go %>% select(go_id, sig_group), by = "go_id") %>%
    filter(sig_group %in% c("Both", "DDID only")) %>%
    group_by(go_domain, sig_group) %>%
    slice_max(rel_rr, n = 5, with_ties = FALSE) %>%
    ungroup() %>%
    pull(go_id)

sig_go_ids_rr <- unique(c(top5_rr_ids, noddid_only_ids))

forest_order_go_rr <- df_oe_go %>%
    filter(cohort == "ddid_proband", go_id %in% sig_go_ids_rr) %>%
    arrange(go_domain, desc(oe)) %>%
    pull(go_name)

df_oe_forest_go_rr <- df_oe_go %>%
    filter(
        cohort %in% c("ddid_proband", "noddid_proband"),
        go_id  %in% sig_go_ids_rr
    ) %>%
    left_join(
        df_scatter_go %>% select(go_id, sig_group),
        by = "go_id"
    ) %>%
    mutate(
        cohort_label = if_else(cohort == "ddid_proband", "DDID", "NoDDID"),
        point_sig = case_when(
            cohort_label == "DDID"   & sig_group %in% c("Both", "DDID only")   ~ "DDID",
            cohort_label == "NoDDID" & sig_group %in% c("Both", "NoDDID only") ~ "NoDDID",
            TRUE                                                                 ~ "ns"
        ) %>% factor(levels = c("DDID", "NoDDID", "ns")),
        go_name_clean = str_wrap(go_name, width = 30),
        go_name_clean = factor(go_name_clean,
                               levels = rev(str_wrap(forest_order_go_rr, width = 30))),
        sig_group     = factor(sig_group,
                               levels = c("Both", "DDID only", "NoDDID only", "Neither"))
    )

n_paths_go_rr <- df_oe_forest_go_rr %>%
    filter(sig_group %in% sig_group_levels) %>%
    group_by(sig_group) %>%
    summarise(n = n_distinct(go_name_clean), .groups = "drop") %>%
    mutate(sig_group = factor(sig_group, levels = sig_group_levels)) %>%
    arrange(sig_group)

forest_rows_go_rr <- map(sig_group_levels, function(grp) {
    make_forest_row_go(
        df         = df_oe_forest_go_rr %>% filter(sig_group == grp),
        grp        = grp,
        show_title = grp == tail(sig_group_levels, 1)
    )
})

p_go_oe_forest_rr <- patchwork::wrap_plots(
    c(list(patchwork::wrap_elements(shared_legend_go)), forest_rows_go_rr),
    ncol    = 1,
    heights = c(0.8, n_paths_go_rr$n)
) +
    patchwork::plot_annotation(title = "GO O/E — top DDID RR vs siblings")

p_go_oe_forest_rr

f_save_plot(
    plot_obj  = p_go_oe_forest_rr,
    plot_name = "04d_go_OE_forest_top_RR.png",
    overwrite = TRUE,
    height    = 10,
    width     = 10,
    dpi       = 300
)
```
