---
title: "μBrain gene-set developmental enrichment"
description: "Tests whether the ASD functional programs (GR/SYN/MORPH) are enriched across laminar zones and cortical regions of the prenatal cortex (μBrain, Ball et al. 2024), above a whole-transcriptome or an ASD-gene background, via linear mixed models."
format: html
---

# --- 0) Setup --- #

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# libraries, directories, topic_colors (incl. lme4/lmerTest/emmeans/ggforce)
project_dir <- "~/Documents/Postdoc/projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

results_dir  <- file.path(analysis_dir, "04_brain_spatial_enrichment")
supp_fig_dir <- file.path(figures_dir, "supplement")
```

## 0.1 Load study data

```{r load_data}
# ASD rare variant data
df_asd_rarevar_zscore <- readRDS(file.path(analysis_dir, "00_data_format", "00_01_ASD-rarevar-stats.RDS"))
asd_genes <- df_asd_rarevar_zscore %>% filter(FDR < 0.001 & !FLAG) %>% pull(gene_name)

# Gene-topic assignments (GR / SYN / MORPH), from 03_02
df_gene_assignments <- read_csv(
    file.path(analysis_dir, "03_module_enrichments", "03_02_gene-topic-assignment.csv"),
    show_col_types = FALSE)
gene_lists <- df_gene_assignments %>%
    split(.$GO_topic) %>%
    map(~ pull(.x, gene_name) %>% unique())
```


## 0.2 Load and format μBrain data

```{r load_mubrain}
# μBrain laser-microdissection microarray of the prenatal human cortex
# (Ball et al. 2024; Zenodo 10.5281/zenodo.10622336)
# Long format: one row per gene x sample, annotated by tissue (laminar zone) and cortical_region.
expression_url  <- "https://zenodo.org/records/10622337/files/uBrain-processed-lmd-data.csv?download=1"
expression_data <- read_csv(expression_url, show_col_types = FALSE)

# z-score each gene's expression across samples (fair comparison across genes
# with different baseline expression)
scaled_data <- expression_data %>%
    group_by(gene) %>%
    mutate(scaled_expr = scale(expression)[, 1]) %>%
    ungroup()

# mean scaled expression per topic x zone x region (feeds the fan geometry)
module_spatial_profiles <- scaled_data %>%
    inner_join(df_gene_assignments, by = join_by(gene == gene_name)) %>%
    group_by(GO_topic, tissue, cortical_region) %>%
    summarize(mean_scaled_expr = mean(scaled_expr, na.rm = TRUE),
              n_genes_present  = n_distinct(gene),
              .groups = "drop")
```


# --- 1) Gene-set enrichment above background (linear mixed models) --- #

Each topic is tested against a **background** gene set:
`member` = gene is in the topic (1) vs background (0), interacted with region / zone.

Because expression is Z-scored per gene, the region/zone main effect absorbs the
background's spatial pattern, so the `member` contrast (topic − background, via
emmeans) is enrichment *above background*. 

The region model is fit separately within each of the 5 zones, 
and the laminar model pools the balanced regions with `(1 | cortical_region:tissue)`.

## 1.1 Helper functions

```{r enrich_helpers}
topics <- c("GR", "SYN", "MORPH")

# Keep topic + background genes, tag membership (topic vs background)
add_member <- function(data, topic_genes, background_genes) {
    data %>%
        filter(gene %in% union(topic_genes, background_genes)) %>%
        mutate(member = factor(if_else(gene %in% topic_genes, "topic", "background"),
                               levels = c("background", "topic")))
}

# Per-region enrichment above background, fit SEPARATELY within each zone
enrich_region_by_tissue <- function(topic, background_genes, data) {
    add_member(data, gene_lists[[topic]], background_genes) %>%
        nest(.by = tissue) %>%
        mutate(res = map(data, function(d) {
            
            ## Main LME ##
            m <- lmer(scaled_expr ~ member * cortical_region + (1 | gene), data = d)
            
            ## Extract per-topic contrasts ##
            emmeans(m, ~ member | cortical_region, lmer.df = "asymptotic") %>%
                contrast("revpairwise") %>%              # topic - background
                summary(infer = TRUE, adjust = "BH") %>%
                as_tibble() %>%
                rename(statistic = any_of("t.ratio")) %>%
                rename(statistic = any_of("z.ratio")) %>%
                transmute(cortical_region, enrichment = estimate, statistic,
                          FDR = p.value, significant = FDR < 0.05)
        })) %>%
        dplyr::select(tissue, res) %>%
        unnest(res) %>%
        mutate(topic = topic)
}

# Per-zone (laminar) enrichment above background, on the balanced regions
enrich_layer <- function(topic, background_genes, data, regions) {
    d <- add_member(data, gene_lists[[topic]], background_genes) %>%
        filter(cortical_region %in% regions)
    
    ## Main LME ##
    m <- lmer(scaled_expr ~ member * tissue + (1 | gene) + (1 | cortical_region:tissue), data = d)
    ##
    
    ## Extract per-topic contrasts ##
    emmeans(m, ~ member | tissue, lmer.df = "asymptotic") %>%
        contrast("revpairwise") %>%
        summary(infer = TRUE, adjust = "BH") %>%
        as_tibble() %>%
        rename(statistic = any_of("t.ratio")) %>%
        rename(statistic = any_of("z.ratio")) %>%
        transmute(tissue, enrichment = estimate, statistic,
                  FDR = p.value, significant = FDR < 0.05, topic = topic)
}
```

## 1.2 Whole-transcriptome background

```{r enrich_transcriptome}
# Background: the whole measured transcriptome in mubrain (n = ~10k genes)
# The full ~10k-gene background makes these mixed models very large (10k gene
# random effects x 5 zones x n_topics), so we draw a large random background --
# the transcriptome mean is stable with a few thousand genes. Set n_background = Inf
# to use every gene (slower).
n_background <- Inf
#set.seed(1)
bg_transcriptome <- {
    g <- setdiff(unique(scaled_data$gene), unlist(gene_lists))
    if (is.finite(n_background)) sample(g, min(n_background, length(g))) else g
}

# --- Regions --- #
df_region_enrich_bg <- map_dfr(topics, ~ enrich_region_by_tissue(.x, bg_transcriptome, scaled_data))

# --- Layers --- #

# laminar zones, deep (ventricular) to superficial (cortical plate)
zone_levels <- c("ventricular", "subventricular", "intermediate", "subplate", "cortical_plate")

# Regions sampled in every zone (drops paleocortex, retrosplenial, and the
# partially-sampled insula/temporal/limbic regions) so layers are comparable
balanced_regions <- scaled_data %>%
    inner_join(df_gene_assignments, by = join_by(gene == gene_name)) %>%
    distinct(cortical_region, tissue) %>%
    count(cortical_region) %>%
    filter(n == length(zone_levels)) %>%
    pull(cortical_region)

# Run LME
df_layer_enrich_bg <- map_dfr(topics, ~ enrich_layer(.x, bg_transcriptome, scaled_data, balanced_regions))

df_layer_enrich_bg
```

## 1.3 ASD-gene background

Same models, but the background is the set of ASD risk genes, so this asks whether
each topic is enriched *relative to ASD genes generally* rather than the whole
transcriptome.

```{r enrich_asd}
# Background (b): all ASD risk genes (TADA FDR < 0.001, unflagged).
df_region_enrich_asd <- map_dfr(topics, ~ enrich_region_by_tissue(.x, asd_genes, scaled_data))
df_layer_enrich_asd  <- map_dfr(topics, ~ enrich_layer(.x, asd_genes, scaled_data, balanced_regions))

df_layer_enrich_asd
```

## 1.4 Export results as tables

```{r export_results}
write_xlsx(
    list(
        "A. transcriptome-wide bg" = df_region_enrich_bg,
        "B. ASD gene bg" = df_region_enrich_asd
    ),
    path = file.path(results_dir, "04_04_mubrain-developmental-enrichment.xlsx"),
)
```



# --- 2) Plots--- #

The fan reproduces the wedge layout of Ball et al. 2024 (μBrain, Fig. 2e): each
wedge is one laminar zone × cortical region, drawn as an annular sector. Radius =
laminar depth (ventricular innermost → cortical plate outermost), angle = cortical
region ordered anterior → posterior. Fill = per-region enrichment above background
(§1); black outline = significantly enriched (FDR < 0.05).

## 2.1 Define fan geometry

```{r fan_geometry}
# --- Axis ordering --------------------------------------------------------- #

# Radius grows outward (zone_levels above, VZ innermost) so the plot mirrors cortical anatomy.
zone_labels <- c("VZ", "SVZ", "IZ", "SP", "CP")

# Anterior -> posterior ordering of the 27 μBrain cortical regions.
# The atlas orders regions on a continuous geodesic A-P axis
# (surface_data/parcellated-axis-data.csv), which is not released with the
# resource; this vector is an anatomical approximation of that axis
region_ap_order <- c(
  "frontal_pole", "orbitofrontal", "rostral_cingulate", "subcallosal_cingulate",
  "dorsolateral_pfc", "ventrolateral_pfc", "motor", "midcingulate",
  "granular_insula", "dysgranular_insula",
  "sensory", "dorsal_parietal", "ventral_parietal", "caudal_cingulate", "retrosplenial",
  "auditory", "dorsolateral_temporal", "ventrolateral_temporal",
  "entorhinal", "perirhinal", "parahippocampal", "hippocampus", "paleocortex",
  "lateral_temporal-occipital", "medial_temporal-occipital", "extrastriate", "primary_visual"
)


# --- Wedge geometry -------------------------------------------------------- #

# Fan spans a 220-degree arc across the top, anterior region on the left,
# matching the start/end angles used in the published figure.
arc_span_deg  <- 220
arc_start_deg <- -110          # left (anterior) end, measured clockwise from 12 o'clock
region_gap    <- 0.12          # angular gap between neighbouring region wedges
zone_gap      <- 0.10          # radial gap between zone rings

fan_geom <- module_spatial_profiles %>%
  filter(tissue %in% zone_levels, cortical_region %in% region_ap_order) %>%
  mutate(
    zone_i   = match(tissue, zone_levels),            # 1 = inner (VZ) .. 5 = outer (CP)
    region_i = match(cortical_region, region_ap_order)
  )

n_region <- length(region_ap_order)
step_deg <- arc_span_deg / n_region

fan_geom <- fan_geom %>%
  mutate(
    start = (arc_start_deg + (region_i - 1) * step_deg) * pi / 180,
    end   = (arc_start_deg + (region_i - 1) * step_deg + step_deg * (1 - region_gap)) * pi / 180,
    r0    = zone_i,
    r     = zone_i + 1 - zone_gap
  )

# Symmetric, robust colour limits centred on 0 (values are z-scores)
fan_lim <- quantile(abs(fan_geom$mean_scaled_expr), 0.98, na.rm = TRUE)


# --- Ring / spoke labels --------------------------------------------------- #

# ggforce measures angles clockwise from 12 o'clock: x = r*sin(theta), y = r*cos(theta)

# Zone labels (VZ..CP) along the anterior (left) edge of the fan, one per ring,
# rotated to follow the radius (inner VZ -> outer CP)
zone_ang_deg <- arc_start_deg - 0.5 * step_deg      # just left of the first region wedge
zone_lab_df <- tibble(
  label  = zone_labels,
  radius = seq_along(zone_labels) + 0.4
) %>%
  mutate(
    x     = radius * sin(zone_ang_deg * pi / 180),
    y     = radius * cos(zone_ang_deg * pi / 180),
    angle = 90 - zone_ang_deg,
    angle = ifelse(angle > 90 & angle < 270, angle - 180, angle)  # keep upright
  )

# Region labels around the outer edge, one per region, rotated radially
region_lab_df <- tibble(
  cortical_region = region_ap_order,
  mid_deg = arc_start_deg + (seq_len(n_region) - 1) * step_deg + step_deg * (1 - region_gap) / 2
) %>%
  mutate(
    x     = 6.4 * sin(mid_deg * pi / 180),
    y     = 6.4 * cos(mid_deg * pi / 180),
    angle = 90 - mid_deg,
    hjust = ifelse(angle > 90 & angle < 270, 1, 0),
    angle = ifelse(angle > 90 & angle < 270, angle - 180, angle) # keep text upright
  ) %>%
    mutate(label = case_when(
        cortical_region == "frontal_pole" ~ "anterior",
        cortical_region == "primary_visual" ~ "posterior",
        TRUE ~ NA_character_
    ))

# --- (tissue, region) -> wedge geometry lookup used by the plot helper ------ #
fan_grid <- expand_grid(tissue = zone_levels, cortical_region = region_ap_order) %>%
    mutate(zone_i   = match(tissue, zone_levels),
           region_i = match(cortical_region, region_ap_order),
           start = (arc_start_deg + (region_i - 1) * step_deg) * pi / 180,
           end   = (arc_start_deg + (region_i - 1) * step_deg + step_deg * (1 - region_gap)) * pi / 180,
           r0    = zone_i,
           r     = zone_i + 1 - zone_gap)
```

## 2.2 Plot helper

```{r fan_helper}
# One fan for a single topic: fill = enrichment above background (white anchored
# at 0), black outline = significantly enriched (FDR < 0.05 & statistic > 0).
f_plot_fan_enrich <- function(df, topic_sel, topic_colors) {
    dat <- df %>%
        filter(topic == topic_sel) %>%
        inner_join(fan_grid, by = c("tissue", "cortical_region"))

    # robust upper limit; white anchored at 0 (= background level). Use
    # limits = range(dat$enrichment) instead if you also want to show depletion.
    lim <- quantile(dat$enrichment, 0.98, na.rm = TRUE)

    ggplot(dat) +
        geom_arc_bar(aes(x0 = 0, y0 = 0, r0 = r0, r = r, start = start, end = end,
                         fill = enrichment, color = FDR < 0.05 & statistic > 0),
                     linewidth = 0.20) +
        geom_text(data = region_lab_df %>% mutate(cortical_region = str_replace_all(cortical_region, "_", " ")),
                  aes(x = x, y = y, label = cortical_region, hjust = hjust, angle = angle),
                  size = 1.8, color = "grey20") +
        geom_text(data = zone_lab_df, aes(x = x, y = y, label = label, angle = angle),
                  size = 2, fontface = "bold") +
        scale_fill_gradient2(
            low = "gray40", mid = "white", high = topic_colors[[topic_sel]],
            limits = c(-lim, lim), oob = scales::squish, na.value = "grey85",
            name = paste0(topic_sel, " enrichment\n(vs background)")) +
        scale_color_manual(values = c("TRUE" = "black", "FALSE" = "gray80"),
                           name = "sig. enriched above bg\n(FDR < 0.05)") +
        coord_fixed(xlim = c(-6.8, 6.8), ylim = c(-3, 6.7), clip = "off") +
        theme_void() +
        labs(title = topic_sel) +
        theme(
            plot.title = element_text(size = 6.5, face = "bold", margin = margin(b = -5)),
            legend.position       = "bottom",
            legend.title.position = "top",
            legend.title = element_text(size = 5, hjust = 0.5),
            legend.text  = element_text(size = 4),
            legend.key.height = unit(0.15, "cm"),
            legend.key.width  = unit(0.35, "cm"),
            legend.margin = margin(t = -50),
            plot.margin = margin(0, 0, 0, 0)
        )
}
```

## 2.3 Whole-transcriptome-background fans

```{r fan_plot_bg, fig.width = 6, fig.height = 8}
# Fill = per-region x per-zone enrichment above the transcriptome background
# (effect size = emmeans estimate, z units) from df_region_enrich_bg; one fan per
# topic, sequential white -> topic colour.
p_supp_fan_bg <- map(names(topic_colors),
                     ~ f_plot_fan_enrich(df_region_enrich_bg, .x, topic_colors)) %>%
    wrap_plots(ncol = 1)

p_supp_fan_bg

ggsave(
    filename = file.path(supp_fig_dir, "04_04_mubrain-fan-transcriptome-bg.pdf"),
    plot     = p_supp_fan_bg,
    height   = 8,
    width    = 6,
    dpi      = 300
)
```

## 2.4 ASD-gene-background fans

```{r fan_plot_asd, fig.width = 6, fig.height = 8}
# Same fans, enrichment relative to the ASD risk-gene background (df_region_enrich_asd).
p_supp_fan_asd <- map(names(topic_colors),
                     ~ f_plot_fan_enrich(df_region_enrich_asd, .x, topic_colors)) %>%
    wrap_plots(ncol = 1)

p_supp_fan_asd

ggsave(
    filename = file.path(supp_fig_dir, "04_04_mubrain-fan-ASD-bg.pdf"),
    plot     = p_supp_fan_asd,
    height   = 8,
    width    = 6,
    dpi      = 300
)
```
