---
title: "00b. Format ASD rare variant data"
description: "Convert the TADA FDR values for each gene to Z-scores to get a normally distributed, continuous metric of ASD association for our model. Also add gene length for regression model covariates."
date: 2026-07-06
format: html
---

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r env}
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "00_data_format")
```


# --- 1) Load ASD rare variant data --- #

```{r load_data}
df_asd_rarevar <- read_table(paste0(data_dir, "full_results_wcounts_2025-10-08.txt")) %>% 
    dplyr::rename(
        "gene_name" = "gene",
        "ensembl_gene_id" = "gene_id"
    ) %>% 
    
    # Calculate gene-level Bayes Factor (multiply across variant types)
    mutate(BF = BF_PTV * BF_Mis2 * BF_Mis1 * BF_DEL * BF_DUP) %>%
    dplyr::select(gene_name, ensembl_gene_id, FLAG, FDR, BF) %>% 
    
    # log-transform and scale Bayes Factor
    mutate(log_BF = log10(BF)) %>% 
    mutate(z_log_BF = scale(log_BF)[,1])
```


# --- 2) Convert FDR to Z-score --- #

## 2.1 Write transformation function

[pval = FDR * rank(FDR) / max(FDR) * length(FDR)] \n
Binarize and combine data (modules + ASD risk) into one wide tibble

```{r transform_p_function}
fdr_to_p <- function(fdr_vec, bf_vec = NULL, target_max_z = 10) {
    
    N <- length(fdr_vec)
    
    # --- 1: PREP DATA ---
    # Ensure bf_vec exists; if NULL, use zeros so it doesn't affect ranking
    if(is.null(bf_vec)) bf_vec <- rep(0, N)
    
    # --- 2: COLLAPSE THE BOTTOM (null genes) ---
    # Scientific honesty: If log(BF) is 0, get rid of the alphabetical noise (in datasheet Kathryn & Bernie sent)
    # Force these genes (N = ~7k) to the maximum observed FDR so they group together
    fdr_clean <- fdr_vec
    bf_is_zero <- (bf_vec == 0)
    max_fdr_val <- max(fdr_vec, na.rm = TRUE)
    fdr_clean[bf_is_zero] <- max_fdr_val
    
    # --- 3: CREATE A SORTING KEY ---
    # Rank by FDR (ascending) then by BF (descending).
    # Since rank() works on a single vector, we create a combined score:
    # Low FDR + High BF = top rank
    # Use -bf_vec because higher BF should result in a smaller rank number
    combined_score <- fdr_clean + (rank(-bf_vec, ties.method = "average") / (N * 10))
    
    # --- 4: GENERATE RANKS ---
    # ties.method = "average" ensures that if FDR and BF are identical, they get the exact same rank 
    # (as is the case for the bottom ~7k genes)
    fdr_ranks <- rank(combined_score, ties.method = "average")
    
    # --- 5: TRANSFORM TO P-VALUES ---
    # Use the rank-based power shift to avoid the "p-value wall" at 0
    p_target <- pnorm(target_max_z, lower.tail = FALSE)
    
    # Create base uniform distribution from ranks
    p_base <- (fdr_ranks - 0.5) / N
    
    # Calculate power shift to anchor the top gene to target_max_z
    power_shift <- log(p_target) / log(min(p_base))
    p_vals <- p_base ^ power_shift
    
    # --- 6: SAFETY TRUNCATION ---
    # Cap at 1 - 1e-16 to avoid Z-scores of -Inf
    p_vals <- pmin(1 - 1e-16, p_vals)
    
    return(p_vals)
}
```

## 2.2 Quantify ASD risk as FDR-derived Z-score

```{r run_transform_p}
df_asd_rarevar_zscore <- df_asd_rarevar %>% 
    
    # convert FDR to p
    mutate(p_fdr = fdr_to_p(FDR, bf_vec = log_BF, target_max_z = 10), .after = FDR) %>%
    
    # Calculate Z-score from p-value
    mutate(z_fdr = qnorm(p_fdr, lower.tail = FALSE), .after = p_fdr) %>% 
    arrange(-z_fdr)
```


# --- 3) Gene info: Get gene lengths and biotype --- #

## 3.1 Load gtf

```{r load_gtf}
# Load GRCh38 version 29
url_GRCh38.p14_GENCODE_v29 <- paste0("https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_29/gencode.v29.annotation.gtf.gz")
gtf_GRCh38.p14_GENCODE_v29 <- rtracklayer::import(url_GRCh38.p14_GENCODE_v29)

# Convert to tibble
df_GRCh38.p14_GENCODE_v29 <- gtf_GRCh38.p14_GENCODE_v29 %>% as_tibble()
rm(list="gtf_GRCh38.p14_GENCODE_v29") # clean up to save space
```

## 3.2 Calculate the length of each gene as the sum of all non-overlapping exon segments 

```{r union_cds_function}
calculate_union_cds <- function(object) {
    
    # If the input object is of GRanges class, convert to tibble
    object_class <- class(object)[1]
    if ( object_class == "GRanges" ) {
        df_genome <- object %>% as_tibble()
    } else if ( object_class == "tbl_df" ) {
        df_genome <- object
    } else { stop("Input object must be of class GRanges or tibble") }
    
    # Calculate the union CDS length for each gene by merging non-overlapping segments of all exons
    df_cds_length <- df_genome %>% 
        
        # Filter for exons only
        filter(type == "exon") %>%
        
        # Group by gene ID
        group_by(gene_id) %>% 
        dplyr::select(gene_id, exon_id, start, end) %>% 
        
        # Merging logic: new block if start/end don't overlap
        arrange(start) %>% # Arrange tibble in order of start site
        mutate(end_lag = lag(end, default = 0)) %>% # reference the previous exon's end coordinate
        mutate(is_new_block = start > end_lag + 1) %>% # marks the start of a non-overlapping region
        group_by(block_id = cumsum(is_new_block)) %>% # groups overlapping regions together
        
        # Summarize to find the merged block boundaries
        group_by(gene_id, block_id) %>% 
        summarise(
            block_start = min(start),
            block_end = max(end),
            .groups = "drop"
        ) %>% 
        
        # Calculate the block lengths and final sum
        group_by(gene_id) %>% 
        mutate(block_length = block_end - block_start + 1) %>%  # calculate the length of the merged block
        summarise(union_cds_length = sum(block_length))
    
    # Return the data frame
    return(df_cds_length)
    
}
```
```{r calc_cds}
# Calculate union CDS length as the sum of non-overlapping exon lengths
df_cds_length <- calculate_union_cds(df_GRCh38.p14_GENCODE_v29)

# Combine with biotype for full gene annotation
df_gene_info <- df_GRCh38.p14_GENCODE_v29 %>% 
    dplyr::select(gene_id, gene_name, gene_type) %>%
    left_join(df_cds_length) %>% 
    distinct() %>%
    
    # Rename columns for interpretability
    dplyr::rename(ensembl_gene_id_version = gene_id,
                  gene_biotype = gene_type,
                  gene_name_gtf = gene_name
    ) %>% 
    
    # Get the ensembl_gene_id without the version number
    mutate(ensembl_gene_id = str_remove(ensembl_gene_id_version, "\\..*$")) %>% 
    
    # Transform CDS length for models
    mutate(
        z_union_cds = scale(union_cds_length)[,1], # z-score union CDS
        log_union_cds = log10(union_cds_length + 1), # log10 transform union CDS
        z_log_union_cds = scale(log_union_cds)[,1] # z-score log10-transformed union CDS
    )
```


# --- 4) Combine & save --- #

## 4.1 Combine ASD rare variant statistics with gene info

```{r combine_tibbles}
df_asd_rarevar_zscore <- df_asd_rarevar_zscore %>% 
    left_join(df_gene_info, by = join_by(ensembl_gene_id))
```

```{r save}
saveRDS(df_asd_rarevar_zscore, file = file.path(notebook_out_dir, "00_01_ASD-rarevar-stats.RDS"))
```


# --- 5) Plots --- #

## 5.1 log10(BF) distribution

```{r plot_log10BF, fig.height = 2.5, fig.width = 3}
p_BF <- df_asd_rarevar_zscore %>% 
    ggplot(aes(x = log10(BF))) +
    geom_histogram(aes(y = after_stat(density)),
                   color = "#2c2c2c", alpha = 0.5) +
    geom_density(color = "purple") +
    labs(title = "log10(BF) distribution",
         x = "log10(BF)")
p_BF
# ggsave(
#     filename = file.path(figures_dir, "00_data_format", "00_01_log10BF_distribution.png"),
#     plot = p_BF,
#     height = 2.5,
#     width = 3,
#     dpi = 300
# )
```

## 5.2 FDR distribution

```{r plot_FDR, fig.height = 2.5, fig.width = 3}
p_FDR <- df_asd_rarevar_zscore %>% 
    ggplot(aes(x = FDR)) +
    geom_histogram(aes(y = after_stat(density)),
                   color = "#2c2c2c", alpha = 0.5) +
    geom_density(color = "purple") +
    labs(title = "FDR distribution",
         x = "FDR")
p_FDR
# ggsave(
#     filename = file.path(figures_dir, "00_data_format", "00_01_FDR_distribution.png"),
#     plot = p_FDR,
#     height = 2.5,
#     width = 3,
#     dpi = 300
# )
```

## 5.3 P-value distribution

```{r plot_p, fig.height = 2.5, fig.width = 3}
p_P <- df_asd_rarevar_zscore %>% 
    ggplot(aes(x = p_fdr)) +
    geom_histogram(aes(y = after_stat(density)),
                   color = "#2c2c2c", alpha = 0.5) +
    geom_density(color = "purple") +
    labs(title = "FDR-derived P-value distribution",
         x = "P-value")
p_P
# ggsave(
#     filename = file.path(figures_dir, "00_data_format", "00_01_P_distribution.png"),
#     plot = p_FDR,
#     height = 2.5,
#     width = 3,
#     dpi = 300
# )
```

## 5.4 FDR-derived Z-score distribution

```{r plot_z, fig.height = 2.5, fig.width = 3}
p_Z <- df_asd_rarevar_zscore %>% 
    ggplot(aes(x = z_fdr)) +
    geom_histogram(aes(y = after_stat(density)),
                   color = "#2c2c2c", alpha = 0.5) +
    geom_density(color = "purple") +
    labs(title = "FDR-derived Z-score distribution",
         x = "Z-score")
p_Z
# ggsave(
#     filename = file.path(figures_dir, "00_data_format", "00_01_Z_distribution.png"),
#     plot = p_FDR,
#     height = 2.5,
#     width = 3,
#     dpi = 300
# )
```

## 5.5 Combine & save

```{r combine_all, fig.height = 5, fig.width = 6}
p_stats <- (p_BF + p_FDR) / (p_P + p_Z)
p_stats
ggsave(
    filename = file.path(figures_dir, "supplement", "00_01_ASD_stats_distributions.png"),
    plot = p_stats,
    height = 5,
    width = 6,
    dpi = 300
)
```


