---
title: "Format network data"
description: "Format and combine networks for downstream enrichment analyses (includes bulk & single-cell coexpression networks & regulons)"
date: 2026-07-06
format: html
---

# --- 0) Setup --- #

## 0.0 Load setup & data paths

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r env}
# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# List available networks files
networks_files <- list.files(file.path(data_dir, "networks"))
networks_paths <- file.path(data_dir, "networks", networks_files)

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "00_data_format")
```

## 0.1 Helper functions

```{r fix_excel_date_genes}
# Recover gene symbols that Excel silently coerced into dates.
#
# Excel converts symbols like MARCH1, SEPT7, DEC1 into dates, which read back as
# numeric serials (e.g. 43894 -> 2020-03-04 -> "MARCH4"). Only these date-like
# gene families coerce, so the original symbol is always the pre-2020 HGNC name
# (MARCH#/SEPT#/DEC#). That nomenclature also matches the annotations these
# papers used (Gandal 2022: GENCODE release 25/lift37; Wamsley 2024: Ensembl 88),
# so recovered symbols align with the rest of each file and map cleanly to
# Ensembl gene IDs.
fix_excel_date_genes <- function(gene_vec) {
    suppressWarnings(serial <- as.numeric(gene_vec))
    is_serial <- !is.na(serial) & serial > 30000 & serial < 60000  # Excel date range (~1982–2064)
    if (!any(is_serial)) return(gene_vec)

    # Excel 1900 date system: day 0 = 1899-12-30 (accounts for the 1900 leap-year bug)
    d      <- as.Date(serial[is_serial], origin = "1899-12-30")
    day    <- as.integer(format(d, "%d"))
    family <- c(Mar = "MARCH", Sep = "SEPT", Dec = "DEC")[format(d, "%b")]

    # Rebuild the symbol (e.g. MARCH + 4 -> "MARCH4"); leave any non-gene date untouched
    recovered <- ifelse(is.na(family), gene_vec[is_serial], paste0(family, day))

    if (anyNA(family)) {
        warning("fix_excel_date_genes(): ", sum(is.na(family)),
                " date serial(s) did not decode to a MARCH/SEPT/DEC gene and were left unchanged: ",
                paste(unique(serial[is_serial][is.na(family)]), collapse = ", "))
    }

    gene_vec[is_serial] <- recovered
    gene_vec
}
```


# --- 1) Load and format network data --- #

## 1.1 Wen et al., 2024

Developmental bulk modules (gene & isoform)

```{r wen}
# Path to Wen data
wen_path <- networks_paths[str_detect(networks_paths, "Wen")]
wen_path <- wen_path[!str_detect(wen_path, "\\~\\$")] # in case the file is open

# Load and format Wen data
df_wen <- read_xlsx(path = wen_path, sheet = "ST6-1-Module-Genes") %>% 
    dplyr::rename("ensembl_gene_id" = "gene_id",
                  "ensembl_transcript_id" = "transcript_id") %>%
    
    # Set factor order for modules
    mutate(mod_numeric = str_remove(assigned_module, "M") %>% as.numeric) %>% 
    arrange(mod_numeric) %>% 
    mutate(module = factor(full_dendro_name, levels = unique(.$full_dendro_name)))

# Membership of genes and transcripts to each module
df_wen_modules <- df_wen %>% 
    dplyr::select(module, ensembl_gene_id, ensembl_transcript_id, gene_name) %>% 
    distinct()

print(
    paste0(
        "N = ", length(unique(df_wen_modules$ensembl_gene_id)), " genes mapped to ",
        length(unique(df_wen_modules$module)), " unique co-expression modules"
    )
)

# Module eigengene data (keeping separate for formatting purposes)
df_wen_kme <- df_wen %>% 
    dplyr::select(module, ensembl_gene_id, ensembl_transcript_id, gene_name, kME, kME_p)
```


## 1.2 Gandal et al., 2022

### 1.2.1 Gene-level

```{r gandal_genes}
# Path to Gandal data
gandal_path <- networks_paths[str_detect(networks_paths, "Gandal") & str_detect(networks_paths, "s5")] # Table S6 has enrichment information
gandal_path <- gandal_path[!str_detect(gandal_path, "\\~\\$")] # in case the file is open

# Load Gandal gene-level pancortical module data
df_gandal_genes <- read_xlsx(path = gandal_path, sheet = "Gene_Level") %>% 
    dplyr::rename("gene_name" = "external_gene_name") %>% 
    
    # Recover gene symbols that Excel coerced into dates (MARCH/SEPT/DEC families)
    mutate(gene_name = fix_excel_date_genes(gene_name)) %>%

    # Set module factor order
    mutate(
        module_numeric = str_remove(WGCNA_module, "_.*"),
        module_numeric = str_remove_all(module_numeric, "[^0-9]") %>% as.numeric, .before = 2
    ) %>% 
    arrange(module_numeric) %>% 
    mutate(WGCNA_module = paste0("Gene", WGCNA_module)) %>% 
    mutate(
        module = factor(WGCNA_module, levels = unique(.$WGCNA_module)),
        .before = 1
    ) %>% 
    dplyr::select(-c(module_numeric, WGCNA_module))

# Membership of genes to each module
df_gandal_gene_modules <- df_gandal_genes %>% 
    dplyr::select(module, ensembl_gene_id, gene_name) %>% 
    distinct()

print(
    paste0(
        "N = ", length(unique(df_gandal_gene_modules$ensembl_gene_id)), " genes mapped to ",
        length(unique(df_gandal_gene_modules$module)), " unique co-expression modules"
    )
)


# Module eigengene data (keeping separate for formatting purposes)
df_gandal_genes_kme <- df_gandal_genes %>% 
    dplyr::select(module, ensembl_gene_id, gene_name, contains("kME"))
```

### 1.2.2 Isoform-level

```{r gandal_isoforms}
# Load Gandal isoform-level pancortical module data
df_gandal_isoforms <- read_xlsx(path = gandal_path, sheet = "Isoform_Level") %>% 
    dplyr::rename("gene_name" = "external_gene_name") %>% 
    
    # Recover gene symbols that Excel coerced into dates (MARCH/SEPT families)
    mutate(gene_name = fix_excel_date_genes(gene_name)) %>%

    # Set module factor order
    mutate(
        module_numeric = str_remove(WGCNA_module, "_.*"),
        module_numeric = str_remove_all(module_numeric, "[^0-9]") %>% as.numeric, .before = 2
    ) %>% 
    arrange(module_numeric) %>% 
    mutate(WGCNA_module = paste0("Isoform", WGCNA_module)) %>% 
    mutate(
        module = factor(WGCNA_module, levels = unique(.$WGCNA_module)),
        .before = 1
    ) %>% 
    dplyr::select(-c(module_numeric, WGCNA_module))

# Membership of isoforms to each module
df_gandal_isoform_modules <- df_gandal_isoforms %>% 
    dplyr::select(module, ensembl_gene_id, ensembl_transcript_id, gene_name) %>% 
    distinct()

print(
    paste0(
        "N = ", length(unique(df_gandal_isoform_modules$ensembl_transcript_id)), " isoforms mapped to ",
        length(unique(df_gandal_isoform_modules$ensembl_gene_id)), " unique genes in ",
        length(unique(df_gandal_isoform_modules$module)), " unique co-expression modules"
    )
)

# Module eigengene data (keeping separate for formatting purposes)
df_gandal_isoforms_kme <- df_gandal_isoforms %>% 
    dplyr::select(module, ensembl_gene_id, ensembl_transcript_id, gene_name, contains("kME"))
```


## 1.3 Wang et al., 2025

Developmental single-cell regulons

```{r wang}
# Path to Wang data
wang_path <- networks_paths[str_detect(networks_paths, "Wang") & str_detect(networks_paths, "s13")]
wang_path <- wang_path[!str_detect(wang_path, "\\~\\$")] # in case the file is open

# Format Wang data to get 'module' (regulon) column and gene membership column
df_wang_regulons <- read_xlsx(path = wang_path, sheet = "Supplementary Table 13b") %>% 
    row_to_names(1) %>% 
    clean_names %>% 
    
    # create module and gene_name columns
    separate_rows(target_genes, sep = ",") %>%
    dplyr::rename(gene_name = target_genes, module = e_regulon) %>% 
    dplyr::select(module, gene_name) %>% 
    distinct()

print(paste0("N = ", length(unique(df_wang_regulons$gene_name)), " unique genes"))

# Query the annotation hub for the human EnsDb version 98 (GRCh38 v32, used in Wang et al)
hub <- AnnotationHub()
query(hub, c("EnsDb", "Hsapiens", "98")) # get the access ID
ebd98 <- hub[["AH75011"]] # query the db with that access ID

# Read in Wang background to get full gene search space (background from Yundan)
df_wang_background <- read_csv(paste0(data_dir, "Wang2025_background_genes.csv")) %>% 
    dplyr::select(gene) %>% 
    dplyr::rename(gene_name = gene) %>% 
    distinct() %>% # n = 32470
    
    # map to ensembl_gene_id
    mutate(
        ensembl_gene_id = ensembldb::mapIds(
            ebd98, # GRCh38 v32 (Ensembl 98)
            keys = gene_name,
            keytype = "SYMBOL",
            column = "GENEID",
            multiVals = "first"
        )
    ) # N = 9 genes unmapped

# Map any remaining gene symbols
unmapped_genes <- df_wang_background %>% filter(is.na(ensembl_gene_id)) %>% pull(gene_name) # N = 9 IDs did not map
df_wang_background <- df_wang_background %>% 
    mutate(ensembl_gene_id = ifelse(
        gene_name %in% unmapped_genes,
        ensembldb::mapIds(
            ebd98, # GRCh38 v32 (Ensembl 98)
            keys = gene_name %>% str_remove("\\.\\d+$"),
            keytype = "SYMBOL",
            column = "GENEID",
            multiVals = "first"
        ),
        ensembl_gene_id
    )
    ) %>% # all genes are now mapped
    distinct()

# Combine unmapped genes with regulon data
df_wang_regulons <- df_wang_regulons %>% 
    left_join(df_wang_background, by = join_by(gene_name)) %>% 
    
    # add all genes that are in the background but not mapped to regulons
    bind_rows(
        df_wang_background %>% 
            filter(!(gene_name %in% unique(df_wang_regulons$gene_name))) %>% 
            mutate(module = "none")
    )

print(
    paste0(
        "N = ", length(unique(df_wang_regulons$ensembl_gene_id)), " genes mapped to ",
        length(unique(df_wang_regulons$module)), " unique regulons"
    )
)
```


## 1.4 Wamsley et al., 2024

ASD single-cell regulons

```{r wamsley}
# Path to Wamsley data
wamsley_path <- networks_paths[str_detect(networks_paths, "Wamsley")]
wamsley_path <- wamsley_path[!str_detect(wamsley_path, "\\~\\$")] # in case the file is open

# Regulon membership
df_wamsley_regulons <- read_xlsx(path = wamsley_path, 
                                 sheet = "(A)Regulons", 
                                 skip = 2) %>% 
    pivot_longer(1:ncol(.), names_to = "module", values_to = "gene_name") %>%

    # Recover gene symbols that Excel coerced into dates (MARCH/SEPT families), before mapping to Ensembl IDs
    mutate(gene_name = fix_excel_date_genes(gene_name)) %>%

    na.omit %>%
    arrange(module)


# Query the annotation hub for the human EnsDb version 98 (GRCh38 v26; least number of genes dropped this way)
hub <- AnnotationHub()
query(hub, c("EnsDb", "Hsapiens", "88")) # get the access ID
ebd88 <- hub[["AH53715"]] # query the db with that access ID

# Add ensembl_gene_id column
df_wamsley_regulons <- df_wamsley_regulons %>% 
    mutate(
        ensembl_gene_id = ensembldb::mapIds(
            ebd88, # GRCh38 v26 (Ensembl 88)
            keys = gene_name,
            keytype = "SYMBOL",
            column = "GENEID",
            multiVals = "first"
        )
    ) %>% 
    
    # map genes that are still unmapped (only CBWD6) using newer version
    mutate(ensembl_gene_id = ifelse(is.na(ensembl_gene_id), 
                                    ensembldb::mapIds(ebd98, 
                                                      keys = gene_name, 
                                                      keytype = "SYMBOL", 
                                                      column = "GENEID", 
                                                      multiVals = "first"), 
                                    ensembl_gene_id)
    )


print(
    paste0(
        "N = ", length(unique(df_wamsley_regulons$ensembl_gene_id)), " genes mapped to ",
        length(unique(df_wamsley_regulons$module)), " unique regulons"
    )
)
```



# --- 2) Combine networks, assign labels, and save --- #

## 2.1 Combine gene networks across studies

```{r combine_networks}
# Combine all gene networks into one long tibble (nested below, after labeling)
df_network_data_long <- df_wen_modules %>%
    mutate(study = "Wen", .before = 1) %>%

    bind_rows(
        df_gandal_gene_modules %>%
            mutate(study = "GandalGenes", .before = 1)
    ) %>%

    bind_rows(
        df_gandal_isoform_modules %>%
            mutate(study = "GandalIsoforms", .before = 1)
    ) %>%

    bind_rows(
        df_wang_regulons %>%
            mutate(study = "Wang", .before = 1)
    ) %>%

    bind_rows(
        df_wamsley_regulons %>%
            mutate(study = "Wamsley", .before = 1)
    )
```

## 2.2 Module eigengene (kME) data

```{r format_kme}
# Format module eigengene data
df_module_kme <- df_wen_kme %>%
    mutate(study = "Wen", .before = 1) %>%
    mutate(kME_module = module, .before = kME) %>%

    bind_rows(
        df_gandal_genes_kme %>%
            filter(module != "GeneM0_grey") %>%
            mutate(study = "GandalGenes", .before = 1) %>%
            pivot_longer(contains("kME"), names_to = "kME_module", values_to = "kME") %>%
            mutate(kME_module = str_replace(kME_module, "kME", "GeneM"))
    ) %>%

    bind_rows(
        df_gandal_isoforms_kme %>%
            filter(module != "IsoformM0_grey") %>%
            mutate(study = "GandalIsoforms", .before = 1) %>%
            pivot_longer(contains("kME"), names_to = "kME_module", values_to = "kME") %>%
            mutate(kME_module = str_replace(kME_module, "kME", "IsoformM"))
    )
```

## 2.3 Module display labels

```{r module_labels}
# Hub gene (highest self-kME) for each co-expression module. Co-expression
# studies store each module's self-kME where module == kME_module; regulons have
# no kME and are labeled by their TF name + activation (+) / repression (-) sign.
hub_gene_map <- df_module_kme %>%
    filter(module == kME_module) %>%
    group_by(module) %>%
    slice_max(kME, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module, hub_gene = gene_name)

# module_label: hub gene for co-expression modules; "TF (+/-)" for regulons
df_module_labels <- df_network_data_long %>%
    distinct(module) %>%
    left_join(hub_gene_map, by = "module") %>%
    mutate(
        is_regulon   = is.na(hub_gene),
        regulon_gene = str_extract(module, "^[^_\\-\\+]+"),
        regulon_sign = str_extract(str_remove(module, "^[^_\\-\\+]+"), "[\\+\\-]"),
        module_label = if_else(is_regulon, paste0(regulon_gene, " (", regulon_sign, ")"), hub_gene)
    ) %>%
    # one co-expression hub gene reads as a date-coerced number; restore symbol
    mutate(module_label = if_else(module_label == "43894", "MARCHF4", module_label)) %>%
    dplyr::select(module, module_label)
```

## 2.4 Attach labels, nest, and save

```{r save_networks}
# Attach module_label to each gene, then nest by study
df_network_data <- df_network_data_long %>%
    left_join(df_module_labels, by = "module") %>%
    dplyr::select(study, module, module_label, everything()) %>% 
    group_by(study) %>%
    nest()

# Save network, kME, and module-label objects for downstream analyses
saveRDS(df_network_data,  file = file.path(notebook_out_dir, "00_00_gene-networks-all.RDS"))
saveRDS(df_module_kme,    file = file.path(notebook_out_dir, "00_00_co-expression-kME.RDS"))
```
