---
title: "Fig2"
format: html
---

# --- 0) Setup ---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- paste0(here::here(), "/")
source(paste0(project_dir, "setup.R"))
```

## 0.0 Load data

```{r load_data}
# Module eigengene data
df_module_kme <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_co-expression-kME.RDS"))

# ASD module association statistics
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "02_module_identification", "02_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results %>% 
    mutate(module = factor(module, levels = unique(.$module)))

# GO topic model results (spectral fit)
topic_model <- readRDS(file.path(analysis_dir, "03_module_enrichments", "03_00_topic-model-K3.RDS"))
df_gamma <- readRDS(file.path(analysis_dir, "03_module_enrichments", "03_00_topic-module-gamma.RDS"))
df_beta <- readRDS(file.path(analysis_dir, "03_module_enrichments", "03_00_topic-word-beta.RDS"))

# Restart stability of the topic model
topic_stability <- readRDS(
    file.path(analysis_dir, "03_module_enrichments", "03_00_topic-restart-stability.RDS")
)
df_assignment_prop <- topic_stability$assignment_prop

# AUCell results
aucell_high_prop_pct <- read.csv(
    file.path(analysis_dir, "03_module_enrichments", "03_01_AUCell", "high_proportion_pct.csv"),
    row.names = 1, check.names = FALSE
)

# Gene-topic assignments
df_gene_assignments <- read_csv(file.path(analysis_dir, "03_module_enrichments", "03_02_gene-topic-assignment.csv"))

# GR subclusters
GR_net <- readRDS(file.path(analysis_dir, "03_module_enrichments", "03_03_GR-GO-similarity-graph.RDS"))
df_GR_subclusters <- read_csv(file.path(analysis_dir, "03_module_enrichments", "03_03_GR-subcluster-by-GO-similarity.csv"),
                        show_col_types = FALSE)

# BrainSpan trajectories
df_traj_topic <- readRDS(file.path(analysis_dir, "03_module_enrichments", "03_04_brainspan-ASD-gene-trajectories.RDS"))

# BrainSpan prenatal t-stats
tstat_data <- readRDS(file.path(analysis_dir, "03_module_enrichments", "03_04_brainspan-prenatal-tstats.RDS"))
df_tstats <- tstat_data$gene_tstats
df_tstat_comparison <- tstat_data$topic_vs_background
```

## 0.1 Prep: order modules & cell types

```{r prep_main_fig}
# Consensus topic per module (the topic it is assigned to in most restarts)
df_top_topics <- df_assignment_prop %>%
    group_by(module) %>%
    slice_max(assignment_prop, n = 1, with_ties = FALSE) %>%
    select(module, top_topic = topic) %>%
    ungroup()

# Initial topic-based ordering (used as the within-topic starting point for clustering)
module_order_topic <- df_assignment_prop %>%
    select(module, topic, assignment_prop) %>%
    pivot_wider(names_from = topic, values_from = assignment_prop) %>%
    left_join(df_top_topics, by = "module") %>%
    arrange(
        top_topic,
        desc(case_when(
            top_topic == "genomic regulation / \nRNA processing" ~
                `genomic regulation / \nRNA processing`,
            top_topic == "synaptic transmission / \nion transport" ~
                `synaptic transmission / \nion transport`,
            top_topic == "neuron development / \nmorphogenesis" ~
                `neuron development / \nmorphogenesis`
        ))
    ) %>%
    pull(module)

# Hierarchically cluster modules within each topic group based on AUCell cell type enrichment
cell_order_aucell <- c(
    "RG-vRG", "RG-tRG", "RG-oRG",
    "IPC-EN", "EN-Newborn", "EN-IT-Immature",
    "EN-L2_3-IT", "EN-L4-IT", "EN-L5-IT", "EN-L6-IT",
    "Cajal-Retzius cell", "EN-Non-IT-Immature", "EN-L5-ET", "EN-L5_6-NP", "EN-L6-CT", "EN-L6b",
    "IPC-Glia",
    "IN-dLGE-Immature", "IN-CGE-Immature", "IN-MGE-Immature",
    "IN-CGE-VIP", "IN-CGE-SNCG", "IN-Mix-LAMP5",
    "IN-MGE-SST", "IN-MGE-PV",
    "Astrocyte-Immature", "Astrocyte-Protoplasmic", "Astrocyte-Fibrous",
    "OPC", "Oligodendrocyte-Immature", "Oligodendrocyte",
    "Microglia", "Vascular", "Unknown"
)

# Subset cell types that were included in AUCell
cell_types_for_clust <- cell_order_aucell[
    cell_order_aucell %in% colnames(aucell_high_prop_pct) &
    !cell_order_aucell %in% c("Unknown", "Vascular")
]

# Cluster each topic separately
mat <- aucell_high_prop_pct[, cell_types_for_clust]
mat[is.na(mat)] <- 0

topic_order_seq <- unique(df_top_topics$top_topic[match(module_order_topic, df_top_topics$module)])

module_order <- purrr::map(topic_order_seq, function(topic) {
    mods <- module_order_topic[module_order_topic %in% filter(df_top_topics, top_topic == topic)$module]
    if (length(mods) == 1) return(mods)
    sub_mat <- mat[mods, , drop = FALSE]
    hc <- hclust(dist(sub_mat), method = "complete")
    mods_ordered <- mods[hc$order]

    # MORPH is the middle facet, so reversing it places the MORPH modules carrying the
    # most SYN signal (FOXP2, ZBTB18) at the bottom edge, adjacent to the SYN block
    if (topic == "neuron development / \nmorphogenesis") mods_ordered <- rev(mods_ordered)

    # GR reversed to align developmental modules
    if (topic == "genomic regulation / \nRNA processing") mods_ordered <- rev(mods_ordered)
    
    # SYN: these five modules are assigned with equal confidence and cluster too tightly
    # for the dendrogram to give a meaningful order, so it is set by hand
    # (listed bottom -> top as they appear in the figure)
    if (topic == "synaptic transmission / \nion transport") {
        mods_ordered <- c(
            "M84-gene",          # SRCIN1
            "IsoformM3_brown",  # PRKCE
            "IsoformM4_yellow", # MARCH4
            "M88-xx",           # FBXO41
            "MEF2C_+_+"       # MEF2C (+)
        )
        # guard against a module silently dropping out of the figure if the SYN set changes
        stopifnot(setequal(mods_ordered, mods))
    }

    mods_ordered
}) %>% unlist()
```


# --- 1) Panel A: Module results colored by GO theme ---

```{r module_significance, fig.height = 3, fig.width = 2.34}
p_mods_sig <- df_stepwise_res %>%
    
    # add GO topic information (consensus assignment, as used in panel B)
    left_join(
        df_assignment_prop %>%
            group_by(module) %>%
            slice_max(assignment_prop, n = 1, with_ties = FALSE) %>%
            ungroup() %>%
            dplyr::select(module, topic_label),
        by = join_by(module)
    ) %>%
    
    # add regulon sign to point
    separate(module_label, into = c("label", "regulon_sign"), sep = "\\(") %>% 
    mutate(regulon_sign = str_remove(regulon_sign, "\\)") %>% str_trim,
           label = str_trim(label)) %>% 

    # Layout
    ggplot(aes(x = beta_std, y = -log10(p_adj))) +
    #geom_segment(aes(x = conf_low, xend = conf_high)) + # turn on for error bars
    geom_point(aes(fill = topic_label, color = hub_is_risk_gene, size = n, shape = is.na(regulon_sign))) +
    geom_text(aes(label = regulon_sign),
              size = 2.5, color = "white", fontface = "bold", vjust = 0.5) +
    ggrepel::geom_text_repel(
        aes(label = label, fontface = ifelse(hub_is_risk_gene, "bold", "plain")),
        size = 1.75,
        min.segment.length = 0,
        max.overlaps = Inf,
        force = 1,
        box.padding = 0.25,
        point.padding = 0.5,
        segment.color = "grey50",
        color = "black",
        show.legend = FALSE,
        inherit.aes = TRUE
    ) +
    
    # Aesthetics
    scale_color_manual(values = c("TRUE" = "black", "FALSE" = "gray90")) +
    scale_fill_manual(values = topic_colors, guide = "none") +
    scale_shape_manual(values = c("FALSE" = 21, "TRUE" = 22),
                       labels = c(
                           "FALSE"   = "regulon",
                           "TRUE" = "co-expression module"
                       )
    ) +
    #scale_fill_paletteer_d("ggsci::default_locuszoom") +
    scale_size_continuous(range = c(2, 5)) +
    scale_x_log10() +
    guides(#fill = guide_legend(title = NULL, label.hjust = 1),
           size = guide_legend(title = "Gene set size", label.hjust = 1, ncol = 1,
                               override.aes = aes(shape = 21, fill = "white")),
           color = guide_legend(title = "Hub / TF is risk gene", label.hjust = 1,
                                override.aes = aes(shape = 21, fill = "white")),
           shape = guide_legend(title = NULL)
    ) +
    #facet_wrap(vars(network_list), scales = "free") +
    coord_cartesian(clip = "off") +
    labs(
        x = "Standardized effect size",
        y = "-log10(FDR)",
        title = "Modules enriched for ASD risk genes"
    ) +
    theme(
        plot.title = element_text(size = 7),
        axis.text = element_text(size = 5.5),
        axis.title.y = element_text(size = 6),
        axis.title.x = element_text(size = 6),#element_text(size = 6, margin = margin(t = -57)),
        legend.text = element_text(size = 5.5, margin = margin(l = 1)),
        legend.title = element_text(size = 6),
        legend.key.size = unit(0.3, "cm"),
        legend.background = element_rect(color = "transparent", linewidth = 0.25),
        legend.position = "bottom", #c(0.5, -0.25),
        legend.box = "horizontal",
        legend.direction = "vertical",
        legend.box.just = "top",
        legend.title.position = "top",
        legend.box.margin = margin(t = -7),
        legend.spacing = unit(0.15, "cm"),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.25, color = "gray40")
    )

p_mods_sig
```

# --- 2) Panel B: Topic modeling results ---

## i Word-topic probability (beta from spectral fit)

```{r topic_beta, fig.height = 3, fig.width = 1}
p_beta <- df_beta %>% 
    mutate(topic_label = case_when(
        topic == 1 ~ "GR",
        topic == 2 ~ "SYN",
        topic == 3 ~ "MORPH"
    )) %>% 
    #group_by(topic) %>% 
    #slice_max(beta, n = 20) %>% # Keep top words per topic
    #ungroup %>% 
    filter(p_adj < 0.05) %>% 
    filter(!(topic_label == "SYN" & term == "development")) %>% 

    ggplot(aes(x = beta, y = reorder_within(term, beta, topic))) +
    geom_col(aes(fill = topic_label, alpha = z_score), color = "transparent") +
    facet_wrap(vars(topic_label), scales = "free_y", space = "free_y", ncol = 1) +
    
    scale_alpha_continuous(limits = c(0, 15.3), range = c(0, 0.8), name = "Z-score") +
    scale_fill_manual(values = topic_colors, guide = "none") +
    scale_color_manual(values = c("TRUE" = "black", "FALSE" = "gray"), name = "FDR < 0.05") +
    scale_y_reordered() +
    scale_x_continuous(n.breaks = 3) +
    guides(color = guide_legend(override.aes = list(fill = "white"))) +
    labs(x = "Beta \n(word-topic probability)", y = NULL,
         title = "Topic words" #"Highest probability per topic"
         ) +
    coord_cartesian(clip = "off") +
    theme(
        plot.title = element_text(size = 7),
        legend.position = "none",
        panel.grid = element_blank(),
        strip.background = element_blank(),
        #strip.text = element_text(face = "bold", size = 6),
        strip.text = element_blank(),
        strip.clip = "off",
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 5, hjust = 0, color = "black", margin = margin(r = -50)),
        axis.ticks.y = element_blank(),
        axis.title.x = element_text(size = 6),
        plot.margin = margin(t = 0, b = 0, l = 10, r = 0),
        panel.spacing = unit(0.2, "lines"),
        panel.border = element_blank()
    )
p_beta
```

## ii Module-topic assignment (# of assignments across topic model restarts)

```{r module_GO, fig.height = 4.12, fig.width = 1.83}
# Bold the y-axis label for modules whose hub/TF is itself an ASD risk gene
y_label_df <- df_stepwise_res %>%
    distinct(module_label, hub_is_risk_gene) %>%
    mutate(module_label = as.character(module_label),
           md = ifelse(hub_is_risk_gene, paste0("<b>", module_label, "</b>"), module_label))
y_label_lookup <- setNames(y_label_df$md, y_label_df$module_label)

# Plot
p_module_topic <- df_assignment_prop %>%
    left_join(df_top_topics) %>%
    mutate(module = factor(module, levels = module_order)) %>%
    arrange(module) %>%
    left_join(df_stepwise_res %>% dplyr::select(module, module_label)) %>%
    mutate(module_label = factor(module_label, levels = unique(.$module_label))) %>%

    ggplot(aes(x = topic_label, y = module_label)) +
    #geom_hline(yintercept = seq(1.5, nrow(df_sig_mods) - 0.5, 1), linewidth = 0.25, color = "gray") +
    geom_tile(aes(fill = topic_label, alpha = assignment_prop), color = NA) +
    facet_grid(rows = vars(top_topic), scales = "free_y", space = "free_y") +

    scale_x_discrete(
        labels = c(
            "GR"   = "<span style='color:#ea801c'><b>GR</b></span>",
            "MORPH" = "<span style='color:#0d7d87'><b>MORPH</b></span>",
            "SYN"   = "<span style='color:#4a2377'><b>SYN</b></span>"
        )
    ) +
    scale_y_discrete(labels = y_label_lookup) +
    scale_fill_manual(
        values = topic_colors,
        labels = c(
            "GR"   = "genomic regulation",
            "MORPH" = "neuron development",
            "SYN"   = "synaptic transmission"
        ),
        guide = "none"
    ) +
    scale_alpha_continuous(range = c(0, 1), limits = c(0, 1),
                           name = "assignment reproducibility") +
    
    coord_cartesian(clip = "off") +
    labs(x = NULL, y = NULL,
         title = "GO topic\nassignments") +
    theme(
        plot.title = element_text(size = 7),
        legend.key.width = unit(0.2, "cm"),
        legend.key.height = unit(0.2, "cm"),
        legend.position = "bottom",
        legend.title.position = "top",
        legend.text = element_text(size = 5, margin = margin(l = 0)),
        legend.title = element_text(size = 5.5),
        panel.grid = element_blank(),
        panel.border = element_rect(linewidth = 0.25),
        strip.text = element_blank(),
        panel.spacing = unit(0.0, "lines"),
        axis.text.x.bottom = ggtext::element_markdown(angle = 45, hjust = 0.9, size = 6.5),
        axis.text.y = ggtext::element_markdown(size = 5, color = "black"),
        plot.margin = margin(l = 0)
    )
p_module_topic
```

# --- 3) Panel C: AUCell enrichments ---

```{r plot_aucell_res, fig.height = 4.12, fig.width = 2.92}
df_cell_type_order_aucell <- tibble(cell_type = cell_order_aucell) %>%
    mutate(cell_order_rank = row_number(), .before = 1) %>% 
    arrange(cell_order_rank) %>%
    mutate(
        broad_group = case_when(
            cell_type %in% c("RG-vRG", "RG-tRG", "RG-oRG", "IPC-EN", "EN-Newborn",
                             "EN-IT-Immature", "EN-L2_3-IT", "EN-L4-IT", "EN-L5-IT", "EN-L6-IT",
                             "EN-Non-IT-Immature", "EN-L5-ET", "EN-L5_6-NP",
                             "EN-L6-CT", "EN-L6b", "Cajal-Retzius cell") ~ "EN",
            cell_type %in% c("IPC-Glia", "IN-dLGE-Immature", "IN-CGE-Immature",
                             "IN-MGE-Immature", "IN-CGE-VIP", "IN-CGE-SNCG",
                             "IN-Mix-LAMP5", "IN-MGE-SST", "IN-MGE-PV")  ~ "IN",
            cell_type %in% c("Astrocyte-Immature", "Astrocyte-Protoplasmic",
                             "Astrocyte-Fibrous", "OPC", "Oligodendrocyte-Immature",
                             "Oligodendrocyte", "Microglia")              ~ "Glia"
        ),
        sub_group = case_when(
            cell_type %in% c("RG-vRG", "RG-tRG", "RG-oRG",
                             "IPC-EN", "EN-Newborn", "Cajal-Retzius cell") ~ "Newborn",
            cell_type %in% c("EN-IT-Immature", "EN-L2_3-IT", "EN-L4-IT",
                             "EN-L5-IT", "EN-L6-IT")                       ~ "IT",
            cell_type %in% c("EN-Non-IT-Immature", "EN-L5-ET",
                             "EN-L5_6-NP", "EN-L6-CT", "EN-L6b")          ~ "non-IT",
            cell_type %in% c("IPC-Glia", "IN-dLGE-Immature",
                             "IN-CGE-Immature", "IN-MGE-Immature")         ~ "Immature",
            cell_type %in% c("IN-CGE-VIP", "IN-CGE-SNCG",
                             "IN-Mix-LAMP5", "IN-MGE-SST", "IN-MGE-PV")   ~ "Mature",
            cell_type %in% c("Astrocyte-Immature", "Astrocyte-Protoplasmic",
                             "Astrocyte-Fibrous")                           ~ "Astro",
            cell_type %in% c("OPC", "Oligodendrocyte-Immature",
                             "Oligodendrocyte")                             ~ "Oligo",
            cell_type == "Microglia"                                        ~ "Micro"
        ),
        abbrev_name = case_when(
            cell_type == "RG-vRG"                ~ "vRG",
            cell_type == "RG-tRG"                ~ "tRG",
            cell_type == "RG-oRG"                ~ "oRG",
            cell_type == "IPC-EN"                ~ "IPC-E",
            cell_type == "EN-Newborn"            ~ "Newborn",
            cell_type == "EN-IT-Immature"        ~ "IT-Immature",
            cell_type == "EN-L2_3-IT"            ~ "L2/3",
            cell_type == "EN-L4-IT"              ~ "L4",
            cell_type == "EN-L5-IT"              ~ "L5",
            cell_type == "EN-L6-IT"              ~ "L6",
            cell_type == "EN-Non-IT-Immature"    ~ "nIT-Immature",
            cell_type == "EN-L5-ET"              ~ "L5-ET",
            cell_type == "EN-L5_6-NP"            ~ "L5/6-NP",
            cell_type == "EN-L6-CT"              ~ "L6-CT",
            cell_type == "EN-L6b"                ~ "L6b",
            cell_type == "IPC-Glia"              ~ "Tri-IPC",
            cell_type == "IN-dLGE-Immature"      ~ "dLGE",
            cell_type == "IN-CGE-Immature"       ~ "CGE",
            cell_type == "IN-MGE-Immature"       ~ "MGE",
            cell_type == "IN-CGE-VIP"            ~ "VIP",
            cell_type == "IN-CGE-SNCG"           ~ "SNCG",
            cell_type == "IN-Mix-LAMP5"          ~ "LAMP5",
            cell_type == "IN-MGE-SST"            ~ "SST",
            cell_type == "IN-MGE-PV"             ~ "PV",
            cell_type == "Astrocyte-Immature"    ~ "Astro-Immature",
            cell_type == "Astrocyte-Protoplasmic" ~ "Protoplasmic",
            cell_type == "Astrocyte-Fibrous"     ~ "Fibrous",
            cell_type == "OPC"                   ~ "OPC",
            cell_type == "Oligodendrocyte-Immature" ~ "Oligo-Immature",
            cell_type == "Oligodendrocyte"       ~ "Oligo",
            cell_type == "Cajal-Retzius cell"    ~ "Cajal-Retzius",
            cell_type == "Microglia"             ~ "Micro"
        )
    )

abbrev_name_order <- df_cell_type_order_aucell %>%
    filter(!cell_type %in% c("Unknown", "Vascular")) %>%
    pull(abbrev_name)

# Pre-compute per-tile hex color: white → topic color scaled by pct_high
topic_ramps <- map(topic_colors, ~ colorRamp(c("white", .x)))

df_aucell <- aucell_high_prop_pct %>%
    as.data.frame() %>%
    tibble::rownames_to_column("module") %>%
    pivot_longer(-module, names_to = "cell_type", values_to = "pct_high") %>%
    left_join(df_cell_type_order_aucell, by = "cell_type") %>%
    left_join(df_assignment_prop %>%
                  group_by(module) %>%
                  slice_max(assignment_prop, n = 1, with_ties = FALSE) %>%
                  ungroup() %>%
                  dplyr::select(module, top_topic = topic_label),
              by = "module") %>%
    filter(!cell_type %in% c("Unknown", "Vascular"), !is.na(broad_group)) %>%
    mutate(
        fill_color = {
            pct_norm <- pmin(pmax(pct_high / 100, 0), 1)
            pct_norm[is.na(pct_norm)] <- 0
            map2_chr(pct_norm, top_topic, function(p, t) {
                rgb_vals <- topic_ramps[[t]](p)
                rgb(rgb_vals[1], rgb_vals[2], rgb_vals[3], maxColorValue = 255)
            })
        },
        abbrev_name = factor(abbrev_name, levels = abbrev_name_order),
        broad_group = factor(broad_group, levels = c("EN", "IN", "Glia")),
        sub_group   = factor(sub_group,   levels = c("Newborn", "IT", "non-IT",
                                                     "Immature", "Mature",
                                                     "Astro", "Oligo", "Micro")),
        module      = factor(module,      levels = module_order)
    )

p_cell_type <- df_aucell %>%
    ggplot(aes(x = abbrev_name, y = module)) +
    geom_tile(aes(fill = fill_color)) +
    geom_point(aes(color = pct_high), shape = NA) +  # invisible; drives legend only
    ggh4x::facet_nested(
        rows = vars(top_topic),
        cols = vars(broad_group, sub_group),
        scales = "free", space = "free",
        nest_line = element_line(linewidth = 0.3, color = "gray40"),
        strip = ggh4x::strip_nested(
            text_x  = element_text(size = 6),
            background_x = list(element_rect(fill = NA, color = NA),
                                element_rect(fill = NA, color = NA))
        )
    ) +
    scale_fill_identity(na.value = "grey90") +
    scale_color_gradientn(
        colors   = c("white", "black"),
        name     = "Proportion highly\nenriched (%)",
        na.value = "grey90"
    ) +
    guides(color = guide_colorbar(
        barwidth  = unit(1.5, "cm"),
        barheight = unit(0.15, "cm"),
        title.position = "top"
    )) +
    labs(x = NULL, y = NULL,
         title = "AUCell module enrichment by cell type") +
    theme(
        axis.text.x      = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 5.5),
        axis.text.y      = element_blank(),
        axis.ticks.y     = element_blank(),
        panel.grid       = element_line(color = "white"),
        panel.spacing    = unit(0.0, "lines"),
        panel.border     = element_rect(linewidth = 0.25),
        strip.background = element_blank(),
        strip.clip       = "off",
        strip.text.y     = element_blank(),
        plot.margin      = margin(r = 0),
        plot.title       = element_text(size = 7),
        legend.key.width  = unit(0.5, "cm"),
        legend.key.height = unit(0.15, "cm"),
        legend.position   = "bottom",
        legend.title.position = "top",
        legend.title      = element_text(size = 6, hjust = 0.5),
        legend.text       = element_text(size = 5.5)
    )

p_cell_type
```

# --- 4) Panel D: Developmental trajectory curves --- #

```{r plot_brainspan_curves, fig.width = 3, fig.height = 2.5}
# Set x-breaks based on Kang 2011
stage_x_breaks <- df_traj_topic %>%
    mutate(dev_stage = factor(dev_stage, levels = stage_levels)) %>%
    group_by(dev_stage) %>%
    summarise(x = median(log10_pcw), .groups = "drop") %>%
    arrange(dev_stage)

# Plot
p_brainspan_traj <- df_traj_topic %>% 
    ggplot(aes(x = log10_pcw, y = mean_z, color = GO_topic)) +
    
    # prenatal rectangle background
    annotate("rect",
             xmin = -Inf, xmax = log10(40),
             ymin = -Inf, ymax = Inf,
             fill = "grey92", alpha = 0.6) +
    geom_vline(xintercept = log10(40), linetype = "dashed",
               color = "grey60", linewidth = 0.3) +
    geom_hline(yintercept = 0, linetype = "dashed",
               color = "grey60", linewidth = 0.3) +

    # sample expression means
    geom_point(size = 0.05, alpha = 0.2) +
    
    # loess curves
    geom_smooth(
        aes(fill = GO_topic),
        method = "loess", span = 0.4,
        linewidth = 0.8, se = TRUE, alpha = 0.15, color = NA,
        show.legend = FALSE
    ) +
    geom_smooth(method = "loess", span = 0.4, linewidth = 0.4) +

    # aesthetics
    scale_color_manual(values = topic_colors, name = NULL) +
    scale_fill_manual(values  = topic_colors, guide = "none") +
    scale_x_continuous(
        breaks = stage_x_breaks$x,
        labels = as.character(stage_x_breaks$dev_stage) %>%
            str_replace("Neonatal and early infancy", "Neonatal and\nearly infancy") %>%
            str_replace("Middle and late childhood",  "Middle and\nlate childhood"),
        expand = expansion(mult = 0.02)
    ) +
    labs(x = NULL, y = "Within-timepoint Z-score", 
         title = "Developmental expression trajectories\nof genes by GO topic") +
    theme(
        plot.title         = element_text(size = 7, hjust = 0.5),
        axis.title.y       = element_text(size = 7),
        axis.text.x        = element_text(size = 6, angle = 90, hjust = 1, vjust = 0.5, color = "black"),
        axis.text.y = element_text(color = "black"),
        legend.position    = "none",
        panel.border = element_rect(color = "gray50"),
        panel.grid.major.x = element_blank()
    )

p_brainspan_traj
```

# --- 5) Panel E: Prenatal t-statistic density for topic gene lists

```{r plot_tstat_distributions, fig.width = 2, fig.height = 2.5}
# Per-topic annotation: median t-stat (distribution midpoint), the density height at that
# median, and the FDR (topic vs background Wilcoxon) → drawn as a coloured line from just
# above each distribution's midpoint to 0, labelled with significance stars.
df_topic_sig <- df_tstats %>%
    filter(GO_topic != "Background") %>%
    group_by(GO_topic) %>%
    summarise(
        median_tstat = median(prenatal_tstat, na.rm = TRUE),
        y_med = { d <- density(prenatal_tstat, na.rm = TRUE)
                  approx(d$x, d$y, xout = median(prenatal_tstat, na.rm = TRUE))$y },
        .groups = "drop"
    ) %>%
    left_join(df_tstat_comparison %>% select(GO_topic, p_adj_wilcox), by = "GO_topic") %>%
    mutate(
        y_line = y_med * 1.06,   # slightly above the distribution's midpoint (raise factor to lift)
        star = case_when(
            p_adj_wilcox < 0.001 ~ "***",
            p_adj_wilcox < 0.01  ~ "**",
            p_adj_wilcox < 0.05  ~ "*",
            TRUE                 ~ ""
        )
    )

# Plot
p_tstat_distr <- df_tstats %>%
    ggplot() +
    geom_density(
        data = df_tstats %>% filter(GO_topic == "Background"),
        aes(x = prenatal_tstat, fill = "Background"),
        color = "grey50", linetype = "dashed", linewidth = 0.15
    ) +
    geom_density(
        data = df_tstats %>% filter(GO_topic != "Background"),
        aes(x = prenatal_tstat, color = GO_topic),
        linewidth = 0.7
    ) +
    geom_vline(xintercept = 0, linetype = "dashed",
               color = "grey50", linewidth = 0.3) +
    # each topic's shift from 0: coloured line from just above its midpoint to 0, with
    # significance stars (topic vs background Wilcoxon FDR)
    geom_segment(
        data = df_topic_sig,
        aes(x = median_tstat, xend = 0, y = y_line, yend = y_line, color = GO_topic),
        linewidth = 0.4, show.legend = FALSE
    ) +
    geom_text(
        data = df_topic_sig,
        aes(x = median_tstat / 2, y = y_line, label = star, color = GO_topic),
        vjust = -0.10, size = 1.5, show.legend = FALSE
    ) +
    scale_fill_manual(values = c("Background" = "grey88"), name = NULL) +
    scale_color_manual(values = topic_colors, name = NULL, guide = "none") +
    guides(
        fill  = guide_legend(ncol = 1, order = 2,
                             override.aes = list(linetype = "dashed", color = "grey50", linewidth = 0.3))
    ) +
    labs(x = "t-statistic\n(pre- vs post-natal)", y = "Density", 
         title = "Prenatal bias of GO topic genes") +
    theme(
        plot.title        = element_text(size = 7, hjust = 0.5),
        axis.title       = element_text(size = 7),
        legend.position   = "bottom",
        legend.direction  = "vertical",
        legend.box        = "vertical",
        legend.key.size   = unit(0.3, "cm"),
        legend.spacing.x  = unit(0.05, "cm"),
        legend.spacing.y  = unit(0.35, "cm"),
        legend.margin     = margin(t = -3),
        legend.background = element_blank()
    )

p_tstat_distr
```

# --- 6) Panel F: Cortical enrichment patterns of GO topics [[PYTHON NOTEBOOK]] ---


# --- 7) Panel G: GR subclusters ---

```{r GR_subclusters, fig.height = 85/25.4, fig.width = 180/25.4}
GR_color <- topic_colors[["GR"]]

# --- Data --- #

# Per-gene subcluster assignment (Misc / cluster 8 is not plotted)
cluster_vec <- df_GR_subclusters %>%
    separate_rows(genes, sep = ", ") %>%
    dplyr::select(genes, GR_cluster) %>%
    distinct() %>%
    deframe

# Node attributes (shared by all GO-similarity plots: GO topic (fill) + log10(BF) (size))
node_stats <- df_gene_assignments %>%
    mutate(log_bf = log10(BF)) %>%
    dplyr::select(gene_name, GO_topic, log_bf)

# Bold/outline genes: top hub (rank 1 by kME) + TFs of ASD modules
hubs <- df_module_kme %>%
    filter(module == kME_module & module %in% df_stepwise_res$module) %>%
    group_by(module) %>% slice_max(kME, n = 1, with_ties = FALSE) %>% pull(gene_name)
tfs <- df_stepwise_res %>%
    filter(study %in% c("Wamsley", "Wang")) %>%
    mutate(module_label = str_remove(module_label, "\\ .*")) %>% pull(module_label)
nodes_to_bold <- c(hubs, tfs)

# Short panel titles
GR_short_labels <- c(`1` = "Histone & DNA modification",
                     `2` = "Chromatin remodeling",
                     `3` = "Coactivators & machinery",
                     `4` = "Corepressors & silencing",
                     `5` = "RNA processing",
                     `6` = "Ubiquitin & proteostasis",
                     `7` = "PI3K-mTOR signaling")

# One within-cluster graph per subcluster: each node linked to its top-3 non-zero-similarity
# partners (go_knn_edges); genes left with no edge are dropped.
gr_nets <- map(set_names(names(GR_short_labels)), function(cn) {
    cl    <- intersect(names(cluster_vec[cluster_vec == cn]), rownames(GR_net$sim_mat))
    edges <- go_knn_edges(GR_net$sim_mat[cl, cl, drop = FALSE], k_nn = 3) %>%
        filter(similarity > 0)
    graph_from_data_frame(edges, directed = FALSE)
})

# Shared scale limits so the size / edge-width legends are comparable across panels
sim_rng   <- gr_nets %>% map(~ E(.x)$similarity) %>% unlist() %>% range()
logbf_rng <- node_stats %>%
    filter(gene_name %in% unlist(map(gr_nets, ~ V(.x)$name))) %>%
    pull(log_bf) %>% range(na.rm = TRUE)


# --- Plot --- #

# plot_go_graph() (graphopt layout, GO-topic fill, log10(BF) size, similarity-scaled edges,
# bold hubs/TFs) is sourced from functions/plot_GO_similarity_graphs.R

# Per-panel padding: wide-but-short cells get more horizontal padding (compress x) and less
# vertical (stretch y) to keep node spacing even across panels.
xmult_by_cluster <- c(`1` = 0.10, `2` = 0.15, `3` = 0.40, `4` = 0.15, `5` = 0.35, `6` = 0.25, `7` = 0.40)
ymult_by_cluster <- c(`1` = 0.10, `2` = 0.10, `3` = 0.05, `4` = 0.20, `5` = 0.10, `6` = 0.10, `7` = 0.05)

gr_plots <- imap(gr_nets, ~ plot_go_graph(
    net = .x, node_stats = node_stats, bold_genes = nodes_to_bold,
    title = GR_short_labels[[.y]], size_range = c(1, 4),
    size_limits = logbf_rng, edge_limits = sim_rng,
    seed = 1221, xmult = xmult_by_cluster[[.y]], ymult = ymult_by_cluster[[.y]]
))

# Single full-width header strip (kept separate from the plot function): three category
# boxes connected by arrows, shown above the subcluster grid.
GR_fill_light <- colorRampPalette(c("white", GR_color))(5)[2]  # light tint of GR orange
header_df <- tibble(x = c(1, 2, 3),
                    label = c("Chromatin", "Transcription", "Post-transcription"))
header_strip <- ggplot(header_df, aes(x = x, y = 0, label = label)) +
    # arrows between adjacent boxes (sit in the gaps between box centres)
    annotate("segment", x = 1.45, xend = 1.55, y = 0, yend = 0,
             arrow = arrow(length = unit(0.04, "in"), type = "closed"),
             linewidth = 0.4, color = "grey30") +
    annotate("segment", x = 2.45, xend = 2.55, y = 0, yend = 0,
             arrow = arrow(length = unit(0.04, "in"), type = "closed"),
             linewidth = 0.4, color = "grey30") +
    geom_label(fill = GR_fill_light, color = GR_color, fontface = "bold",
               size = 2.2, label.size = 0.5, label.r = unit(0.1, "lines")) +
    geom_text(color = "black", fontface = "bold", size = 2.2) +
    scale_x_continuous(limits = c(0.5, 3.5)) +
    coord_cartesian(clip = "off") +
    theme_void()

# Subcluster grid: panel AREA proportional to gene count (bigger clusters get more space).
#   col 1: cluster 1 over 2   |   col 2: cluster 3 over 4   |   col 3: cluster 5, 6, 7
layout <- c(
    area(t = 1,    b = 28, l = 1, r = 1),  # 1
    area(t = 28.5, b = 44, l = 1, r = 1),  # 2
    area(t = 1,    b = 13, l = 2, r = 2),  # 3
    area(t = 20,   b = 44, l = 2, r = 2),  # 4
    area(t = 1,    b = 17, l = 3, r = 3),  # 5
    area(t = 18,   b = 32, l = 3, r = 3),  # 6
    area(t = 33,   b = 44, l = 3, r = 3)   # 7
)
p_gr_grid <- gr_plots[["1"]] + gr_plots[["2"]] + gr_plots[["3"]] + gr_plots[["4"]] +
    gr_plots[["5"]] + gr_plots[["6"]] + gr_plots[["7"]] +
    plot_layout(design = layout, widths = c(1.35, 1, 1))

# Header strip on top, subcluster grid below
p_gr_subcluster <- header_strip / p_gr_grid +
    plot_layout(heights = c(0.12, 1), guides = "collect") &
    theme(legend.margin = margin(l = -15))
p_gr_subcluster
```


# --- 8) Combined figure ---

```{r combined_figure, fig.width = 7.09, fig.height = 9.72}
row_top <- p_mods_sig + p_beta + p_module_topic + p_cell_type +
    plot_layout(widths = c(4.5, 0.75, 1, 4.75), nrow = 1)

row_mid <- p_brainspan_traj + free(p_tstat_distr) + plot_spacer() +
    plot_layout(widths = c(1.5, 0.8, 1.5), nrow = 1)

row_bottom <- wrap_elements(p_gr_subcluster)

fig2_combined <- (row_top / row_mid / row_bottom) +
    plot_layout(heights = c(1.65, 0.65, 1.8)) +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(face = "bold", size = 10))

fig2_combined

ggsave(
    filename = file.path(figures_dir, "Fig2", "Fig2_raw.pdf"),
    plot = fig2_combined,
    height    = 247 / 25.4,
    width     = 180 / 25.4,
    dpi       = 300
)
```
