---
title: "Test gene sets for DNM enrichment"
description: "Test synaptic, transcription factor, and epigenetic gene sets for ASD DNM enrichment relative to brain-expressed background"
date: 2026-07-09
format: html
---


# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "05_SynGO")
```

## 0.1 Load data

Load study data

```{r load_data}
# ASC de novo mutation counts and per-gene background mutation rates
# Damaging variants: PTV + Mis2 + Mis1; Mis0/Syn/CNV (Del/Dup) excluded
df_asd_dnm <- read_tsv(
    file.path(data_dir, "full_results_wcounts_2025-10-08.txt"),
    show_col_types = FALSE
) %>%
    clean_names() %>%
    select(gene, gene_id, loeuf, mu_lof, mu_mis2, mu_mis1,
           ptv_proband, ptv_sibling,
           mis2_proband, mis2_sibling,
           mis1_proband, mis1_sibling,
           del_proband, del_sibling,
           dup_proband, dup_sibling) %>%
    mutate(
        mu_total   = mu_lof + mu_mis2 + mu_mis1,
        dn_proband = ptv_proband + mis2_proband + mis1_proband, #+ del_proband + dup_proband,
        dn_sibling = ptv_sibling + mis2_sibling + mis1_sibling #+ del_sibling + dup_sibling
    )

# Gene–topic assignments (from 03_02)
df_gene_assignments   <- read_csv(file.path(analysis_dir, "03_module_enrichments", "03_02_gene-topic-assignment.csv"))
```

Load external gene sets

```{r load_gene_sets}
# Load SynGO data
syngo_dir <- file.path(data_dir, "SynGO")
df_syngo_annot <- read_xlsx(file.path(syngo_dir, "syngo_annotations.xlsx"))
df_syngo_ontol <- read_xlsx(file.path(syngo_dir, "syngo_ontologies.xlsx"))
df_syngo_genes <- read_xlsx(file.path(syngo_dir, "syngo_genes.xlsx"))

# Brain-expressed gene list (Human Protein Atlas, non-HBA definition)
# Source: https://www.proteinatlas.org/search/NOT+tissue_category_rna%3Abrain%3Bnot+detected?format=tsv&download=yes
df_brain <- read_tsv(
    "https://www.proteinatlas.org/search/NOT+tissue_category_rna%3Abrain%3Bnot+detected?format=tsv&download=yes",
    show_col_types = FALSE
) %>%
    clean_names()

# Epigenetic regulators: Engelen et al. 2023 (doi:10.1080/15548627.2022.2139067), Supp Table S2
df_chromatin <- read_xlsx(
    file.path(data_dir, "gene_annotations", "kepi_a_2139067_sm6382(1).xlsx"),
    sheet = 2, skip = 2
) %>% clean_names()

# Transcription factors: AnimalTFDB v1.01 (Ensembl IDs; 2 non-Ensembl rows excluded)
df_tf <- read_tsv(
    file.path(data_dir, "gene_annotations", "TFs_Ensembl_v_1.01.txt"),
    col_names = "ensembl_gene_id",
    show_col_types = FALSE
) %>%
    filter(str_starts(ensembl_gene_id, "ENSG"))
```

## 0.2 Helper function to run mutation-rate-matched permutation

```{r permutation_fn}
# For a given gene set (is_set_col = logical column in df), draw 1,000 mutation-rate-
# matched samples from the remaining brain-expressed genes and compute the Fisher exact
# odds ratio (OR) of proband:sibling DNMs in the gene set vs. each sampled background.
#
# Matching: genes binned into 10 mutation-rate deciles based on the gene set's own
# distribution; background genes sampled proportional to those bin weights.
run_rate_permutation <- function(df, is_set_col, n_iter = 1000, seed = 424) {
    is_set <- df[[is_set_col]]

    cutoffs    <- quantile(df$mu_total[is_set], probs = seq(0, 1, 0.1), na.rm = TRUE)
    cutoffs[1] <- 0

    df     <- df %>% mutate(mu_bin = .bincode(mu_total, breaks = cutoffs)) %>% filter(!is.na(mu_bin))
    is_set <- df[[is_set_col]]

    tbl     <- table(df$mu_bin, is_set)   # rows = bin, cols = FALSE / TRUE
    weights <- tbl[, 2] / tbl[, 1]       # gene-set proportion per bin → sampling weight

    foreground <- df %>% filter(.data[[is_set_col]])
    background <- df %>% filter(!.data[[is_set_col]])
    n_set      <- nrow(foreground)
    bg_weights <- weights[as.character(background$mu_bin)]

    dn_pro_fg <- sum(foreground$dn_proband)
    dn_sib_fg <- sum(foreground$dn_sibling)

    set.seed(seed)
    map_dfr(seq_len(n_iter), function(i) {
        idx     <- sample(nrow(background), n_set, prob = bg_weights, replace = FALSE)
        sampled <- background[idx, ]
        ft      <- fisher.test(matrix(
            c(dn_pro_fg, dn_sib_fg, sum(sampled$dn_proband), sum(sampled$dn_sibling)),
            nrow = 2, byrow = TRUE
        ))
        tibble(iter = i, odds_ratio = ft$estimate, p = ft$p.value)
    })
}
```


# --- 1) Synaptic genes --- #

## 1.1 Define gene set and brain-expressed background

```{r syn_prep}
# SynGO genes are force-included in the brain-expressed background
df_syn <- df_asd_dnm %>%
    mutate(
        is_synaptic = gene_id %in% df_syngo_genes$ensembl_id,
        is_brain    = gene_id %in% df_brain$ensembl | is_synaptic
    ) %>%
    filter(is_brain, !is.na(mu_total), mu_total > 0)

cat("Brain-expressed background N =", nrow(df_syn), "\n")
cat("Synaptic gene set N =", sum(df_syn$is_synaptic), "\n")
```

## 1.2 Run permutation (1,000 iterations)

```{r syn_perm}
res_syn <- run_rate_permutation(df_syn, "is_synaptic")
t_syn <- t.test(res_syn$odds_ratio)

cat(sprintf(
    "Synaptic: mean OR = %.2f (95%% CI: %.2f–%.2f), one-sample t-test p < 2.2e-16\n",
    t_syn$estimate, t_syn$conf.int[1], t_syn$conf.int[2]
))
```


# --- 2) Transcription factors --- #

## 2.1 Define gene set

```{r tf_prep}
# TFs and epigenetic regulators are treated as mutually exclusive sets;
# TF genes that are also chromatin remodellers are assigned to the epigenetic set
df_tf_analysis <- df_asd_dnm %>%
    mutate(
        is_chromatin = gene_id %in% df_chromatin$ensembl_gene_id,
        is_tf        = gene_id %in% df_tf$ensembl_gene_id & !is_chromatin,
        is_brain     = gene_id %in% df_brain$ensembl | is_tf | is_chromatin
    ) %>%
    filter(is_brain, !is.na(mu_total), mu_total > 0)

cat("Brain-expressed background N =", nrow(df_tf_analysis), "\n")
cat("TF gene set N (excluding epigenetic regulators) =", sum(df_tf_analysis$is_tf), "\n")
```

## 2.2 Run permutation

```{r tf_perm}
res_tf <- run_rate_permutation(df_tf_analysis, "is_tf")
t_tf <- t.test(res_tf$odds_ratio)

cat(sprintf(
    "TF: mean OR = %.2f (95%% CI: %.2f–%.2f), one-sample t-test p < 2.2e-16\n",
    t_tf$estimate, t_tf$conf.int[1], t_tf$conf.int[2]
))
```


# --- 3) Epigenetic regulators --- #

## 3.1 Define gene set

```{r chrom_prep}
# Shares the same brain-expressed background as the TF analysis (df_tf_analysis)
cat("Epigenetic regulator gene set N =", sum(df_tf_analysis$is_chromatin), "\n")
```

## 3.2 Run permutation

```{r chrom_perm}
res_chrom <- run_rate_permutation(df_tf_analysis, "is_chromatin")
t_chrom <- t.test(res_chrom$odds_ratio)

cat(sprintf(
    "Epigenetic: mean OR = %.2f (95%% CI: %.2f–%.2f), one-sample t-test p < 2.2e-16\n",
    t_chrom$estimate, t_chrom$conf.int[1], t_chrom$conf.int[2]
))
```


# --- 4) ASD gene-GO topic lists --- #

```{r GO_topic_genes}
df_topic_analysis <- df_asd_dnm %>%
    mutate(
        is_GR = gene %in% c(df_gene_assignments %>% filter(GO_topic == "GR") %>% pull(gene_name)),
        is_MORPH = gene %in% c(df_gene_assignments %>% filter(GO_topic == "MORPH") %>% pull(gene_name)),
        is_SYN = gene %in% c(df_gene_assignments %>% filter(GO_topic == "SYN") %>% pull(gene_name)),
        
        is_brain     = gene_id %in% df_brain$ensembl | is_GR | is_MORPH | is_SYN 
    ) %>%
    filter(is_brain, !is.na(mu_total), mu_total > 0)

res_topics <- map(
    .x = c("is_GR", "is_MORPH", "is_SYN"),
    .f = ~ run_rate_permutation(df_topic_analysis, .x)
)
t_topics <- map(
    .x = 1:length(res_topics),
    .f = ~ t.test(res_topics[[.x]]$odds_ratio)
)

cat(sprintf(
    "GR: mean OR = %.2f (95%% CI: %.2f–%.2f), one-sample t-test p < 2.2e-16\n",
    t_topics[[1]]$estimate, t_topics[[1]]$conf.int[1], t_topics[[1]]$conf.int[2]
))
cat(sprintf(
    "MORPH: mean OR = %.2f (95%% CI: %.2f–%.2f), one-sample t-test p < 2.2e-16\n",
    t_topics[[2]]$estimate, t_topics[[2]]$conf.int[1], t_topics[[2]]$conf.int[2]
))
cat(sprintf(
    "SYN: mean OR = %.2f (95%% CI: %.2f–%.2f), one-sample t-test p < 2.2e-16\n",
    t_topics[[3]]$estimate, t_topics[[3]]$conf.int[1], t_topics[[3]]$conf.int[2]
))
```


# --- 5) Plot --- #

For supplement: Distribution of odds ratios across permutations

## 5.1 Gene sets (not ASD risk genes)

```{r gene_set_OR, fig.width = 3.5, fig.height = 3}
df_gene_set_OR <- bind_rows(
    res_syn   %>% mutate(gene_set = "Synaptic"),
    res_tf    %>% mutate(gene_set = "Transcription\nfactor"),
    res_chrom %>% mutate(gene_set = "Epigenetic\nregulator")
) %>%
    mutate(gene_set = factor(
        gene_set,
        levels = c("Epigenetic\nregulator", "Transcription\nfactor", "Synaptic")
    ))

df_means <- df_gene_set_OR %>%
    group_by(gene_set) %>%
    summarise(
        mean_or = mean(odds_ratio),
        ci_lo   = t.test(odds_ratio)$conf.int[1],
        ci_hi   = t.test(odds_ratio)$conf.int[2],
        .groups = "drop"
    )

p_gene_set_OR <- ggplot(df_gene_set_OR, 
                        aes(x = odds_ratio, y = gene_set)) +
    geom_quasirandom(
        aes(color = gene_set), 
        groupOnX = FALSE, size = 0.3, alpha = 0.25
    ) +
    geom_point(
        data = df_means,
        aes(x = mean_or, y = gene_set),
        shape = 21, fill = "white",
        color = "black", size = 2,
        inherit.aes = FALSE
    ) +
    geom_vline(xintercept = 1, lty = 2, color = "gray40") +
    scale_color_manual(
        values = c(topic_colors[["GR"]], topic_colors[["MORPH"]], topic_colors[["SYN"]]),
        guide = "none"
    ) +
    labs(
        x     = "Odds ratio",
        y     = NULL,
        title = "DNM enrichment of gene sets vs\nbrain-expressed background"
    ) +
    theme(
        plot.title = element_text(size = 7),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 6.5)
    )
p_gene_set_OR
```

## 5.2 ASD risk genes by GO topic

```{r p_ASD_genes, fig.width = 3.5, fig.height = 3}
df_GO_topic_OR <- bind_rows(
    res_topics[[1]]   %>% mutate(GO_topic = "GR"),
    res_topics[[2]]   %>% mutate(GO_topic = "MORPH"),
    res_topics[[3]]   %>% mutate(GO_topic = "SYN")
) %>%
    mutate(GO_topic = factor(
        GO_topic,
        levels = c("GR", "MORPH", "SYN")
    ))

df_means <- df_GO_topic_OR %>%
    group_by(GO_topic) %>%
    summarise(
        mean_or = mean(odds_ratio),
        ci_lo   = t.test(odds_ratio)$conf.int[1],
        ci_hi   = t.test(odds_ratio)$conf.int[2],
        .groups = "drop"
    )

p_GO_topic_OR <- ggplot(df_GO_topic_OR, 
                        aes(x = odds_ratio, y = GO_topic)) +
    geom_quasirandom(
        aes(color = GO_topic), 
        groupOnX = FALSE, size = 0.3, alpha = 0.25
    ) +
    geom_point(
        data = df_means,
        aes(x = mean_or, y = GO_topic),
        shape = 21, fill = "white",
        color = "black", size = 2,
        inherit.aes = FALSE
    ) +
    geom_vline(xintercept = 1, lty = 2, color = "gray40") +
    scale_color_manual(
        values = topic_colors,
        guide = "none"
    ) +
    labs(
        x     = "Odds ratio",
        y     = NULL,
        title = "DNM enrichment of ASD genes vs\nbrain-expressed background"
    ) +
    theme(
        plot.title = element_text(size = 7),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 6.5)
    )
p_GO_topic_OR
```


## 5.3 Combine

```{r p_combine, fig.height = 3.5, fig.width = 6.5}
p_combined <- p_gene_set_OR + p_GO_topic_OR &
    plot_annotation(tag_levels = "A")
ggsave(
    file = file.path(figures_dir, "supplement", "05_00_gene-set-OR-vs-brain-bg.pdf"),
    plot = p_combined,
    height = 3,
    width = 6.5,
    dpi = 300
)
```
