---
title: "Module synapse and HotNet enrichment"
description: "Fisher's exact overlap of each significant module with SynGO synaptic genes overall, then with individual SynGO HotNet neighborhoods"
format: html
---

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- paste0(here::here(), "/")
source(paste0(project_dir, "setup.R"))

# Set output directory for this notebook
notebook_out_dir <- file.path(analysis_dir, "06_MORPH_annotation")
```

```{r env}
# ASD-associated modules
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "02_module_identification", "02_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results

# Genes in each module
df_network_data <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_gene-networks-all.RDS"))

# Module-topic assignments (restart stability of the topic model)
topic_stability <- readRDS(
    file.path(analysis_dir, "03_module_enrichments", "03_00_topic-restart-stability.RDS")
)
df_top_topic <- topic_stability$consensus %>% 
    dplyr::select(module, topic_label = consensus_topic)

# SynGO HotNet gene sets (from 04_01)
hotnet_gene_lists <- readRDS(file.path(analysis_dir, "05_SynGO", "05_02_hotnet-gene-lists.RDS"))

# SynGO ontology background
syngo_dir <- file.path(data_dir, "SynGO")
df_syngo_ontol <- read_xlsx(file.path(syngo_dir, "syngo_ontologies.xlsx"))

# Brain-expressed gene list (Human Protein Atlas, non-HBA definition)
# Synaptic enrichment test background
df_brain <- read_tsv(
    "https://www.proteinatlas.org/search/NOT+tissue_category_rna%3Abrain%3Bnot+detected?format=tsv&download=yes",
    show_col_types = FALSE
) %>%
    clean_names()
```


# --- 1) Overall synapse enrichment --- #

Before asking which synaptic neighborhood a module maps onto (section 2), test whether
each module is enriched for synaptic genes at all (Fisher's exact test, two-sided).

SynGO splits into two branches, so all three definitions are tested and reported: the
**CC root** (synapse, GO:0045202, 1,487 genes) captures genes physically localized to the
synapse; the **BP root** (process in the synapse, SYNGO:synprocess, 1,126 genes) captures
genes participating in a synaptic process; and their union is every gene SynGO curates
(1,602). All three share one background so only the foreground differs.

In practice the branches are close to nested: 1,011 of the 1,126 BP genes are also
annotated to CC, leaving only 115 BP-only genes. SynGO curates genes *at* the synapse and
its BP terms describe what those genes do there, so "location" and "process" do not
separate cleanly within this resource. Consistent with that, all three definitions flag
the same 21 modules, and topic-level odds ratios differ only slightly (GR 1.34 / 1.37 /
1.40, MORPH 2.65 / 2.42 / 2.71, SYN 3.93 / 4.03 / 3.82 for CC / BP / union). The CC root
is used for the main figure as the most direct test of synaptic localization; the other
two are retained as a sensitivity analysis.

Both sections use two-sided tests so that odds ratios come with finite confidence
intervals; a one-sided test returns an infinite upper bound by construction.

Background: the module side is that study's network genes, and the gene-set side is the
HPA brain-expressed list used in 04_00; `ORA_fisher()` intersects the two, so each module
is compared against other brain-expressed genes *in its own network* and never against
the genome. Restricting to brain-expressed genes is what guards against the standard
confound that synaptic genes are brain-expressed, and it keeps the baseline synaptic rate
comparable across studies, which matters because networks differ several-fold in how many
non-brain and unannotated genes they carry.

Substituting an all-GO-annotated background changes no conclusion: the same 21 modules
reach BH < 0.05 with fold enrichment > 1 under either definition, and under all
GO-annotated genes every odds ratio is slightly larger (SYN median 4.20 vs 3.82, GR 1.52
vs 1.40) because a genome-wide background dilutes the synaptic baseline. The
brain-expressed background is the more conservative of the two.

Note this is not independent of the GO topic assignments — the SYN topic was itself
derived from GO enrichment of these same modules. SynGO is expert-curated where the
topic model is text-derived, so this is corroboration by an independent resource rather
than independent evidence.

## 1.1 Run ORA: modules × SynGO synaptic genes

```{r syngo_module_ora}
# Helper: genes annotated to one SynGO term (semicolon-free, comma-delimited symbols)
get_syngo_term_genes <- function(term_id) {
    df_syngo_ontol %>%
        filter(id == term_id) %>%
        pull(hgnc_symbol) %>%
        str_split(", ") %>%
        unlist() %>%
        str_trim() %>%
        unique()
}

# All SynGO-annotated genes (CC or BP). Also the background for the HotNet tests in
# section 2, two of which are BP-derived and so need BP genes in their universe
syngo_gene_set <- df_syngo_ontol %>%
    dplyr::select(hgnc_symbol) %>%
    separate_rows(hgnc_symbol, sep = ", ") %>%
    mutate(hgnc_symbol = str_trim(hgnc_symbol)) %>%
    pull(hgnc_symbol) %>%
    unique()

# Three definitions of "synaptic", tested separately. The CC root asks about physical
# location at the synapse; the BP root asks about participation in a synaptic process;
# the union is every gene SynGO curates. The two branches overlap heavily but are not
# nested, so which one is used is a substantive choice rather than a technicality
synaptic_gene_sets <- list(
    "Synapse (CC)"          = get_syngo_term_genes("GO:0045202"),
    "Synaptic process (BP)" = get_syngo_term_genes("SYNGO:synprocess"),
    "All SynGO"             = syngo_gene_set
)

cat("gene set sizes:\n")
walk2(names(synaptic_gene_sets), synaptic_gene_sets,
      ~ cat(sprintf("  %-22s %d genes\n", .x, length(.y))))
cat(sprintf("  CC and BP overlap: %d | CC only: %d | BP only: %d\n",
            length(intersect(synaptic_gene_sets[[1]], synaptic_gene_sets[[2]])),
            length(setdiff(synaptic_gene_sets[[1]], synaptic_gene_sets[[2]])),
            length(setdiff(synaptic_gene_sets[[2]], synaptic_gene_sets[[1]]))))

# Brain-expressed background, as HGNC symbols to match gene_name in the network data.
# Every SynGO gene is force-included (as in 04_00) so that a synaptic gene missing from
# the HPA list is not dropped, and so that all three tests share one identical universe
brain_background <- union(df_brain$gene, syngo_gene_set)

# Run ORA: each module against each of the three synaptic gene set definitions
df_module_syngo_ora <- df_stepwise_res %>%
    dplyr::select(study, module) %>%
    left_join(df_top_topic) %>%
    expand_grid(gene_set = names(synaptic_gene_sets)) %>%
    mutate(
        ora_res = pmap(
            .l = list(..1 = study, ..2 = module, ..3 = gene_set),
            .f = ~ {
                df_study_tmp <- df_network_data %>%
                    filter(study == ..1) %>%
                    unnest(cols = c(data))

                study_background <- df_study_tmp %>% pull(gene_name) %>% unique()
                module_genes     <- df_study_tmp %>% filter(module == ..2) %>%
                    pull(gene_name) %>% unique()

                ORA_fisher(
                    gene_list1       = module_genes,
                    gene_list2       = synaptic_gene_sets[[..3]],
                    background_list1 = study_background,
                    background_list2 = brain_background,
                    alternative      = "two.sided"
                )
            }
        )
    ) %>%
    unnest(cols = c(ora_res)) %>%
    # each gene set definition is its own family of tests across modules
    group_by(gene_set) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"), .before = n_overlap) %>%
    ungroup() %>%
    dplyr::rename(n_module = n_list1, n_gene_set = n_list2) %>%
    mutate(
        pct_module = n_overlap / n_module * 100,
        gene_set   = factor(gene_set, levels = names(synaptic_gene_sets))
    ) %>%
    arrange(gene_set, p_value) %>%
    left_join(df_stepwise_res %>% dplyr::select(module, module_label)) %>%
    dplyr::select(study, module, module_label, topic_label, gene_set, everything())

# Compare the three definitions: how many modules are enriched, and how strongly
df_module_syngo_ora %>%
    group_by(gene_set, topic_label) %>%
    summarise(median_OR = round(median(odds_ratio), 2),
              n_sig     = sum(p_adj < 0.05 & fold_enrichment > 1),
              n         = n(), .groups = "drop") %>%
    pivot_wider(names_from = topic_label, values_from = c(median_OR, n_sig, n)) %>%
    print()
```

## 1.2 Save results

```{r save_module_syngo_ora}
saveRDS(df_module_syngo_ora,
        file.path(notebook_out_dir, "06_00_module-SynGO-ORA.RDS")
)
```


# --- 2) Module HotNet enrichment --- #

Modules from the SYN topic are expected to be enriched for synaptic genes. Here we
test, for each module × HotNet neighborhood combination, whether module genes are
over-represented among HotNet genes (Fisher's exact test, two-sided). The background
for each module is the set of genes in that study network.

## 2.1 Run ORA: modules × HotNet neighborhoods

```{r hotnet_module_ora}
# Background for each module = all genes in that study's network
# Foreground = module gene set
# Gene set 2 = HotNet neighborhood genes (same across all studies)
# BH correction applied within each HotNet neighborhood

hotnet_names <- names(hotnet_gene_lists)

df_module_hotnet_ora <- df_stepwise_res %>%
    dplyr::select(study, module) %>%
    left_join(df_top_topic) %>%
    expand_grid(hotnet_group = hotnet_names) %>%
    mutate(
        ora_res = pmap(
            .l = list(..1 = study, ..2 = module, ..3 = hotnet_group),
            .f = ~ {
                df_study_tmp <- df_network_data %>%
                    filter(study == ..1) %>%
                    unnest(cols = c(data))
                
                study_background <- df_study_tmp %>% pull(gene_name) %>% unique()
                module_genes     <- df_study_tmp %>% filter(module == ..2) %>%
                    pull(gene_name) %>% unique()
                hotnet_gs        <- hotnet_gene_lists[[..3]]
                
                ORA_fisher(
                    gene_list1       = module_genes,
                    gene_list2       = hotnet_gs,
                    background_list1 = study_background,
                    background_list2 = syngo_gene_set,
                    alternative      = "two.sided"
                )
            }
        )
    ) %>%
    unnest(cols = c(ora_res)) %>%
    group_by(hotnet_group) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"), .before = n_overlap) %>%
    ungroup() %>%
    dplyr::rename(n_module = n_list1, n_hotnet = n_list2) %>%
    mutate(pct_module = n_overlap / n_module * 100) %>%
    arrange(hotnet_group, p_value) %>% 
    left_join(df_stepwise_res %>% dplyr::select(module, module_label)) %>% 
    dplyr::select(study, module, module_label, everything())

print(df_module_hotnet_ora %>% filter(p_value < 0.05))
```

## 2.2 Save results

```{r save_module_hotnet_ora}
saveRDS(df_module_hotnet_ora, 
        file.path(notebook_out_dir, "06_00_module-HotNet-ORA.RDS")
)
```


# --- 3) Export results tables ---

```{r export_tables}
write_csv(
    df_module_syngo_ora,
    file = file.path(notebook_out_dir, "06_00_module-SynGO-ORA.csv")
)
write_csv(
    df_module_hotnet_ora,
    file = file.path(notebook_out_dir, "06_00_module-HotNet-ORA.csv")
)
```


