---
title: "Module HotNet enrichment"
description: "Fisher's exact overlap of each significant module with SynGO HotNet neighborhoods"
format: html
---

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set output directory for this notebook
notebook_out_dir <- file.path(analysis_dir, "04_MORPH_MEF2C")
```

```{r env}
# ASD-associated modules
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "01_module_identification", "01_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results

# Genes in each module
df_network_data <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_gene-networks-all.RDS"))

# Module-topic assignments
df_gamma <- readRDS(file.path(analysis_dir, "02_module_enrichments", "02_00_topic-module-gamma.RDS"))
df_top_topic <- df_gamma %>%
    filter(top_topic == "yes") %>%
    distinct(module, topic_label)

# SynGO HotNet gene sets (from 03_01)
hotnet_gene_lists <- readRDS(file.path(analysis_dir, "03_SynGO", "03_01_hotnet-gene-lists.RDS"))

# SynGO ontology background
syngo_dir <- file.path(data_dir, "SynGO")
df_syngo_ontol <- read_xlsx(file.path(syngo_dir, "syngo_ontologies.xlsx"))
```


# --- 1) Module HotNet enrichment --- #

Modules from the SYN topic are expected to be enriched for synaptic genes. Here we
test, for each module × HotNet neighborhood combination, whether module genes are
over-represented among HotNet genes (Fisher's exact test, one-sided). The background
for each module is the set of genes in that study network.

## 1.1 Run ORA: modules × HotNet neighborhoods

```{r hotnet_module_ora}
# Background for each module = all genes in that study's network
# Foreground = module gene set
# Gene set 2 = HotNet neighborhood genes (same across all studies)
# BH correction applied within each HotNet neighborhood

syngo_background <- df_syngo_ontol %>%
    dplyr::select(hgnc_symbol) %>%
    separate_rows(hgnc_symbol, sep = ", ") %>%
    mutate(hgnc_symbol = str_trim(hgnc_symbol)) %>%
    pull(hgnc_symbol) %>% unique()

hotnet_names <- names(hotnet_gene_lists)

df_module_hotnet_ora <- df_stepwise_res %>%
    dplyr::select(study, module) %>%
    left_join(df_top_topic) %>%
    expand_grid(hotnet_group = hotnet_names) %>%
    mutate(
        ora_res = pmap(
            .l = list(..1 = study, ..2 = module, ..3 = hotnet_group),
            .f = ~ {
                df_study_tmp <- df_network_data %>%
                    filter(study == ..1) %>%
                    unnest(cols = c(data))
                
                study_background <- df_study_tmp %>% pull(gene_name) %>% unique()
                module_genes     <- df_study_tmp %>% filter(module == ..2) %>%
                    pull(gene_name) %>% unique()
                hotnet_gs        <- hotnet_gene_lists[[..3]]
                
                ORA_fisher(
                    gene_list1       = module_genes,
                    gene_list2       = hotnet_gs,
                    background_list1 = study_background,
                    background_list2 = syngo_background,
                    alternative      = "greater"
                )
            }
        )
    ) %>%
    unnest(cols = c(ora_res)) %>%
    group_by(hotnet_group) %>%
    mutate(p_adj = p.adjust(p_value, method = "BH"), .before = n_overlap) %>%
    ungroup() %>%
    dplyr::rename(n_module = n_list1, n_hotnet = n_list2) %>%
    mutate(pct_module = n_overlap / n_module * 100) %>%
    arrange(hotnet_group, p_value) %>% 
    left_join(df_stepwise_res %>% dplyr::select(module, module_label)) %>% 
    dplyr::select(study, module, module_label, everything())

print(df_module_hotnet_ora %>% filter(p_value < 0.05))
```

## 1.2 Save results

```{r save_module_hotnet_ora}
saveRDS(df_module_hotnet_ora, 
        file.path(notebook_out_dir, "04_00_module-HotNet-ORA.RDS")
)
```
