---
title: "ASD-associated module GO and cell-type enrichments"
description: "Run GO ORA on all significant modules; use topic modeling to synthesize biological themes"
date: 2026-07-07
format: html
---


# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "02_module_enrichments")
```

Load data

```{r env}
# Gene network assignments
df_network_data <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_gene-networks-all.RDS"))

# Significant ASD modules
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "01_module_identification", "01_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results %>% 
    mutate(module_label = factor(module_label, levels = unique(.$module_label)))
```


# --- 1) GO enrichment --- #

## 1.1 Load GO terms

```{r load_GO}
# GO term description to ID mapping
go_terms <- AnnotationDbi::select(
    GO.db,
    keys = keys(GO.db),
    columns = c("TERM", "ONTOLOGY"),
    keytype = "GOID"
) %>% 
    as_tibble() %>% 
    clean_names %>% 
    dplyr::rename(go_id = goid)

# GO ID to gene mapping
go_mapping <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = keys(org.Hs.eg.db, keytype = "ENSEMBL"),
    columns = c("GO", "ONTOLOGY"),
    keytype = "ENSEMBL"
) %>% 
    as_tibble() %>% 
    clean_names %>% 
    dplyr::rename(go_id = go, ensembl_gene_id = ensembl) %>% 
    dplyr::select(-evidence)

# Identify GO database background
go_background <- go_mapping %>% pull(ensembl_gene_id) %>% unique
```

## 1.2 Prep GO data

```{r prep_GO_data}
## To save time, precompute gene lists ##

# Precompute study & module gene sets
df_network_unnested <- df_network_data %>%
    tidyr::unnest(data)

# Study-level genes (for enrichment backgrounds)
study_genes_list <- df_network_unnested %>%
    group_by(study) %>% 
    summarise(study_genes = list(unique(ensembl_gene_id)), .groups = "drop") %>%
    deframe()  # named list: study -> genes

# Module-level genes
module_genes_list <- df_network_unnested %>%
    group_by(study, module) %>%
    summarise(module_genes = list(unique(ensembl_gene_id)), .groups = "drop") %>% 
    filter(module %in% df_stepwise_res$module)

# GO --> genes list (filtered)
go_genes_list <- go_mapping %>%
    group_by(go_id) %>%
    summarise(genes = list(unique(ensembl_gene_id)), .groups = "drop") %>%
    filter(lengths(genes) >= 5 & lengths(genes) <= 500) %>% # remove GO terms with fewer than 5 genes or greater than 500 genes (removes broad, non-informative terms)
    deframe()
```

## 1.3 Run GO ORA

```{r run_GO_ORA}
## Run ORA on all combinations of modules and GO paths

results <- list()

for(i in seq_len(nrow(module_genes_list))) {
    study <- module_genes_list$study[i]
    module <- module_genes_list$module[[i]]
    module_genes <- module_genes_list$module_genes[[i]]
    
    print(paste0("Running GO enrichment on ", module))
    
    # Only GO terms that overlap > 1
    go_overlap <- names(go_genes_list)[map_lgl(go_genes_list, ~ length(intersect(.x, module_genes)) > 1)]
    
    for(go_id in go_overlap) {
        go_genes <- go_genes_list[[go_id]]
        study_genes <- study_genes_list[[study]]
        
        results[[length(results)+1]] <- ORA_fisher(
            gene_list1 = module_genes,
            background_list1 = study_genes,
            gene_list2 = go_genes,
            background_list2 = go_background,
            alternative = "greater"
        ) %>% 
            mutate(study = study, module = module, go_id = go_id)
    }
}

df_go_res <- bind_rows(results)

# Format results
df_go_res <- df_go_res %>% 
    left_join(go_terms) %>% 
    dplyr::select(study, module, go_id, term, ontology, everything()) %>% 
    group_by(study, module) %>% 
    mutate(p_adj = p.adjust(p_value, method = "BH"), .before = n_overlap) %>% 
    mutate(module = factor(module, levels = df_stepwise_res$module)) %>% 
    arrange(module, p_value) %>% 
    dplyr::rename(n_module = n_list1, n_pathway = n_list2) %>% 
    ungroup %>% 
    mutate(pct_module = n_overlap / n_module * 100)
```

## 1.4 Export results as object and table for supplement

```{r save_GO_res}
# Save output as RDS file
saveRDS(
    df_go_res,
    file = file.path(notebook_out_dir, "02_00_sig-mod-GO-enrichments.RDS")
)
# Write significant results to Excel
write_csv(
    df_go_res %>% filter(p_value < 0.05),
    path = file.path(notebook_out_dir, "02_00_sig-mod-GO-enrichments-p0.05.csv")
)
```


# --- 2) Topic modeling --- #

Run topic modeling on all significant GO terms to extract common themes among enrichments, and quantify module relationship to each term. \n NOTE: We have too few documents (modules, n = 29) to run searchK and identify optimal number of modules. We select K = 3 based on qualitiative evaluation of biological themes

## 2.1 Prep sparse module-word matrix

```{r data_prep}
# Unnest tokens to get GO terms within each module (group)
df_go_tokens <- df_go_res %>% 
    dplyr::filter(!is.na(term)) %>% 
    group_by(module) %>% 
    # if any term in the module has p_adj < 0.05, keep only those.
    # else, keep terms where p_value < 0.05
    dplyr::filter(
        if (any(p_adj < 0.05)) {
            p_adj < 0.05
        } else {
            p_value < 0.05
        }
    ) %>% 
    ungroup() %>% 
    unnest_tokens(word, term)

## Create a sparse matrix using only meaningful words that appear at least twice within module
generic_terms <- c("positive", "negative", "regulation", "response", "protein", "activity",
                   "pathway", "process", "involved", "cell", 
                   "binding", "complex", "signaling", "localization", "organization", 
                   "component", "system", "level", "activation", "production",
                   "kinase", "membrane", "dependent", "factor", "substrate", "mediated", "type", "formation")

# 1. Identify which modules drop out at n > 2
modules_all <- unique(df_go_tokens$module)
modules_at_n2 <- df_go_tokens %>%
    dplyr::count(module, word) %>%
    anti_join(stop_words, by = "word") %>% 
    dplyr::filter(!(word %in% generic_terms)) %>% 
    dplyr::filter(n > 2) %>%
    pull(module) %>%
    unique()

missing_modules <- setdiff(modules_all, modules_at_n2)

# 2. Extract the top words for those specific missing modules
whitelist_words <- df_go_tokens %>%
    filter(module %in% missing_modules) %>%
    dplyr::count(module, word) %>%
    anti_join(stop_words, by = "word") %>%
    dplyr::filter(!(word %in% generic_terms)) %>%
    group_by(module) %>%
    slice_max(n, n = 3) %>% # Keep top 3 words per missing module
    pull(word) %>%
    unique()

# 3. Create the sparse matrix with n > 2 OR Whitelist
go_sparse <- df_go_tokens %>%
    dplyr::count(module, word) %>%
    anti_join(stop_words, by = "word") %>% 
    dplyr::filter(!(word %in% generic_terms)) %>% 
    # Logic: Keep if frequency > 2 OR if it's a vital word for the missing modules
    dplyr::filter(n > 2 | word %in% whitelist_words) %>%
    cast_sparse(module, word, n)
```

## 2.2 Select k

```{r select_k, fig.height = 4, fig.width = 7}
k_range <- 2:10
k_models <- map(k_range, ~ stm(go_sparse, K = .x, verbose = FALSE, seed = 234))

df_k_metrics <- map2_dfr(k_range, k_models, function(k, m) {
    tibble(
        K      = k,
        semcoh = mean(semanticCoherence(m, go_sparse)),
        exclus = mean(exclusivity(m))
    )
})
```

## 2.3 Select final topic model

```{r run_topic_model}
# Fit topic model on K = 3
topic_model <- k_models[[2]]

# Save fitted topic model for downstream analyses & figures
saveRDS(topic_model, file = file.path(notebook_out_dir, "02_00_topic-model-K3.RDS"))
```

## 2.4 Extract topic model metrics

Beta = word-topic probability

```{r extract_beta}
# Beta = overall probability that a word belongs to a given topic

# Quantify word-topic significance
df_beta <- tidytext::tidy(topic_model, matrix = "beta") %>%
    group_by(topic) %>%
    mutate(
        # Z-score for how much more likely this word is in this topic vs others
        z_score = (beta - mean(beta)) / sd(beta),
        p_value = pnorm(z_score, lower.tail = FALSE),
        p_adj = p.adjust(p_value, method = "BH")
    ) %>%
    ungroup()

# Save for figure
saveRDS(df_beta, file = file.path(notebook_out_dir, "02_00_topic-word-beta.RDS"))
```

Gamma = module-topic probability

```{r extract_gamma}
# Quantify module-topic probabilities & label topics
df_gamma <- tidytext::tidy(
    topic_model,
    matrix = "gamma",
    document_names = rownames(go_sparse)
) %>% 
    dplyr::rename(module = document) %>% 
    
    # Calculate z-score and p-value of module association with a given topic compared to other modules
    group_by(topic) %>% 
    mutate(
        z_score = (gamma - mean(gamma)) / sd(gamma),
        p_value = pnorm(z_score, lower.tail = FALSE),
        p_adj = p.adjust(p_value, method = "BH")
    ) %>%
    ungroup() %>% 
    
    # rename modules based on word associations
    mutate(topic = case_when(
        topic == 1 ~ "genomic regulation / \nRNA processing",
        topic == 2 ~ "synaptic transmission / \nion transport",
        topic == 3 ~ "neuron development / \nmorphogenesis"
    ),
    topic = factor(topic, 
                   levels = c("genomic regulation / \nRNA processing", 
                              "neuron development / \nmorphogenesis",
                              "synaptic transmission / \nion transport")
    ),
    topic_label = case_when(
        topic == "synaptic transmission / \nion transport" ~ "SYN",
        topic == "neuron development / \nmorphogenesis" ~ "MORPH",
        topic == "genomic regulation / \nRNA processing" ~ "GR"
    ),
    topic_color = case_when(
        topic_label == "SYN" ~ "#4a2377",
        topic_label == "MORPH" ~ "#0d7d87",
        topic_label == "GR" ~ "#ea801c"
    )
    ) %>% 
    
    # Identify if the topic is the top topic for a given module
    group_by(module) %>% 
    mutate(top_topic = ifelse(gamma == max(gamma), "yes", "no")) %>% 
    ungroup

# Save for downstream analyses & figures
saveRDS(df_gamma, file = file.path(notebook_out_dir, "02_00_topic-module-gamma.RDS"))
```



# --- 3) Supplement figures --- #


## 3.1 Select k for topic model

```{r plot_select_k, fig.height = 3, fig.width = 3}
# p_select_k <- df_k_metrics %>%
#     pivot_longer(-K, names_to = "metric", values_to = "value") %>%
#     mutate(metric = recode(metric,
#         semcoh = "Semantic coherence",
#         exclus = "Exclusivity"
#     )) %>%
#     ggplot(aes(x = K, y = value)) +
#     geom_line(color = "steelblue", linewidth = 0.7) +
#     geom_point(color = "steelblue", size = 1.5) +
#     facet_wrap(~ metric, scales = "free_y", nrow = 1) +
#     scale_x_continuous(breaks = k_range) +
#     labs(
#         x     = "Number of topics (K)",
#         y     = NULL,
#         title =  NULL #"K selection: semantic coherence & exclusivity"
#     ) +
#     theme(
#         strip.text       = element_text(face = "bold"),
#         strip.background = element_blank()
#     )

p_k_tradeoff <- df_k_metrics %>%
    mutate(color = ifelse(K == 3, "firebrick", "steelblue")) %>% 

    ggplot(aes(x = semcoh, y = exclus, label = K)) +
    geom_path(color = "grey70", linewidth = 0.5) +
    geom_point(aes(color = I(color)), size = 2) +
    ggrepel::geom_text_repel(size = 2.5, min.segment.length = 0) +
    labs(
        x     = "Semantic coherence",
        y     = "Exclusivity",
        title = "Coherence vs exclusivity tradeoff"
    )
p_k_tradeoff

ggsave(
    filename = file.path(figures_dir, "supplement", "02_02_topic-model-K-selection.png"),
    plot = p_k_tradeoff,
    width = 3,
    height = 3,
    dpi = 300
)
```


# --- 4) Tables --- #

## 4.1 Topic model summary

Export the top words per topic across STM metrics (highest prob / FREX / lift / score),
wide format with one column block per topic (SYN, GR, MORPH).

```{r topic_top_words_table}
n_top_words <- 7
lt <- stm::labelTopics(topic_model, n = n_top_words)

# Map each topic number -> theme label via a signature top-probability word
# (robust to topic renumbering across re-runs)
topic_theme <- apply(lt$prob, 1, function(words) {
    if      ("synaptic"          %in% words) "SYN"
    else if ("acetyltransferase" %in% words) "GR"
    else if ("morphogenesis"     %in% words) "MORPH"
    else                                     NA_character_
})
topic_for_label <- setNames(seq_along(topic_theme), topic_theme)

# Assemble <LABEL>_<metric> columns in SYN, GR, MORPH order
df_topic_top_words <- purrr::map_dfc(c("SYN", "GR", "MORPH"), function(label) {
    k <- topic_for_label[[label]]
    setNames(
        list(lt$prob[k, ], lt$frex[k, ], lt$lift[k, ], lt$score[k, ]),
        paste0(label, c("_prob", "_frex", "_lift", "_score"))
    ) %>% as_tibble()
})

write_csv(df_topic_top_words,
          file.path(notebook_out_dir, "02_00_topic-top-words.csv")
)
```

## 4.2 Module-topic probabilities (gamma)

```{r module_gamma}
df_gamma %>% 
    dplyr::select(module, topic_label, topic, top_topic, gamma, z_score, p_value, p_adj) %>% 
    mutate(topic = str_remove(topic, "\\n")) %>% 
    
    write_csv(file.path(notebook_out_dir, "02_00_topic-module-gamma.csv"))
```


## DEPR

## 3.2  Word-topic probabilities (beta)

```{r plot_beta, fig.height = 2.5, fig.width = 6.5}
# Plot module-word associations to define topics
## Assign colors
topic_colors <- c(
    "SYN" = "#4a2377",
    "GR" = "#ea801c",
    "MORPH" = "#0d7d87"
)

p_beta <- df_beta %>% 
    mutate(topic_label = case_when(
        topic == 1 ~ "GR",
        topic == 2 ~ "SYN",
        topic == 3 ~ "MORPH"
    )) %>% 
    group_by(topic) %>% 
    slice_max(beta, n = 20) %>% # Keep top words per topic
    ungroup %>% 

    ggplot(aes(x = beta, y = reorder_within(term, beta, topic))) +
    geom_col(aes(fill = topic_label, alpha = z_score, color = p_adj < 0.05), linewidth = 0.25) +
    facet_wrap(vars(topic_label), scales = "free_y", nrow = 1) +
    
    scale_alpha_continuous(limits = c(0, 15.3), name = "Z-score") +
    scale_fill_manual(values = topic_colors, guide = "none") +
    scale_color_manual(values = c("TRUE" = "black", "FALSE" = "gray"), name = "FDR < 0.05") +
    scale_y_reordered() +
    scale_x_continuous(breaks = c(0, 0.05, 0.09)) +
    guides(color = guide_legend(override.aes = list(fill = "white"))) +
    labs(x = "Beta (word-topic probability)", y = NULL,
         title = NULL #"Highest probability per topic"
         ) +
    theme(legend.key.size = unit(0.3, "cm"),
          legend.margin = margin(l = -5),
          panel.grid.major.y = element_blank()
    )
p_beta
```

