---
title: "Assign genes to GO topics"
description: "Assignment of ASD risk genes to topics (SYN/GR/MORPH) by GO-pathway overlap"
date: 2026-07-08
format: html
---

Each ASD risk gene (TADA FDR < 0.001) is assigned to a single topic (SYN, GR, or
MORPH) by counting how many of its GO Biological Process terms overlap each topic's
characteristic GO vocabulary, then taking the topic with the greatest overlap.

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "02_module_enrichments")
```

## 0.1 Load study data

```{r load_data}
# ASD rare variant statistics
df_asd_rarevar_zscore <- readRDS(file.path(analysis_dir, "00_data_format", "00_01_ASD-rarevar-stats.RDS"))

# Significant ASD modules
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "01_module_identification", "01_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results %>% 
    mutate(module_label = factor(module_label, levels = unique(.$module_label)))

# Module GO enrichments
df_go_res <- readRDS(file.path(notebook_out_dir, "02_00_sig-mod-GO-enrichments.RDS"))

# Module-topic gamma
df_gamma <- readRDS(file.path(notebook_out_dir, "02_00_topic-module-gamma.RDS"))
```

## 0.2 Load GO data

```{r load_GO_data}
# GO ID to gene mapping
go_mapping <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = keys(org.Hs.eg.db, keytype = "ENSEMBL"),
  columns = c("GO", "ONTOLOGY"),
  keytype = "ENSEMBL"
) %>%
    as_tibble() %>%
    clean_names %>%
    dplyr::rename(go_id = go, ensembl_gene_id = ensembl) %>%
    dplyr::select(-evidence) #%>%

    # # filter pathways by size
    # filter(ontology == "BP") %>%
    # group_by(go_id) %>%
    # mutate(n_genes = n_distinct(ensembl_gene_id)) %>%
    # ungroup() %>%
    # filter(n_genes > 5 & n_genes < 1000)
```


# --- 1) Topic vocabulary & per-gene topic overlap scores --- #

```{r identify_module_GO_paths}
# Identify ASD risk genes
asd_risk_genes <- c(df_asd_rarevar_zscore %>%
                        filter(FDR < 0.001 & !FLAG) %>%
                        pull(ensembl_gene_id) 
)

# Step 1: build topic vocabulary from enriched GO terms
# joining to df_gamma to get topic assignment per module
# (all GO IDs mapped to each topic)
topic_vocab <- df_go_res %>%
    filter(p_adj < 0.05, ontology == "BP") %>%
    left_join(df_gamma %>%
                  filter(top_topic == "yes") %>%
                  select(module, topic),
              by = "module") %>%
    filter(!is.na(topic)) %>%
    distinct(topic, go_id) %>%

    # add topic label
    mutate(
        topic = case_when(
            str_detect(topic, "^syn") ~ "SYN",
            str_detect(topic, "^neu") ~ "MORPH",
            str_detect(topic, "^gen") ~ "GR"
        )
    )

# Step 2: Map each risk gene to all of its GO terms
gene_go_terms <- go_mapping %>%
    filter(ensembl_gene_id %in% asd_risk_genes,
           ontology == "BP") %>%
    # add gene name
    left_join(df_asd_rarevar_zscore %>% 
                  dplyr::select(ensembl_gene_id, gene_name)
    ) %>%
    distinct(gene_name, ensembl_gene_id, go_id)
# N=3 genes are missing (no GO annotations in database): PRR12, GSE1, ZCCHC14


# Step 3: compute per-gene topic overlap scores
gene_topic_scores <- gene_go_terms %>%
    
    # combine gene-term mapping and topic-term mapping
    left_join(topic_vocab, by = "go_id") %>%
    
    # count the total number of GO terms each gene is mapped to
    group_by(gene_name, ensembl_gene_id) %>%
    mutate(total_terms = n_distinct(go_id)) %>%
    
    # count the number of terms in each topic each gene is mapped to
    group_by(gene_name, ensembl_gene_id, topic) %>%
    summarise(
        n_overlap = sum(!is.na(topic)),
        total_terms = dplyr::first(total_terms),
        .groups = "drop"
    ) %>%
    filter(!is.na(topic)) %>%
    
    # calculate the proportion of terms in each topic to the total number of terms
    group_by(ensembl_gene_id) %>%
    mutate(topic_share = n_overlap / sum(n_overlap)) %>%
    slice_max(n_overlap, n = 1, with_ties = FALSE) %>%
    ungroup %>%
    arrange(-n_overlap/total_terms)
```


# --- 2) Manual assignments --- #

## 2.1 Overrides

```{r manual_overrides}
# Manually override the genes whose GO-overlap call (§1.2) differs from the locked
# final assignment, so §1.6 reproduces 2026.06.30_FINAL-gene-topic-assignment.xlsx
# exactly (with CHR relabelled GR). One row per gene; grouped by target topic. Every
# gene in gene_topic_scores whose §1.2 topic already matches the final is left as-is.
manual_overrides <- tribble(
    ~gene_name, ~topic,

    # --- reassign to GR (genomic regulation / RNA processing) ---
    "ADNP",     "GR",   # ChAHP chromatin-remodeling complex (CHD4/HP1)
    "AHDC1",    "GR",
    "ANKRD11",  "GR",   # HDAC3-recruiting chromatin cofactor
    "ARID1B",   "GR",   # BAF (SWI/SNF) chromatin-remodeling subunit
    "CHD7",     "GR",   # chromodomain helicase, SWI/SNF-related (CHD8 family)
    "CIC",      "GR",
    "EEF1A2",   "GR",   # translation elongation factor -> RNA processing / protein homeostasis
    "HIST1H1E", "GR",   # histone H1.4, chromatin compaction
    "KDM6B",    "GR",   # H3K27 demethylase, opposes PRC2
    "MBD5",     "GR",   # methyl-CpG-binding, chromatin
    "PTEN",     "GR",
    "PTPN11",   "GR",
    "SMARCB1",  "GR",   # BAF (SWI/SNF) core subunit, chromatin
    "SPEN",     "GR",
    "TLE3",     "GR",   # Groucho/TLE corepressor
    "TRAF7",    "GR",
    "VEZF1",    "GR",   # broad (non-neural) ZNF transcriptional activator

    # --- reassign to MORPH (neuron development / morphogenesis) ---
    "ATAT1",    "MORPH",
    "BCR",      "MORPH", # Rho-GAP, RAC1/CDC42-mediated axon growth cone morphology
    "BRSK2",    "MORPH", # brain-specific kinase, tau phosphorylation and axonal polarity
    "CAMTA1",   "MORPH",
    "COL4A3BP", "MORPH", # CERT, ceramide transfer, axon myelination / neurite outgrowth
    "DEAF1",    "MORPH",
    "DYNC1H1",  "MORPH",
    "EXT1",     "MORPH",
    "FOXP1",    "MORPH", # forkhead TF, cortical neuron identity
    "MAF",      "MORPH", # bZIP TF, interneuron fate
    "MAP1A",    "MORPH", # microtubule-associated protein, axon cytoskeleton
    "MAPK8IP3", "MORPH",
    "MARK2",    "MORPH", # MAP/microtubule affinity-regulating kinase, neuronal polarity
    "MEF2C",    "MORPH", # TF specifying GABAergic interneuron fate, neuronal differentiation
    "NR3C2",    "MORPH", # mineralocorticoid nuclear receptor TF
    "NR4A2",    "MORPH", # NURR1 nuclear receptor, dopaminergic identity
    "PLXNA1",   "MORPH",
    "PPP2R1A",  "MORPH",
    "PTCH1",    "MORPH", # Hedgehog receptor, neuronal fate specification
    "RALGAPB",  "MORPH", # RAL-GAP, neurite outgrowth and cell polarity
    "RFX3",     "MORPH", # RFX TF, ciliogenesis / neuronal development
    "SOX5",     "MORPH", # SoxD TF, deep-layer neuron fate
    "SPTBN1",   "MORPH", # spectrin βII, axon initial segment and axonal transport
    "TAOK1",    "MORPH", # TAO kinase, microtubule stability and dendritic arborization
    "TM9SF4",   "MORPH", # nine-TM protein, membrane homeostasis
    "YWHAG",    "MORPH", # 14-3-3γ, tau phosphorylation and neurite outgrowth
    "ZBTB20",   "MORPH", # BTB/ZNF TF, hippocampal / cortical development
    "ZFHX3",    "MORPH", # ZNF-homeobox TF, neuronal differentiation

    # --- reassign to SYN (synaptic transmission / ion transport) ---
    "AP2S1",    "SYN",   # AP-2 complex σ subunit, clathrin-mediated synaptic vesicle endocytosis
    "ARF3",     "SYN",   # ADP-ribosylation factor, Golgi vesicle trafficking
    "CACNA1A",  "SYN",
    "CAMK2A",   "SYN",
    "CLTC",     "SYN",
    "CSNK1E",   "SYN",   # CK1ε, synaptic vesicle phosphorylation / circadian
    "DNM1",     "SYN",
    "GNAI1",    "SYN",   # Gαi1, inhibitory GPCR signaling at synapse
    "GNB1",     "SYN",   # Gβ subunit, synaptic GPCR signaling
    "NBEA",     "SYN",
    "PPP2CA",   "SYN",
    "RAB14",    "SYN",   # Rab GTPase, endosomal vesicle trafficking
    "SYNGAP1",  "SYN"
)
gene_topic_update <- gene_topic_scores %>%
    rows_upsert(manual_overrides, by = "gene_name")
```

## 2.2 Missing genes

```{r missing_genes}
# Check for missing genes
asd_risk_genes <- df_asd_rarevar_zscore %>% filter(FDR < 0.001 & !FLAG) %>% pull(gene_name)
missing_genes <- setdiff(asd_risk_genes, gene_topic_update$gene_name)

# Manually map missing genes
manual_assignments <- tribble(
    ~gene_name, ~topic,
    "TCF20",   "MORPH",
    "EBF3",    "MORPH",
    "MED13L",  "GR",
    "SRCAP",   "GR",
    "ZNF292",  "GR",
    "ZMYM2",   "GR",
    "RFX7",    "GR",
    "PRR12",   "GR",
    "ZNF865",  "GR",
    "SETBP1",  "GR",
    "HIVEP2",  "MORPH", # large ZNF TF, neurodevelopmental
    "DHDDS",   "GR",   # dolichol synthesis / N-glycosylation → protein homeostasis
    "LMTK3",   "SYN",   # reassigned: AMPA receptor recycling at synapse
    "PHF12",   "GR",
    "GSE1",    "GR",
    "NACC1",   "GR",
    "HSF2",    "MORPH", # heat-shock TF, cortical development / migration
    "ZNF644",  "GR",
    "ZCCHC14", "GR",
    "CXXC1",   "GR",
    "MED13",   "GR",

    "CHAMP1",  "GR",
    "DIP2C",   "GR",
    "EMC7",    "GR",   # ER membrane protein complex / protein biogenesis → protein homeostasis
    "MKX",     "MORPH", # Mohawk homeobox TF (developmental identity factor)
    "WDFY3",   "MORPH", # reassigned: axonal autophagy, neurodevelopmental phenotype
    "YLPM1",   "GR",
    "ZBTB34",  "GR",
)
gene_topic_final <- gene_topic_update %>%
    dplyr::select(gene_name, topic) %>%
    rows_upsert(manual_assignments, by = "gene_name")
```


# --- 3) Export

## 3.1 Format (add TADA stats)

```{r add_HotNet}
# Add TADA stats
df_gene_assignments <- gene_topic_final %>%

    # TADA stats
    left_join(
        df_asd_rarevar_zscore %>%
            dplyr::select(gene_name, FDR, BF)
    ) %>%

    # Final format
    dplyr::select(gene_name, BF, FDR, GO_topic = topic) %>%
    arrange(-BF)
```

## 3.2 Export

```{r export_table}
# Export as table for supplement
df_gene_assignments %>% 
    write_csv(file.path(notebook_out_dir, "02_02_gene-topic-assignment.csv"))
```



# DEPR

## GO-similarity clustering of all genes into 3 groups (instead of GO overlap)

Exploratory comparison: instead of assigning genes to topics by GO-vocabulary overlap
(section 1), cluster all genes directly by pairwise GO cosine similarity (the section-2
approach) and cut the tree into k = 3 groups. The three data-driven clusters are then
mapped to SYN/GR/MORPH by majority overlap with the section-1 labels, and the two
assignments are compared.

## 1 GO similarity matrix for all genes

```{r all_gene_go_similarity}
# All section-1 assigned genes
all_genes_k3 <- df_gene_assignments$gene_name

gene_go_all3 <- go_mapping_full %>%
    filter(ensembl_gene_id %in% {
        df_asd_rarevar_zscore %>%
            filter(gene_name %in% all_genes_k3) %>%
            pull(ensembl_gene_id)
    }) %>%
    left_join(df_asd_rarevar_zscore %>% dplyr::select(ensembl_gene_id, gene_name),
              by = "ensembl_gene_id") %>%
    distinct(gene_name, go_id)

go_mat_all3 <- gene_go_all3 %>%
    mutate(present = 1L) %>%
    pivot_wider(id_cols = gene_name, names_from = go_id,
                values_from = present, values_fill = 0L) %>%
    column_to_rownames("gene_name") %>%
    as.matrix()

sim_mat_all3 <- cos_sim_genes(go_mat_all3)

# Genes with no GO annotation: included with zero similarity (placed by variance minimization)
genes_no_go3 <- all_genes_k3[!all_genes_k3 %in% rownames(sim_mat_all3)]
if (length(genes_no_go3) > 0) {
    message("No GO data: ", paste(genes_no_go3, collapse = ", "))
    n_existing <- nrow(sim_mat_all3)
    n_new      <- length(genes_no_go3)
    zero_rows  <- matrix(0, nrow = n_new, ncol = n_existing,
                         dimnames = list(genes_no_go3, rownames(sim_mat_all3)))
    zero_cols  <- matrix(0, nrow = n_existing + n_new, ncol = n_new,
                         dimnames = list(c(rownames(sim_mat_all3), genes_no_go3), genes_no_go3))
    sim_mat_all3 <- cbind(rbind(sim_mat_all3, zero_rows), zero_cols)
    diag(sim_mat_all3) <- 1
}

cat("GO similarity matrix:", nrow(sim_mat_all3), "×", ncol(sim_mat_all3), "\n")
cat("Genes with no GO data:", length(genes_no_go3), "\n")
```

## 2 Hierarchical clustering into 3 groups

```{r all_gene_cluster_k3}
# Ward.D2 hierarchical clustering on GO cosine distance, cut at k = 3
hc_all3       <- hclust(as.dist(1 - sim_mat_all3), method = "ward.D2")
cluster_vec_3 <- cutree(hc_all3, k = 3)

cat("Raw cluster sizes (k = 3):\n")
print(table(cluster_vec_3))
```

## 3 Map clusters to topic labels & compare to section 1

```{r compare_k3_to_section1}
# Tibble of cluster ID + section-1 GO topic per gene
df_k3 <- tibble(
    gene_name      = names(cluster_vec_3),
    sim_cluster    = cluster_vec_3
) %>%
    left_join(df_gene_assignments %>% dplyr::select(gene_name, GO_topic), by = "gene_name")

# Map each data-driven cluster to the topic it most overlaps in section 1
cluster_to_topic <- df_k3 %>%
    filter(!is.na(GO_topic)) %>%
    count(sim_cluster, GO_topic) %>%
    group_by(sim_cluster) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(sim_cluster, sim_topic = GO_topic)

df_k3 <- df_k3 %>%
    left_join(cluster_to_topic, by = "sim_cluster") %>%
    mutate(agree = sim_topic == GO_topic)

# Confusion matrix: section-1 topic (rows) vs GO-similarity cluster topic (cols)
cat("Section-1 topic (rows) × GO-similarity k=3 topic (cols):\n")
print(table(section1 = df_k3$GO_topic, sim_k3 = df_k3$sim_topic))

cat("\nOverall agreement:",
    scales::percent(mean(df_k3$agree, na.rm = TRUE), accuracy = 0.1), "\n")

# Genes where the two approaches disagree
df_k3 %>%
    filter(!agree) %>%
    dplyr::select(gene_name, section1_topic = GO_topic, sim_k3_topic = sim_topic) %>%
    arrange(section1_topic, sim_k3_topic)

# All three clusters are GR dominated
print(table(sim_cluster = df_k3$sim_cluster, section1 = df_k3$GO_topic))
```

We confirmed that unsupervised GO-similarity clustering of all genes into three groups did not recover the topic structure (all three clusters were GR-dominated; Supplementary Fig X), motivating the use of the topic-anchored vocabulary-overlap method for the primary assignment.
