---
title: "BrainSpan developmental trajectories"
description: "Developmental expression trajectories and prenatal bias of ASD risk genes by GO topic (BrainSpan; Kang et al. 2011 stages, Satterstrom et al. 2020 filtering)"
date: 2026-07-09
format: html
---

# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "02_module_enrichments")
```

```{r env}
# ASD rare variant stats
df_asd_rarevar_zscore <- readRDS(file.path(analysis_dir, "00_data_format", "00_01_ASD-rarevar-stats.RDS"))

# Gene–topic assignments (from 02_02)
df_gene_assignments   <- read_csv(file.path(notebook_out_dir, "02_02_gene-topic-assignment.csv"))
```


# --- 1) Load and process BrainSpan data --- #

## 1.1 Load BrainSpan data

```{r load_brainspan}
# Data source: http://www.brainspan.org/static/download.html
#   → Developmental Transcriptome → "RNA-Seq Gencode v10 summarized to genes"
# Satterstrom et al. (2020) used this dataset (Li et al., 2018)
brainspan_dir <- file.path(data_dir, "BrainSpan")

# gene x sample expression matrix
df_bs_expr    <- read_csv(file.path(brainspan_dir, "expression_matrix.csv"),
                          col_names = FALSE, show_col_types = FALSE)

# row (gene) metadata
df_bs_genes   <- read_csv(file.path(brainspan_dir, "rows_metadata.csv"),
                          show_col_types = FALSE) %>% clean_names()

# column (sample) metadata
df_bs_samples <- read_csv(file.path(brainspan_dir, "columns_metadata.csv"),
                          show_col_types = FALSE) %>% clean_names()

# Convert expression tibble to matrix with gene name rows & sample number columns
mat_expr_full <- as.matrix(df_bs_expr[, -1])
rownames(mat_expr_full) <- df_bs_genes$gene_symbol
colnames(mat_expr_full) <- as.character(df_bs_samples$column_num)
rm(df_bs_expr)

cat("BrainSpan:", nrow(mat_expr_full), "genes ×", ncol(mat_expr_full), "samples loaded\n")
```

## 1.2 Process samples

Helper functions to convert age to pcw; align with Kang et al., 2011 stages

```{r dev_stage_helpers}
# Convert BrainSpan age strings ("8 pcw", "4 mos", "2 yrs") to post-conceptional
# weeks (pcw), assuming full-term birth at 40 pcw
# 52.18 weeks/year
parse_bs_age_pcw <- function(age_str) {
    num  <- as.numeric(str_extract(age_str, "[0-9.]+"))
    unit <- str_extract(age_str, "[a-z]+")
    case_when(
        unit == "pcw" ~ num,
        unit == "mos" ~ 40 + num * (52.18 / 12),
        unit == "yrs" ~ 40 + num * 52.18,
        TRUE          ~ NA_real_
    )
}

# Convert pcw to developmental stage
assign_dev_stage <- function(pcw) {
    factor(case_when(
        pcw <   8                      ~ "Embryonic",
        pcw <  13                      ~ "Early fetal",
        pcw <  19                      ~ "Early mid-fetal",
        pcw <  24                      ~ "Late mid-fetal",
        pcw <  38                      ~ "Late fetal",
        pcw < (40 +  6 * 52.18 / 12)  ~ "Neonatal and early infancy",
        pcw < (40 + 12 * 52.18 / 12)  ~ "Late infancy",
        pcw < (40 +  6 * 52.18)       ~ "Early childhood",
        pcw < (40 + 12 * 52.18)       ~ "Middle and late childhood",
        pcw < (40 + 20 * 52.18)       ~ "Adolescence",
        pcw < (40 + 40 * 52.18)       ~ "Young adulthood",
        pcw < (40 + 60 * 52.18)       ~ "Middle adulthood",
        TRUE                          ~ "Late adulthood"
    ), levels = stage_levels) # stage_levels are defined in setup.R because this is used across scripts
}
```

Parse age into pcw, assign developmental stage, & filter

```{r brainspan_sample_filt}
# 11 neocortical regions as used by Satterstrom et al.
neocortex_regions <- c(
    "DFC", "VFC", "MFC", "OFC",
    "M1C", "S1C", "A1C",
    "IPC", "STC", "ITC",
    "V1C"
)

df_samples_filt <- df_bs_samples %>%
    
    # convert age to pcw and assign developmental stage
    mutate(
        pcw         = parse_bs_age_pcw(age),
        dev_stage   = assign_dev_stage(pcw),
        is_prenatal = pcw < 40
    ) %>%
    
    # filter for neocortical regions
    filter(structure_acronym %in% neocortex_regions) %>%
    
    # indicate if a sample is prenatal or postnatal
    mutate(
        epoch        = if_else(is_prenatal, "prenatal", "postnatal"),
        region_epoch = paste(structure_acronym, epoch, sep = "_") # for gene filtering
    )

cat(nrow(df_samples_filt), "of", nrow(df_bs_samples), "samples from neocortex\n")
print(count(df_samples_filt, dev_stage, epoch))
```

## 1.3 Filter genes

Criteria match Satterstrom et al. 2020

```{r brainspan_gene_filter}
# Retain genes present at RPKM >= 0.5 in >=80% of samples from at least one
# neocortical region × epoch combination (matches Satterstrom et al. threshold)
sample_cols         <- as.character(df_samples_filt$column_num)
mat_rpkm            <- mat_expr_full[, sample_cols]
region_epoch_groups <- unique(df_samples_filt$region_epoch)

pct_expressed <- sapply(region_epoch_groups, function(re) {
    cols <- df_samples_filt$column_num[df_samples_filt$region_epoch == re]
    rowMeans(mat_rpkm[, as.character(cols), drop = FALSE] >= 0.5)
})
expressed_genes <- rownames(pct_expressed)[apply(pct_expressed, 1, max) >= 0.8]
cat(length(expressed_genes), "genes pass expression threshold\n") # N = 18551

# Convert to log2(RPKM + 1) as in Satterstrom et al.
mat_log2 <- log2(mat_rpkm[expressed_genes, ] + 1)

# Z-score each gene (row) across all samples: mean 0, SD 1 across full trajectory
mat_z <- t(scale(t(mat_log2)))
```


# --- 2) Topic gene list developmental trajectories --- #

## 2.1 Helper function: compute relative expression values at each dev stage

```{r helper_function_traj}
# Compute per-sample mean z-score for a named gene list, returning a long tibble
# with columns: GO_topic, column_num, mean_z, log10_pcw, dev_stage.
# background_means is subtracted per sample so trajectories are centered around 0,
# showing expression relative to the overall background at each timepoint.
compute_traj_samples <- function(gene_lists, 
                                 mat_z, 
                                 df_samples_filt,
                                 background_means = NULL) {
    map_dfr(names(gene_lists), function(label) {
        genes <- intersect(gene_lists[[label]], rownames(mat_z))
        if (length(genes) < 5) {
            warning(label, ": fewer than 5 genes in BrainSpan — skipping")
            return(NULL)
        }
        sample_means <- colMeans(mat_z[genes, , drop = FALSE], na.rm = TRUE)
        if (!is.null(background_means))
            sample_means <- sample_means - background_means[names(sample_means)]
        tibble(GO_topic = label, column_num = names(sample_means), mean_z = sample_means)
    }) %>%
        left_join(
            df_samples_filt %>%
                mutate(column_num = as.character(column_num)) %>%
                select(column_num, pcw, dev_stage),
            by = "column_num"
        ) %>%
        mutate(log10_pcw = log10(pcw))
}
```

## 2.2 Compute expression trajectories for ASD genes in each GO topic

```{r compute_dev_traj}
# Convert df_gene_assignments to list
gene_topic_lists <- df_gene_assignments %>%
    split(.$GO_topic) %>%
    map(~ pull(.x, gene_name) %>% unique())

# Calculate the mean expression for each sample (background)
# mat_z defined in section 1.3
background_sample_means <- colMeans(mat_z, na.rm = TRUE)

# Compute relative expression values at each dev stage
df_traj_topic <- compute_traj_samples(
    gene_topic_lists, 
    mat_z, 
    df_samples_filt,
    background_means = background_sample_means
) %>%
    group_by(pcw) %>% 
    mutate(mean_z = as.numeric(scale(mean_z))) %>% # z-score within each timepoint
    ungroup() %>%
    mutate(GO_topic = factor(GO_topic, levels = c("SYN", "GR", "MORPH")))
```

## 2.3 Export trajectories

For Fig 2f

```{r export_traj}
saveRDS(df_traj_topic, file = file.path(notebook_out_dir, "02_04_brainspan-ASD-gene-trajectories.RDS"))
```


# --- 3) Calculate pre- vs post-natal t-statistic across genes --- #


## 3.1 Data prep

```{r tstat_data_prep}
# Donor-level averaged expression: collapse all regions within each donor to a single
# mean to avoid collinearity between donor identity and prenatal status.

# Pull donor IDs from filtered data
donor_ids <- df_samples_filt %>% distinct(donor_id) %>% pull(donor_id)

# Calculate the donor-level average: collapse all regions within each donor
# to a mean to avoid collinearity between donor identitiy and prenatal status
# outputs gene x donor matrix
mat_donor_avg <- sapply(as.character(donor_ids), function(d) {
    cols <- df_samples_filt %>%
        filter(donor_id == d) %>%
        mutate(column_num = as.character(column_num)) %>%
        pull(column_num)
    rowMeans(mat_log2[, cols, drop = FALSE], na.rm = TRUE)
})

# Extract donor-level metadata from df_samples_filt
df_donor_meta <- df_samples_filt %>%
    distinct(donor_id, gender, is_prenatal) %>%
    mutate(donor_id = as.character(donor_id)) %>%
    slice(match(colnames(mat_donor_avg), donor_id))

cat(ncol(mat_donor_avg), "donors:",
    sum(df_donor_meta$is_prenatal), "prenatal,",
    sum(!df_donor_meta$is_prenatal), "postnatal\n")
```

## 3.2 Pre- vs post-natal t-statistic

For each gene, calculate the t-statistic of pre- vs post-natal expression

```{r calculate_tstat}
# Prenatal t-statistic: sign convention follows Satterstrom et al. Fig 4C
# negative t = higher prenatal expression (positively biased prenatal)
# Include gender as covariate (same as Satterstrom et al)
compute_prenatal_tstat <- function(gene_expr_vec) {
    df_fit <- data.frame(
        expr        = gene_expr_vec,
        is_prenatal = df_donor_meta$is_prenatal,
        gender      = df_donor_meta$gender
    )
    fit <- tryCatch(lm(expr ~ is_prenatal + gender, data = df_fit), error = function(e) NULL)
    if (is.null(fit)) return(NA_real_)
    cs <- coef(summary(fit))
    if (!"is_prenatalTRUE" %in% rownames(cs)) return(NA_real_)
    -cs["is_prenatalTRUE", "t value"]
}

# Prenatal t-statistic per gene: negative = higher prenatal, positive = higher postnatal.
prenatal_tstats <- apply(mat_donor_avg, 1, compute_prenatal_tstat)
df_tstats <- tibble(
    gene_name = names(prenatal_tstats),
    prenatal_tstat = prenatal_tstats
) %>% 
    
    # add gene assignment info
    left_join(df_gene_assignments) %>% 
    mutate(GO_topic = ifelse(is.na(GO_topic), "Background", GO_topic)) %>% 
    mutate(GO_topic = factor(GO_topic, levels = c("Background", "SYN", "MORPH", "GR")))
```

## 3.3 Compare GO topic t-stat distribution to background

For each GO topic, test whether its prenatal t-statistic distribution differs from the
non-ASD background (all genes not assigned to a topic), with a two-sided Wilcoxon
rank-sum test and Cliff's delta as the effect size.

```{r tstat_comparison}
# Background = non-ASD genes: every gene with a t-statistic that is not an assigned ASD
# risk gene. With ~18K background genes the Wilcoxon test is highly powered, so Cliff's
# delta is the primary quantity — bounded [-1, 1]; > 0 = postnatal-biased vs background
# (prenatal_tstat: negative = higher prenatal, positive = higher postnatal).
background_tstats <- df_tstats %>%
    filter(GO_topic == "Background", !is.na(prenatal_tstat)) %>%
    pull(prenatal_tstat)

# Wilcoxon rank-sum + Cliff's delta for one gene set vs the background
compare_to_background <- function(topic_tstats, bg = background_tstats) {
    wt <- wilcox.test(topic_tstats, bg, alternative = "two.sided")
    tibble(
        n_genes      = length(topic_tstats),
        median_tstat = median(topic_tstats),
        # Cliff's delta = 2U / (n_topic * n_bg) - 1 = P(topic > bg) - P(topic < bg)
        cliffs_delta = 2 * as.numeric(wt$statistic) / (length(topic_tstats) * length(bg)) - 1,
        p_wilcox     = wt$p.value
    )
}

df_tstat_comparison <- df_tstats %>%
    filter(GO_topic != "Background" & !is.na(prenatal_tstat)) %>%
    group_by(GO_topic) %>%
    group_modify(~ compare_to_background(.x$prenatal_tstat)) %>%
    ungroup() %>%
    mutate(
        p_adj_wilcox = p.adjust(p_wilcox, method = "BH"),
        sig_wilcox   = case_when(
            p_adj_wilcox < 0.001 ~ "***",
            p_adj_wilcox < 0.01  ~ "**",
            p_adj_wilcox < 0.05  ~ "*",
            TRUE                 ~ "ns"
        ),
        GO_topic = factor(GO_topic, levels = c("SYN", "GR", "MORPH"))
    ) %>%
    arrange(GO_topic)

df_tstat_comparison %>%
    mutate(across(where(is.double), ~ round(.x, 4))) %>%
    print()
```

## 3.4 Export t-stats

```{r export_tstats}
tstat_data <- list(
    "gene_tstats" = df_tstats,
    "topic_vs_background" = df_tstat_comparison
)
saveRDS(tstat_data, file.path(notebook_out_dir, "02_04_brainspan-prenatal-tstats.RDS"))
```






# DEPR

## Within-topic ASD risk gene trajectories across TADA FDR thresholds

```{r data_asd_traj}
# ASD risk gene developmental trajectories for four TADA FDR thresholds.
# Approach: subtract genome-wide per-sample mean z-score first (so y = 0 represents
# the genome-wide background at each timepoint), then within-pcw z-score the topics
# only. Background t-statistics: all expressed genes not in any topic gene list at
# the given FDR threshold (genes can belong to multiple topics; no exclusivity filter).

asd_fdr_thresholds <- c(0.001, 0.01, 0.05, 0.1)

asd_data_list <- map(
    setNames(asd_fdr_thresholds, paste0("fdr_", asd_fdr_thresholds)),
    function(fdr_thresh) {
        asd_genes <- df_asd_rarevar_zscore %>%
            filter(FDR < fdr_thresh & !FLAG) %>%
            pull(gene_name) %>%
            unique()
        
        topic_lists_asd <- df_topic_genes %>%
            filter(hgnc_symbol %in% asd_genes) %>%
            split(.$topic_label) %>%
            map(~ pull(.x, hgnc_symbol) %>% unique())
        
        bg_tstats <- df_tstats %>%
            filter(!hgnc_symbol %in% unlist(topic_lists_asd),
                   !is.na(prenatal_tstat)) %>%
            pull(prenatal_tstat)
        
        background_sample_means <- colMeans(mat_z, na.rm = TRUE)
        
        df_traj_topic_raw <- compute_traj_samples(
            topic_lists_asd, mat_z, df_samples_filt,
            background_means = background_sample_means
        ) %>%
            group_by(pcw) %>%
            mutate(mean_z = as.numeric(scale(mean_z))) %>%
            ungroup()
        
        list(
            df_traj_topic  = df_traj_topic_raw %>%
                mutate(set_label = factor(set_label, levels = c("SYN", "CHR", "MORPH"))),
            df_traj_background = NULL,
            background_tstats  = bg_tstats,
            topic_lists        = topic_lists_asd,
            n_asd              = length(asd_genes)
        )
    }
)

# Gene count summary across thresholds
walk2(asd_data_list, names(asd_data_list), function(d, nm) {
    cat(nm, "— N ASD risk genes:", d$n_asd, "\n")
    cat("  Topic N:", paste(names(d$topic_lists),
                            map_int(d$topic_lists, length),
                            sep = "=", collapse = ", "), "\n")
    cat("  Background N:", length(d$background_tstats), "\n")
})
```

