---
title: "ASD-associated module GO and cell-type enrichments"
description: "Run GO ORA on all significant modules; use topic modeling to synthesize biological themes"
date: 2026-07-07
format: html
---


# --- 0) Setup --- #

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Source setup script
project_dir <- "~/Documents/Postdoc/Projects/ASD-rarevar-annot/"
source(paste0(project_dir, "code/setup.R"))

# Set directory for analysis outputs
notebook_out_dir <- file.path(analysis_dir, "02_module_enrichments")
```

Load data

```{r env}
# Gene network assignments
df_network_data <- readRDS(file.path(analysis_dir, "00_data_format", "00_00_gene-networks-all.RDS"))

# Significant ASD modules
df_stepwise_res <- readRDS(
    file.path(analysis_dir, "01_module_identification", "01_00_stepwise-linear-res-step1p0.001-step2p0.001.RDS")
)$lm_results %>% 
    mutate(module_label = factor(module_label, levels = unique(.$module_label)))
```


# --- 1) GO enrichment --- #

## 1.1 Load GO terms

```{r load_GO}
# GO term description to ID mapping
go_terms <- AnnotationDbi::select(
    GO.db,
    keys = keys(GO.db),
    columns = c("TERM", "ONTOLOGY"),
    keytype = "GOID"
) %>% 
    as_tibble() %>% 
    clean_names %>% 
    dplyr::rename(go_id = goid)

# GO ID to gene mapping
go_mapping <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = keys(org.Hs.eg.db, keytype = "ENSEMBL"),
    columns = c("GO", "ONTOLOGY"),
    keytype = "ENSEMBL"
) %>% 
    as_tibble() %>% 
    clean_names %>% 
    dplyr::rename(go_id = go, ensembl_gene_id = ensembl) %>% 
    dplyr::select(-evidence)

# Identify GO database background
go_background <- go_mapping %>% pull(ensembl_gene_id) %>% unique
```

## 1.2 Prep GO data

```{r prep_GO_data}
## To save time, precompute gene lists ##

# Precompute study & module gene sets
df_network_unnested <- df_network_data %>%
    tidyr::unnest(data)

# Study-level genes (for enrichment backgrounds)
study_genes_list <- df_network_unnested %>%
    group_by(study) %>% 
    summarise(study_genes = list(unique(ensembl_gene_id)), .groups = "drop") %>%
    deframe()  # named list: study -> genes

# Module-level genes
module_genes_list <- df_network_unnested %>%
    group_by(study, module) %>%
    summarise(module_genes = list(unique(ensembl_gene_id)), .groups = "drop") %>% 
    filter(module %in% df_stepwise_res$module)

# GO --> genes list (filtered)
go_genes_list <- go_mapping %>%
    group_by(go_id) %>%
    summarise(genes = list(unique(ensembl_gene_id)), .groups = "drop") %>%
    filter(lengths(genes) >= 5 & lengths(genes) <= 500) %>% # remove GO terms with fewer than 5 genes or greater than 500 genes (removes broad, non-informative terms)
    deframe()
```

## 1.3 Run GO ORA

```{r run_GO_ORA}
## Run ORA on all combinations of modules and GO paths

results <- list()

for(i in seq_len(nrow(module_genes_list))) {
    study <- module_genes_list$study[i]
    module <- module_genes_list$module[[i]]
    module_genes <- module_genes_list$module_genes[[i]]
    
    print(paste0("Running GO enrichment on ", module))
    
    # Only GO terms that overlap > 1
    go_overlap <- names(go_genes_list)[map_lgl(go_genes_list, ~ length(intersect(.x, module_genes)) > 1)]
    
    for(go_id in go_overlap) {
        go_genes <- go_genes_list[[go_id]]
        study_genes <- study_genes_list[[study]]
        
        results[[length(results)+1]] <- ORA_fisher(
            gene_list1 = module_genes,
            background_list1 = study_genes,
            gene_list2 = go_genes,
            background_list2 = go_background,
            alternative = "greater"
        ) %>% 
            mutate(study = study, module = module, go_id = go_id)
    }
}

df_go_res <- bind_rows(results)

# Format results
df_go_res <- df_go_res %>% 
    left_join(go_terms) %>% 
    dplyr::select(study, module, go_id, term, ontology, everything()) %>% 
    group_by(study, module) %>% 
    mutate(p_adj = p.adjust(p_value, method = "BH"), .before = n_overlap) %>% 
    mutate(module = factor(module, levels = df_stepwise_res$module)) %>% 
    arrange(module, p_value) %>% 
    dplyr::rename(n_module = n_list1, n_pathway = n_list2) %>% 
    ungroup %>% 
    mutate(pct_module = n_overlap / n_module * 100)
```

## 1.4 Export results as object and table for supplement

```{r save_GO_res}
# Save output as RDS file
saveRDS(
    df_go_res,
    file = file.path(notebook_out_dir, "02_00_sig-mod-GO-enrichments.RDS")
)
# Write significant results to Excel
write_csv(
    df_go_res %>% filter(p_value < 0.05),
    path = file.path(notebook_out_dir, "02_00_sig-mod-GO-enrichments-p0.05.csv")
)
```


# --- 2) Topic modeling --- #

Run topic modeling on all significant GO terms to extract common themes among enrichments, and quantify module relationship to each term. \n NOTE: We have too few documents (modules, n = 28) to run searchK and identify optimal number of modules. We select K = 3 based on qualitiative evaluation of biological themes

## 2.1 Prep sparse module-word matrix

```{r data_prep}
# Unnest tokens to get GO terms within each module (group)
df_go_tokens <- df_go_res %>% 
    dplyr::filter(!is.na(term)) %>% 
    group_by(module) %>% 
    # if any term in the module has p_adj < 0.05, keep only those.
    # else, keep terms where p_value < 0.05
    dplyr::filter(
        if (any(p_adj < 0.05)) {
            p_adj < 0.05
        } else {
            p_value < 0.05
        }
    ) %>% 
    ungroup() %>% 
    unnest_tokens(word, term)

## Create a sparse matrix using only meaningful words that appear at least twice within module
generic_terms <- c("positive", "negative", "regulation", "response", "protein", "activity",
                   "pathway", "process", "involved", "cell", 
                   "binding", "complex", "signaling", "localization", "organization", 
                   "component", "system", "level", "activation", "production",
                   "kinase", "membrane", "dependent", "factor", "substrate", "mediated", "type", "formation")

# 1. Identify which modules drop out at n > 2
modules_all <- unique(df_go_tokens$module)
modules_at_n2 <- df_go_tokens %>%
    dplyr::count(module, word) %>%
    anti_join(stop_words, by = "word") %>% 
    dplyr::filter(!(word %in% generic_terms)) %>% 
    dplyr::filter(n > 2) %>%
    pull(module) %>%
    unique()

missing_modules <- setdiff(modules_all, modules_at_n2)

# 2. Extract the top words for those specific missing modules
whitelist_words <- df_go_tokens %>%
    filter(module %in% missing_modules) %>%
    dplyr::count(module, word) %>%
    anti_join(stop_words, by = "word") %>%
    dplyr::filter(!(word %in% generic_terms)) %>%
    group_by(module) %>%
    slice_max(n, n = 3) %>% # Keep top 3 words per missing module
    pull(word) %>%
    unique()

# 3. Create the sparse matrix with n > 2 OR Whitelist
go_sparse <- df_go_tokens %>%
    dplyr::count(module, word) %>%
    anti_join(stop_words, by = "word") %>% 
    dplyr::filter(!(word %in% generic_terms)) %>% 
    # Logic: Keep if frequency > 2 OR if it's a vital word for the missing modules
    dplyr::filter(n > 2 | word %in% whitelist_words) %>%
    cast_sparse(module, word, n)
```

## 2.2 Select k

```{r select_k, fig.height = 4, fig.width = 7}
k_range <- 2:10
k_models <- map(k_range, ~ stm(go_sparse, K = .x, verbose = FALSE, seed = 234))

df_k_metrics <- map2_dfr(k_range, k_models, function(k, m) {
    tibble(
        K      = k,
        semcoh = mean(semanticCoherence(m, go_sparse)),
        exclus = mean(exclusivity(m))
    )
})
```

## 2.3 Select final topic model

```{r run_topic_model}
topic_model <- k_models[[2]]

# Save fitted topic model for downstream analyses & figures
saveRDS(topic_model, file = file.path(notebook_out_dir, "02_00_topic-model-K3.RDS"))
```

## 2.4 Extract topic model metrics

Beta = word-topic probability

```{r extract_beta}
# Beta = overall probability that a word belongs to a given topic

# Quantify word-topic significance
df_beta <- tidytext::tidy(topic_model, matrix = "beta") %>%
    group_by(topic) %>%
    mutate(
        # Z-score for how much more likely this word is in this topic vs others
        z_score = (beta - mean(beta)) / sd(beta),
        p_value = pnorm(z_score, lower.tail = FALSE),
        p_adj = p.adjust(p_value, method = "BH")
    ) %>%
    ungroup()

# Save for figure
saveRDS(df_beta, file = file.path(notebook_out_dir, "02_00_topic-word-beta.RDS"))
```

Gamma = module-topic probability

```{r extract_gamma}
# Quantify module-topic probabilities & label topics
df_gamma <- tidytext::tidy(
    topic_model,
    matrix = "gamma",
    document_names = rownames(go_sparse)
) %>% 
    dplyr::rename(module = document) %>% 
    
    # Calculate z-score and p-value of module association with a given topic compared to other modules
    group_by(topic) %>% 
    mutate(
        z_score = (gamma - mean(gamma)) / sd(gamma),
        p_value = pnorm(z_score, lower.tail = FALSE),
        p_adj = p.adjust(p_value, method = "BH")
    ) %>%
    ungroup() %>% 
    
    # rename modules based on word associations
    mutate(topic = case_when(
        topic == 1 ~ "genomic regulation / \nRNA processing",
        topic == 2 ~ "synaptic transmission / \nion transport",
        topic == 3 ~ "neuron development / \nmorphogenesis"
    ),
    topic = factor(topic, 
                   levels = c("genomic regulation / \nRNA processing", 
                              "neuron development / \nmorphogenesis",
                              "synaptic transmission / \nion transport")
    ),
    topic_label = case_when(
        topic == "synaptic transmission / \nion transport" ~ "SYN",
        topic == "neuron development / \nmorphogenesis" ~ "MORPH",
        topic == "genomic regulation / \nRNA processing" ~ "GR"
    ),
    topic_color = case_when(
        topic_label == "SYN" ~ "#4a2377",
        topic_label == "MORPH" ~ "#0d7d87",
        topic_label == "GR" ~ "#ea801c"
    )
    ) %>% 
    
    # Identify if the topic is the top topic for a given module
    group_by(module) %>% 
    mutate(top_topic = ifelse(gamma == max(gamma), "yes", "no")) %>% 
    ungroup

# Save for downstream analyses & figures
saveRDS(df_gamma, file = file.path(notebook_out_dir, "02_00_topic-module-gamma.RDS"))
```

## 2.5 Assignment stability across random restarts

stm defaults to `init.type = "Spectral"`, which is deterministic 
(the `seed` argument above does not vary the fit, so it can't be used to assess stability) 
With only 28 documents the topic solution is weakly identified, 
and a single fit can assign a module with high posterior probability but low support across restarts 
(M56-xx is the example: gamma margin 0.93, restart support 0.18). 
We therefore refit under random initialization, 
match each restart's topics back to the spectral solution, and assign
modules by consensus rather than by a single fit.

```{r topic_restart_consensus}
# Number of random restarts; the estimates below are Monte Carlo, so this is the
# precision/runtime trade-off (100 restarts ~ few minutes for K = 3)
n_restarts <- 1000
set.seed(1221)
restart_seeds <- sample(1e7, n_restarts)

# Topic order in the spectral fit, matching the labels assigned in 2.4
topic_labels <- c("GR", "SYN", "MORPH")

# All permutations of 3 topics; topic identity is arbitrary across fits, so each
# restart is matched back to the spectral solution before its labels can be compared
topic_perms <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))

get_beta_matrix <- function(model) {
    b <- exp(model$beta$logbeta[[1]])
    colnames(b) <- colnames(go_sparse)
    b
}
get_gamma_matrix <- function(model) {
    g <- tidytext::tidy(model, matrix = "gamma", document_names = rownames(go_sparse)) %>%
        pivot_wider(names_from = topic, values_from = gamma) %>%
        as.data.frame()
    rownames(g) <- g$document
    as.matrix(g[, as.character(seq_along(topic_labels)), drop = FALSE])
}

beta_spectral <- get_beta_matrix(topic_model)

# Refit under random initialisation and realign each fit to the spectral topics.
# Matching maximizes total word-topic correlation over permutations, which forces a
# one-to-one mapping (greedy per-topic argmax can map two topics onto the same target)
l_restarts <- purrr::map(restart_seeds, function(s) {
    fit <- stm(go_sparse, K = 3, verbose = FALSE, seed = s, init.type = "Random")
    cor_mat <- cor(t(beta_spectral), t(get_beta_matrix(fit)))
    perm <- topic_perms[[which.max(purrr::map_dbl(topic_perms, ~ sum(cor_mat[cbind(seq_along(topic_labels), .x)])))]]
    list(
        gamma = get_gamma_matrix(fit)[, perm, drop = FALSE],
        # weakest matched topic correlation: flags restarts that found a genuinely
        # different structure rather than a relabelling of the same one
        min_matched_r = min(cor_mat[cbind(seq_along(topic_labels), perm)])
    )
})

cat("Restarts matching all topics at r > 0.5:",
    sum(purrr::map_dbl(l_restarts, "min_matched_r") > 0.5), "/", n_restarts, "\n")

# Proportion of restarts assigning each module to each topic. Like gamma, this is a
# module x topic matrix on [0, 1] summing to 1 per module, so it is a drop-in
# replacement for gamma in figures -- but it measures agreement across fits rather
# than posterior probability within one fit
df_assignment_prop <- purrr::map_dfr(seq_along(l_restarts), function(i) {
    g <- l_restarts[[i]]$gamma
    tibble(restart = i, module = rownames(g), topic_label = topic_labels[apply(g, 1, which.max)])
}) %>%
    count(module, topic_label) %>%
    # topics a module was never assigned to are absent from count(); fill them as 0
    complete(module, topic_label = topic_labels, fill = list(n = 0)) %>%
    group_by(module) %>%
    mutate(assignment_prop = n / sum(n)) %>%
    ungroup() %>% 
    
    # add full labels
    mutate(
        topic = case_when(
            topic_label == "GR"    ~ "genomic regulation / \nRNA processing",
            topic_label == "MORPH" ~ "neuron development / \nmorphogenesis",
            topic_label == "SYN"   ~ "synaptic transmission / \nion transport"
        ),
        topic = factor(topic, levels = levels(df_gamma$topic))
    ) %>%
    dplyr::select(module, topic, topic_label, assignment_prop)

# Consensus assignment: the topic a module is placed in most often
df_topic_consensus <- df_assignment_prop %>%
    group_by(module) %>%
    slice_max(assignment_prop, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(module, consensus_topic = topic_label, assignment_confidence = assignment_prop) %>%
    # compare against the single-fit assignment used in 2.4
    left_join(
        df_gamma %>% filter(top_topic == "yes") %>% dplyr::select(module, spectral_topic = topic_label),
        by = "module"
    ) %>%
    mutate(agrees_with_spectral = consensus_topic == spectral_topic) %>%
    arrange(assignment_confidence)

saveRDS(
    list(
        consensus       = df_topic_consensus,
        assignment_prop = df_assignment_prop,
        n_restarts      = n_restarts
    ),
    file = file.path(notebook_out_dir, "02_00_topic-restart-stability.RDS")
)

# Modules whose consensus assignment differs from the single-fit assignment
df_topic_consensus %>% filter(!agrees_with_spectral)
```


# --- 3) Supplement figures --- #


## 3.1 Select k for topic model

```{r plot_select_k, fig.height = 3, fig.width = 3}
p_k_tradeoff <- df_k_metrics %>%
    mutate(color = ifelse(K == 3, "firebrick", "steelblue")) %>% 

    ggplot(aes(x = semcoh, y = exclus, label = K)) +
    geom_path(color = "grey70", linewidth = 0.5) +
    geom_point(aes(color = I(color)), size = 2) +
    ggrepel::geom_text_repel(size = 2.5, min.segment.length = 0) +
    labs(
        x     = "Semantic coherence",
        y     = "Exclusivity",
        title = "Coherence vs exclusivity tradeoff"
    ) +
    theme(
        plot.title = element_text(face = "bold")
    )
p_k_tradeoff

ggsave(
    filename = file.path(figures_dir, "supplement", "02_02_topic-model-K-selection.pdf"),
    plot = p_k_tradeoff,
    width = 3,
    height = 3,
    dpi = 300
)
```


# --- 4) Tables --- #

## 4.1 Topic model summary

Export the top words per topic across STM metrics (highest prob / FREX / lift / score),
wide format with one column block per topic (SYN, GR, MORPH).

```{r topic_top_words_table}
n_top_words <- 10
lt <- stm::labelTopics(topic_model, n = n_top_words)

# Map each topic number -> theme label via a signature top-probability word
# (robust to topic renumbering across re-runs)
topic_theme <- apply(lt$prob, 1, function(words) {
    if      ("synaptic"          %in% words) "SYN"
    else if ("acetyltransferase" %in% words) "GR"
    else if ("morphogenesis"     %in% words) "MORPH"
    else                                     NA_character_
})
topic_for_label <- setNames(seq_along(topic_theme), topic_theme)

# Assemble <LABEL>_<metric> columns in SYN, GR, MORPH order
df_topic_top_words <- purrr::map_dfc(c("SYN", "GR", "MORPH"), function(label) {
    k <- topic_for_label[[label]]
    setNames(
        list(lt$prob[k, ], lt$frex[k, ], lt$lift[k, ], lt$score[k, ]),
        paste0(label, c("_prob", "_frex", "_lift", "_score"))
    ) %>% as_tibble()
})

write_csv(df_topic_top_words,
          file.path(notebook_out_dir, "02_00_topic-top-words.csv")
)
```

## 4.2 Module-topic assignments

```{r}
saveRDS(
    list(
        consensus       = df_topic_consensus,
        assignment_prop = df_assignment_prop,
        n_restarts      = n_restarts
    ),
    file = file.path(notebook_out_dir, "02_00_topic-restart-stability.RDS")
)
topic_stability <- readRDS(file.path(analysis_dir, "02_module_enrichments", "02_00_topic-restart-stability.RDS"))

df_gamma <- readRDS(file.path(analysis_dir, "02_module_enrichments", "02_00_topic-module-gamma.RDS"))
```


```{r module_topic_assignment}
topic_stability$consensus %>% 
    left_join(df_stepwise_res %>% dplyr::select(module, module_label)) %>% 
    dplyr::select(module, module_label, final_assigned_topic = consensus_topic) %>% 
    left_join(topic_stability$assignment_prop) %>% 
    left_join(df_gamma %>% dplyr::select(-c(z_score, p_value, p_adj, topic_color, top_topic))) %>% 

    write_csv(file.path(notebook_out_dir, "02_00_topic-module-assignments.csv"))
```
